### 2015-09-08 nnt_manu.Rnw
templates from JSS

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Achim Zeileis\\Universit\"at Innsbruck \And 
        Second Author\\Plus Affiliation}
\title{A Capitalized Title: Something about a Package \pkg{foo}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Achim Zeileis, Second Author} %% comma-separated
\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
  The abstract of the article.
}
\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Achim Zeileis\\
  Department of Statistics and Mathematics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Achim.Zeileis@uibk.ac.at}\\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

% knitr options
<<setup,include = F,cache = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign = TRUE, digits = 1)
@
% get online bib file
<<echo = FALSE, cache = FALSE>>=
refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
refs <- rawToChar(refs$content)
writeLines(refs, con = file('refs.bib'))
@

%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[About Java]{About \proglang{Java}}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

\end{document}

### 2015-09-08 nnt_manu.tex
templates from JSS

--- 2015-09-08 nnt_manu.Rnw
@@
-\documentclass[article]{jss}
+\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 % knitr options
-<<setup,include = F,cache = F>>=
-library(knitr)
-# set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
-options(replace.assign = TRUE, digits = 1)
-@
+
 % get online bib file
-<<echo = FALSE, cache = FALSE>>=
-refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
-refs <- rawToChar(refs$content)
-writeLines(refs, con = file('refs.bib'))
-@
 
@@
 %% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}

### 2015-09-08 nnt_manu.Rnw
started adding my content

@@
 %% almost as usual
-\author{Achim Zeileis\\Universit\"at Innsbruck \And 
-        Second Author\\Plus Affiliation}
-\title{A Capitalized Title: Something about a Package \pkg{foo}}
+\author{Marcus W. Beck\\Oak Ridge Institute for Science and Education\\US Environmental Protection Agency}
+\title{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}
 
 %% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Achim Zeileis, Second Author} %% comma-separated
-\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
-\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
+\Plainauthor{Marcus W. Beck} %% comma-separated
+\Plaintitle{NeuralNetTools: Visualization and Analysis Tools for Neural Networks} %% without formatting
+\Shorttitle{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}
 
@@
 \Abstract{
-  The abstract of the article.
+Functions within this package can be used for the interpretation of neural network models created in \proglang{R}, including functions to plot a neural network interpretation diagram, evaluation of variable importance, and a sensitivity analysis of input variables.
 }
-\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
-\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
+\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
+\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
 %% at least one keyword must be supplied
@@
 \Address{
-  Achim Zeileis\\
-  Department of Statistics and Mathematics\\
-  Faculty of Economics and Statistics\\
-  Universit\"at Innsbruck\\
-  6020 Innsbruck, Austria\\
-  E-mail: \email{Achim.Zeileis@uibk.ac.at}\\
-  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+  Marcus W. Beck\\
+  Oak Ridge Institude for Science and Education\\
+  US Environmental Protection Agency\\
+  National Health and Environmental Effects Research Laboratory\\
+  Gulf Ecology Division, 1 Sabine Island Drive\\
+  Gulf Breeze, Florida, 32561, USA\\
+  E-mail: \email{beck.marcus@epa.gov}
 }
@@
 
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{document}
 
-%% include your article here, just as usual
 %% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
 
-\section[About Java]{About \proglang{Java}}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
+\section[Introduction]{Introduction}
+
+\citep{Goh95}
+
+\section[Package structure]{Package structure}
+
+Packages in R that can create neural networks and relevant methods in NeuralNetTools
+
+Cautionary note about finding the `optimal' model.
+
+\subsection{Visualizing neural networks}
+
+The number of existing functions in \pkg{R} to view neural networks is minimal.
+Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
+To our knowledge, only the \pkg{neuralnet} package provides a method for plotting neural networks created with the \code{neuralnet} function.
+Although useful for viewing the basic structure, the output is visually minimal and does not include options for customization (verify).
+
+This function plots a neural network as a neural interpretation diagram as in \citet{Ozesmi99}.
+Options to plot without color-coding or shading of weights are also provided.
+The default settings plot positive weights between layers as black lines and negative weights as grey lines.
+Line thickness is in proportion to relative magnitude of each weight.
+The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables.
+One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn.
+The output layer is plotted last with nodes labeled as O1 through On.
+Bias nodes connected to the hidden and output layers are also shown.
+Neural networks created using \code{mlp} do not show bias layers.
+
+A primary network and a skip layer network can be plotted for nnet models with a skip layer connection.
+The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}.
+If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights.
+Viewing both plots is recommended to see which network has larger relative weights.
+Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0}) will include bias connections to the output layer, whereas these are not included in the plot of the skip layer if size is greater than zero.
+
+Pruned networks in RSNNS.
+
+\subsection{Evaluating variable importance}
+
+The \code{garson} function uses Garson's algorithm to evaluate relative variable importance.
+This function identifies the relative importance of explanatory variables for a single response variable by deconstructing the model weights.
+The importance of each variable can be determined by identifying all weighted connections between the layers in the network.
+That is, all weights connecting the specific input node that pass through the hidden layer to the response variable are identified.
+This is repeated for all other explanatory variables until a list of all weights that are specific to each input variable is obtained.
+The connections are tallied for each input node and scaled relative to all other inputs.
+A single value is obtained for each explanatory variable that describes the relationship with the response variable in the model.
+The results indicate relative importance as the absolute magnitude from zero to one.
+The function cannot be used to evaluate the direction of the response.
+Only neural networks with one hidden layer and one output node can be evaluated.
+
+The \code{olden} function is an alternative and more flexible approach to evaluate variable importance.
+The function calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output neuron and sums the product across all hidden neurons.
+An advantage of this approach is the relative contributions of each connection weight are maintained in terms of both magnitude and sign as compared to Garson's algorithm which only considers the absolute magnitude.
+For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect whereas Garson's algorithm may provide misleading results based on the absolute magnitude.
+An additional advantage is that Olden's algorithm is capable of evaluating neural networks with multiple hidden layers and response variables.
+The importance values assigned to each variable are in units that are based directly on the summed product of the connection weights.
+The actual values should only be interpreted based on relative sign and magnitude between explanatory variables.
+Comparisons between different models should not be made.
+
+Issues with different indications of variable importance as a model is refit...
+
+\subsection{Sensitivity analysis}
+
+The Lek profile method is described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
+The profile method is fairly generic and can be extended to any statistical model in R with a predict method.
+However, it is one of few methods used to evaluate sensitivity in neural networks.
+
+The profile method can be used to evaluate the effect of explanatory variables by returning a plot of the predicted response across the range of values for each separate variable.
+The original profile method evaluated the effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum).
+This is implemented in in the function by creating a matrix of values for explanatory variables where the number of rows is the number of observations and the number of columns is the number of explanatory variables.
+All explanatory variables are held at their mean (or other constant value) while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
+This matrix (or data frame) is then used to predict values of the response variable from a fitted model object.
+This is repeated for each explanatory variable to obtain all response curves.
+Values passed to \code{split_vals} must range from zero to one to define the quantiles for holding unevaluated explanatory variables.
+
+An alternative implementation of the profile method is to group the unevaluated explanatory variables using groupings defined by the statistical properties of the data.
+Covariance among predictors may present unlikely scenarios if holding all unevaluated variables at the same level.
+To address this issue, the function provides an option to hold unevalutaed variable at mean values defined by natural clusters in the data.
+kmeans clustering is used on the input data.frame of explanatory variables if the argument passed to \code{split_vals} is an integer value greater than one.
+The centers of the clusters are then used as constant values for the unevaluated variables.
+An arbitrary grouping scheme can also be passed to \code{split_vals} as a data.frame where the user can specify exact values for holding each value constant (see the examples).
+Examples in \citet{Beck14a} show this...
+
+For all plots, the legend with the 'splits' label indicates the colors that correspond to each group.
+The groups describe the values at which unevaluated explanatory variables were held constant, either as specific quantiles, group assignments based on clustering, or in the arbitrary grouping defined by the user.
+The constant values of each explanatory variable for each split can be viewed as a barplot by using \code{split_show = TRUE}.
+
+Note that there is no predict method for neuralnet objects from the nn package.
+The lekprofile method for nn objects uses the nnet package to recreate the input model, which is then used for the sensitivity predictions.
+This approach only works for networks with one hidden layer.
+
+\section[Future development]{Future development}
+
+\section[Acknowledgments]{Acknowledgments}
+
+\bibliographystyle{jss}
+\bibliography{refs}
 

### 2015-09-08 nnt_manu.tex
started adding my content

@@
 %% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-%% include your article here, just as usual
 %% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
 
-\section[About Java]{About \proglang{Java}}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
+\section[Introduction]{Introduction}
+
+\citep{Goh95}
+
+\section[Package structure]{Package structure}
+
+Packages in R that can create neural networks and relevant methods in NeuralNetTools
+
+Cautionary note about finding the `optimal' model.
+
+\subsection{Visualizing neural networks}
+
+The number of existing functions in \pkg{R} to view neural networks is minimal.
+Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
+To our knowledge, only the \pkg{neuralnet} package provides a method for plotting neural networks created with the \code{neuralnet} function.
+Although useful for viewing the basic structure, the output is visually minimal and does not include options for customization (verify).
+
+This function plots a neural network as a neural interpretation diagram as in \citet{Ozesmi99}.
+Options to plot without color-coding or shading of weights are also provided.
+The default settings plot positive weights between layers as black lines and negative weights as grey lines.
+Line thickness is in proportion to relative magnitude of each weight.
+The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables.
+One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn.
+The output layer is plotted last with nodes labeled as O1 through On.
+Bias nodes connected to the hidden and output layers are also shown.
+Neural networks created using \code{mlp} do not show bias layers.
+
+A primary network and a skip layer network can be plotted for nnet models with a skip layer connection.
+The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}.
+If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights.
+Viewing both plots is recommended to see which network has larger relative weights.
+Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0}) will include bias connections to the output layer, whereas these are not included in the plot of the skip layer if size is greater than zero.
+
+Pruned networks in RSNNS.
+
+\subsection{Evaluating variable importance}
+
+The \code{garson} function uses Garson's algorithm to evaluate relative variable importance.
+This function identifies the relative importance of explanatory variables for a single response variable by deconstructing the model weights.
+The importance of each variable can be determined by identifying all weighted connections between the layers in the network.
+That is, all weights connecting the specific input node that pass through the hidden layer to the response variable are identified.
+This is repeated for all other explanatory variables until a list of all weights that are specific to each input variable is obtained.
+The connections are tallied for each input node and scaled relative to all other inputs.
+A single value is obtained for each explanatory variable that describes the relationship with the response variable in the model.
+The results indicate relative importance as the absolute magnitude from zero to one.
+The function cannot be used to evaluate the direction of the response.
+Only neural networks with one hidden layer and one output node can be evaluated.
+
+The \code{olden} function is an alternative and more flexible approach to evaluate variable importance.
+The function calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output neuron and sums the product across all hidden neurons.
+An advantage of this approach is the relative contributions of each connection weight are maintained in terms of both magnitude and sign as compared to Garson's algorithm which only considers the absolute magnitude.
+For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect whereas Garson's algorithm may provide misleading results based on the absolute magnitude.
+An additional advantage is that Olden's algorithm is capable of evaluating neural networks with multiple hidden layers and response variables.
+The importance values assigned to each variable are in units that are based directly on the summed product of the connection weights.
+The actual values should only be interpreted based on relative sign and magnitude between explanatory variables.
+Comparisons between different models should not be made.
+
+Issues with different indications of variable importance as a model is refit...
+
+\subsection{Sensitivity analysis}
+
+The Lek profile method is described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
+The profile method is fairly generic and can be extended to any statistical model in R with a predict method.
+However, it is one of few methods used to evaluate sensitivity in neural networks.
+
+The profile method can be used to evaluate the effect of explanatory variables by returning a plot of the predicted response across the range of values for each separate variable.
+The original profile method evaluated the effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum).
+This is implemented in in the function by creating a matrix of values for explanatory variables where the number of rows is the number of observations and the number of columns is the number of explanatory variables.
+All explanatory variables are held at their mean (or other constant value) while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
+This matrix (or data frame) is then used to predict values of the response variable from a fitted model object.
+This is repeated for each explanatory variable to obtain all response curves.
+Values passed to \code{split_vals} must range from zero to one to define the quantiles for holding unevaluated explanatory variables.
+
+An alternative implementation of the profile method is to group the unevaluated explanatory variables using groupings defined by the statistical properties of the data.
+Covariance among predictors may present unlikely scenarios if holding all unevaluated variables at the same level.
+To address this issue, the function provides an option to hold unevalutaed variable at mean values defined by natural clusters in the data.
+kmeans clustering is used on the input data.frame of explanatory variables if the argument passed to \code{split_vals} is an integer value greater than one.
+The centers of the clusters are then used as constant values for the unevaluated variables.
+An arbitrary grouping scheme can also be passed to \code{split_vals} as a data.frame where the user can specify exact values for holding each value constant (see the examples).
+Examples in \citet{Beck14a} show this...
+
+For all plots, the legend with the 'splits' label indicates the colors that correspond to each group.
+The groups describe the values at which unevaluated explanatory variables were held constant, either as specific quantiles, group assignments based on clustering, or in the arbitrary grouping defined by the user.
+The constant values of each explanatory variable for each split can be viewed as a barplot by using \code{split_show = TRUE}.
+
+Note that there is no predict method for neuralnet objects from the nn package.
+The lekprofile method for nn objects uses the nnet package to recreate the input model, which is then used for the sensitivity predictions.
+This approach only works for networks with one hidden layer.
+
+\section[Future development]{Future development}
+
+\section[Acknowledgments]{Acknowledgments}
+
+\bibliographystyle{jss}
+\bibliography{refs}
 

### 2015-09-09 nnt_manu.Rnw
draft of intro done - bloated as hell

@@
 
+\usepackage{acronym}
+\acrodef{mlp}[mlp]{multilayer perceptron}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\citep{Goh95}
+The increasing quantity of information and computational capacity to address relevant research questions has contributed to the growth of data science as a legitimate field of study.
+Data science is a relatively new paradigm of analysis that focuses on the synthesis of unstructured information from multiple sources to identify patterns or trends `born from the data' \citep{Kelling09}.
+A central theme is the focus on data exploration and prediction as compared to specific hypothesis-testing of traditional, domain-specific methods of scientific exploration \citep{Kell03}.
+Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the continuing advancement of techniques for rapid acquisition of data.
+Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}).
+By contrast, disciplines that have historically been limited by data quantity, such as ecological studies across broad temporal and spatial scales, have shown a need for data intensive approaches given the improved ability to acquire information (e.g., \citet{Swanson15}).
+Regardless of the discipline, quantitative approaches that explicitly focus on  inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
+
+Statistical methods that have been used to support data exploration for deductive analysis are numerous.
+A common theme among many of these methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns in the data with minimal human intervention.
+Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms.
+Although the conceptual model was introduced several decades ago \citep{Maier00}, neural networks continue to maintain a central role in data intensive science.
+The most popular form of neural network is the \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
+This model is typically used to predict the response of one or more variables given co-occurrence of any of a number of explanatory variables.
+The hallmark feature of the \ac{mlp} is the characterization of relationships between variabiles using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through an iterative training process with the backpropation algorithm.
+Conceptually, the \ac{mlp} is nothing more than a hyper-parameterized non-linear model that can a fit smooth function with almost non-existent residual error for any dataset.
+
+An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of critical model information.
+Information such as variable importance or model sensitivity are necessary aspects of exploraty data analysis that are not easily obtained from a neural network.
+As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables.
+\citet{Olden02} provides a rebuttal to this concern by providing methods to extract information from neural networks, most of which were previously available but not commonly used to evaluate neural networks.
+For example, \citet{Olden02} describes neural interpretation diagrams for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
+These quantitative tools `illuminate the black box' by disaggregating the network parameters as a means to characterize relationships between variables.
+In essence, \ac{mlp} neural networks were developed for prediction but methods in \citep{Olden02} leverage these models to describe data signals.
+Increasing the accesibility of these diagnostic tools will have value for exploratory analysis in data science.
+
+This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.
+Functions provided by the package are those previously described in \citep{Olden02} but have not been available in an open-source proramming environment.
+The reach of the package is all-inclusive such that generic functions were developed using S3 methods for all neural network object classes available in \proglang{R}.
+The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) briefly describe similarities and differences between existing neural network packages in \proglang{R}, and 3) describe the theory and application of the primary functions in the \pkg{NeuralNetTools} package.
+The package is currently available on CRAN, whereas the development version is maintained as a GitHub repository.
+
+\section[Theoretical foundation]{Theoretical foundation and existing R packages}
+
+Packages available in R to create neural networks (similarities, differences)
+
+hyper-dimensional
+
+Formulaically \citep{Ripley96}
+$y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right) $
 
 \section[Package structure]{Package structure}
-
-Packages in R that can create neural networks and relevant methods in NeuralNetTools
-
-Cautionary note about finding the `optimal' model.
 
@@
 
+\section[Conclusions]{Conclusions}
+
+A cautionary note about the `optimal network' and reproducibility of results.
+
 \section[Acknowledgments]{Acknowledgments}

### 2015-09-14 nnt_manu.Rnw
background about neural nets and available packages

--- 2015-09-08 nnt_manu.Rnw
@@
 
-\citep{Goh95}
+The increasing quantity of information and computational capacity to address relevant research questions has contributed to the growth of data science as a legitimate field of study.
+Data science is a relatively new paradigm of analysis that focuses on the synthesis of unstructured information from multiple sources to identify patterns or trends `born from the data' \citep{Kelling09}.
+A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
+Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data.
+Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}).
+Morever, disciplines that have historically been limited by data quantity, such as ecological studies across broad temporal and spatial scales, have also realized the importance of data intensive approaches of analysis given the improved ability to acquire information (e.g., \citet{Swanson15}).
+Regardless of the discipline, quantitative methods that explicitly focus on inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
+
+Statistical methods that have been used to support data exploration for inductive analysis are numerous \citep{Jain00}.
+A common theme among these methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns in the data with minimal human intervention.
+Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.
+Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in data intensive science.
+The most popular form of neural network is the \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
+This model is typically used to predict the response of one or more variables given co-occurrence of one to many explanatory variables.
+The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through an iterative training process with the backpropation algorithm.
+Conceptually, the \ac{mlp} is nothing more than a hyper-parameterized non-linear model that can fit a smooth function to any dataset with almost non-existent residual error \citep{Hornik91}.
+
+An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of critical model information.
+Information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
+As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
+\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks, most of which were previously available but not commonly used.
+For example, \citet{Olden02} describe neural interpretation diagrams for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
+These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.
+In essence, \ac{mlp} neural networks were developed for prediction but methods described in \citep{Olden02} leverage these models to describe data signals.
+Increasing the accesibility of these diagnostic tools will have value for exploratory analysis in data science.
+
+This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.
+Functions provided by the package are those previously described in \citep{Olden02} but have not been available in an open-source programming environment.
+The reach of the package is all-inclusive such that generic functions were developed using S3 methods for all neural network object classes available in \proglang{R}.
+The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) briefly describe similarities and differences between existing neural network packages in \proglang{R}, and 3) describe the theory and application of the primary functions in the \pkg{NeuralNetTools} package.
+The package is currently available on CRAN, whereas the development version is maintained as a GitHub repository.
+
+\section[Theoretical foundation]{Theoretical foundation and existing R packages}
+
+The following is a brief description of the basic structure and methods for creating a network to provide sufficient context for the \pkg{NeuralNetTools} package.
+A detailed discussion of the \ac{mlp} neural network is beyond the scope of this manuscript and interested readers should consult \citet{Rumelhart86,Ripley96} and references therein for a more thorough review.
+An intriguing characteristic of the neural network approach to predictive modelling is the absence of assumptions regarding the distributional characteristics of the response variables.
+Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.
+The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.
+Information travels in one direction where a set of values for one to many variables in the input layer propagates through one or more hidden layers to the resulting layer of the response variables.
+`Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.
+Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.
+`Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.
+
+Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.
+The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.
+Training is commonly achieved using the backpropagation algorithm described in detail in \citep{Rumelhart86}.
+This algorithm identifies the optimal weighting scheme between layers through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.
+The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.
+The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.
+The process is repeated until the chosen error function is minimized.
+Formulaically, a generic \ac{mlp} neural network can be represented as \citep{Ripley96}:
+\begin{equation}
+y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
+\end{equation}
+where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the respective weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node as the information progresses through the network.
+
+<<eval = F, echo = F>>=
+# total downloads for neural network packages off CRAN
+# make sure to change cdate
+library(cranlogs)
+
+tots <- c(
+  sum(cran_downloads('nnet', from = '2000-01-01', to = '2015-09-13')$count),
+  sum(cran_downloads('neuralnet', from = '2000-01-01', to = '2015-09-13')$count),
+  sum(cran_downloads('RSNNS', from = '2000-01-01', to = '2015-09-13')$count),
+  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = '2015-09-13')$count)
+)
+@
+
+Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
+At the time of writing, \pkg{FCNN4R} is a relatively new package that represents less than 1\% of the total downloads of all neural network packages on CRAN.
+As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.
+In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models should be comparable.
+Differences between the packages primarily include default values of the arguments and the ability to handle multiple response variables or multiple hidden layers in the same model.
+The \code{nnet} function can take separate (or combined) input or response inputs as each as a separate \code{data.frame} or as a \code{formula}, the \code{neuralnet} function can only use a \code{formula} as input, and the \code{mlp} function (\pkg{RSNNS}) can only take a \code{data.frame} as combined or separate variables as input.
+The \code{neuralnet} function is not capable of modelling multiple response variables, unless the response is a categorical variable where one node for each outcome is used in the response.
+Additionally, the default output for the \code{neuralnet} function is linear, whereas the opposite is true for the other two functions.
+Other differences that influence how functions in \pkg{NeuralNetTools} handle object classes from the different packages are described below.
 
 \section[Package structure]{Package structure}
-
-Packages in R that can create neural networks and relevant methods in NeuralNetTools
-
-Cautionary note about finding the `optimal' model.
 
@@
 Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
-To our knowledge, only the \pkg{neuralnet} package provides a method for plotting neural networks created with the \code{neuralnet} function.
+To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks.
 Although useful for viewing the basic structure, the output is visually minimal and does not include options for customization (verify).
@@
 
+\section[Conclusions]{Conclusions}
+
+A cautionary note about the `optimal network' and reproducibility of results.
+Note that most literature sources suggest variable standardization prior to using a model yet none of the existing packages in R provide this functionality as a default option (verify).
+
 \section[Acknowledgments]{Acknowledgments}

### 2015-09-14 nnt_manu.tex
background about neural nets and available packages

--- 2015-09-08 nnt_manu.tex
@@
 
-\citep{Goh95}
+The increasing quantity of information and computational capacity to address relevant research questions has contributed to the growth of data science as a legitimate field of study.
+Data science is a relatively new paradigm of analysis that focuses on the synthesis of unstructured information from multiple sources to identify patterns or trends `born from the data' \citep{Kelling09}.
+A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
+Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data.
+Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}).
+Morever, disciplines that have historically been limited by data quantity, such as ecological studies across broad temporal and spatial scales, have also realized the importance of data intensive approaches of analysis given the improved ability to acquire information (e.g., \citet{Swanson15}).
+Regardless of the discipline, quantitative methods that explicitly focus on inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
+
+Statistical methods that have been used to support data exploration for inductive analysis are numerous \citep{Jain00}.
+A common theme among these methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns in the data with minimal human intervention.
+Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.
+Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in data intensive science.
+The most popular form of neural network is the \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
+This model is typically used to predict the response of one or more variables given co-occurrence of one to many explanatory variables.
+The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through an iterative training process with the backpropation algorithm.
+Conceptually, the \ac{mlp} is nothing more than a hyper-parameterized non-linear model that can fit a smooth function to any dataset with almost non-existent residual error \citep{Hornik91}.
+
+An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of critical model information.
+Information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
+As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
+\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks, most of which were previously available but not commonly used.
+For example, \citet{Olden02} describe neural interpretation diagrams for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
+These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.
+In essence, \ac{mlp} neural networks were developed for prediction but methods described in \citep{Olden02} leverage these models to describe data signals.
+Increasing the accesibility of these diagnostic tools will have value for exploratory analysis in data science.
+
+This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.
+Functions provided by the package are those previously described in \citep{Olden02} but have not been available in an open-source programming environment.
+The reach of the package is all-inclusive such that generic functions were developed using S3 methods for all neural network object classes available in \proglang{R}.
+The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) briefly describe similarities and differences between existing neural network packages in \proglang{R}, and 3) describe the theory and application of the primary functions in the \pkg{NeuralNetTools} package.
+The package is currently available on CRAN, whereas the development version is maintained as a GitHub repository.
+
+\section[Theoretical foundation]{Theoretical foundation and existing R packages}
+
+The following is a brief description of the basic structure and methods for creating a network to provide sufficient context for the \pkg{NeuralNetTools} package.
+A detailed discussion of the \ac{mlp} neural network is beyond the scope of this manuscript and interested readers should consult \citet{Rumelhart86,Ripley96} and references therein for a more thorough review.
+An intriguing characteristic of the neural network approach to predictive modelling is the absence of assumptions regarding the distributional characteristics of the response variables.
+Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.
+The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.
+Information travels in one direction where a set of values for one to many variables in the input layer propagates through one or more hidden layers to the resulting layer of the response variables.
+`Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.
+Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.
+`Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.
+
+Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.
+The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.
+Training is commonly achieved using the backpropagation algorithm described in detail in \citep{Rumelhart86}.
+This algorithm identifies the optimal weighting scheme between layers through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.
+The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.
+The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.
+The process is repeated until the chosen error function is minimized.
+Formulaically, a generic \ac{mlp} neural network can be represented as \citep{Ripley96}:
+\begin{equation}
+y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
+\end{equation}
+where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the respective weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node as the information progresses through the network.
+
+Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
+At the time of writing, \pkg{FCNN4R} is a relatively new package that represents less than 1\% of the total downloads of all neural network packages on CRAN.
+As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.
+In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models should be comparable.
+Differences between the packages primarily include default values of the arguments and the ability to handle multiple response variables or multiple hidden layers in the same model.
+The \code{nnet} function can take separate (or combined) input or response inputs as each as a separate \code{data.frame} or as a \code{formula}, the \code{neuralnet} function can only use a \code{formula} as input, and the \code{mlp} function (\pkg{RSNNS}) can only take a \code{data.frame} as combined or separate variables as input.
+The \code{neuralnet} function is not capable of modelling multiple response variables, unless the response is a categorical variable where one node for each outcome is used in the response.
+Additionally, the default output for the \code{neuralnet} function is linear, whereas the opposite is true for the other two functions.
+Other differences that influence how functions in \pkg{NeuralNetTools} handle object classes from the different packages are described below.
 
 \section[Package structure]{Package structure}
-
-Packages in R that can create neural networks and relevant methods in NeuralNetTools
-
-Cautionary note about finding the `optimal' model.
 

### 2015-09-15 nnt_manu.Rnw
more about package structure, need to trim the fat

@@
 # set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
+opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
 options(replace.assign = TRUE, digits = 1)
@@
 
-The following is a brief description of the basic structure and methods for creating a network to provide sufficient context for the \pkg{NeuralNetTools} package.
-A detailed discussion of the \ac{mlp} neural network is beyond the scope of this manuscript and interested readers should consult \citet{Rumelhart86,Ripley96} and references therein for a more thorough review.
 An intriguing characteristic of the neural network approach to predictive modelling is the absence of assumptions regarding the distributional characteristics of the response variables.
@@
 Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.
 An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
-At the time of writing, \pkg{FCNN4R} is a relatively new package that represents less than 1\% of the total downloads of all neural network packages on CRAN.
+At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.
 As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.
-In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models should be comparable.
-Differences between the packages primarily include default values of the arguments and the ability to handle multiple response variables or multiple hidden layers in the same model.
-The \code{nnet} function can take separate (or combined) input or response inputs as each as a separate \code{data.frame} or as a \code{formula}, the \code{neuralnet} function can only use a \code{formula} as input, and the \code{mlp} function (\pkg{RSNNS}) can only take a \code{data.frame} as combined or separate variables as input.
-The \code{neuralnet} function is not capable of modelling multiple response variables, unless the response is a categorical variable where one node for each outcome is used in the response.
-Additionally, the default output for the \code{neuralnet} function is linear, whereas the opposite is true for the other two functions.
-Other differences that influence how functions in \pkg{NeuralNetTools} handle object classes from the different packages are described below.
+In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.
+Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.
+The primary functions in each package also vary in the type of input used to specify the model structure.
+Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.
 
 \section[Package structure]{Package structure}
+
+The stable release of \pkg{NeurelNetTools} can be installed from CRAN and the development version can be installed from GitHub:
+
+<<eval = F>>=
+# install from CRAN
+install.packages('NeuralNetTools')
+library(NeuralNetTools)
+
+# install from GitHub
+install.package('devtools')
+library(devtools)
+install_github('fawda123/NeuralNetTools')
+library(NeuralNetTools)
+@
+
+\pkg{NeuralNetTools} includes four main functions that were developed following techniques described in \citet{Olden02} and references therein.
+These functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.
+Most of the functions require the extraction of model weights in a common format for each of the neural network object classes in \proglang{R}.
+The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.
+A two-element \code{list} is returned with the first element describing the structure of the network and the second element as a named list of weight values for the input model.
+The function is used internally within the main package functions but may be useful for comparing networks of different object classes.
+
+A sample dataset is also included with \pkg{NeuralNetTools}.
+The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.
+The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.
+The response variables are also standardized from zero to one.
+A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
+The \code{neuraldat} package provides data following this structure.
+Novel datasets can be preprocessed in the common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
 
@@
 Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
-To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks.
-Although useful for viewing the basic structure, the output is visually minimal and does not include options for customization (verify).
+To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.
+Although useful for viewing the basic structure, the output is visually minimal and does not include extensive options for customization.
 
-This function plots a neural network as a neural interpretation diagram as in \citet{Ozesmi99}.
-Options to plot without color-coding or shading of weights are also provided.
+Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes many options to customize plot aesthetics.
+A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respecitively.
 The default settings plot positive weights between layers as black lines and negative weights as grey lines.
@@
 Bias nodes connected to the hidden and output layers are also shown.
-Neural networks created using \code{mlp} do not show bias layers.
+Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.
+
+<<echo = F>>=
+library(NeuralNetTools)
+@
+<<>>=
+# create a model using the nnet package
+# neuraldat is included with NeuralDat
+@
 

### 2015-09-15 nnt_manu.tex
more about package structure, need to trim the fat

@@
 Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.
 An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
-At the time of writing, \pkg{FCNN4R} is a relatively new package that represents less than 1\% of the total downloads of all neural network packages on CRAN.
+At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.
 As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.
-In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models should be comparable.
-Differences between the packages primarily include default values of the arguments and the ability to handle multiple response variables or multiple hidden layers in the same model.
-The \code{nnet} function can take separate (or combined) input or response inputs as each as a separate \code{data.frame} or as a \code{formula}, the \code{neuralnet} function can only use a \code{formula} as input, and the \code{mlp} function (\pkg{RSNNS}) can only take a \code{data.frame} as combined or separate variables as input.
-The \code{neuralnet} function is not capable of modelling multiple response variables, unless the response is a categorical variable where one node for each outcome is used in the response.
-Additionally, the default output for the \code{neuralnet} function is linear, whereas the opposite is true for the other two functions.
-Other differences that influence how functions in \pkg{NeuralNetTools} handle object classes from the different packages are described below.
+In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.
+Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.
+The primary functions in each package also vary in the type of input used to specify the model structure.
+Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.
 
 \section[Package structure]{Package structure}
+
+The stable release of \pkg{NeurelNetTools} can be installed from CRAN and the development version can be installed from GitHub:
+
+\begin{kframe}
+\begin{alltt}
+\hlcom{# install from CRAN}
+\hlkwd{install.packages}\hlstd{(}\hlstr{'NeuralNetTools'}\hlstd{)}
+\hlkwd{library}\hlstd{(NeuralNetTools)}
+
+\hlcom{# install from GitHub}
+\hlkwd{install.package}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
+\hlkwd{library}\hlstd{(devtools)}
+\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/NeuralNetTools'}\hlstd{)}
+\hlkwd{library}\hlstd{(NeuralNetTools)}
+\end{alltt}
+\end{kframe}
+
+\pkg{NeuralNetTools} includes four main functions that were developed following techniques described in \citet{Olden02} and references therein.
+These functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.
+Most of the functions require the extraction of model weights in a common format for each of the neural network object classes in \proglang{R}.
+The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.
+A two-element \code{list} is returned with the first element describing the structure of the network and the second element as a named list of weight values for the input model.
+The function is used internally within the main package functions but may be useful for comparing networks of different object classes.
+
+A sample dataset is also included with \pkg{NeuralNetTools}.
+The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.
+The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.
+The response variables are also standardized from zero to one.
+A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
+The \code{neuraldat} package provides data following this structure.
+Novel datasets can be preprocessed in the common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
 
@@
 Bias nodes connected to the hidden and output layers are also shown.
-Neural networks created using \code{mlp} do not show bias layers.
+Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.
+
+\begin{kframe}
+\begin{alltt}
+\hlcom{# create a model using the nnet package}
+\hlcom{# neuraldat is included with NeuralDat}
+\end{alltt}
+\end{kframe}
 

### 2015-09-17 nnt_manu.Rnw
added refs for neural network packages, need to mention others in the RSNNS paper

@@
 library(cranlogs)
+today <- as.character(as.Date(Sys.time()))
 
 tots <- c(
-  sum(cran_downloads('nnet', from = '2000-01-01', to = '2015-09-13')$count),
-  sum(cran_downloads('neuralnet', from = '2000-01-01', to = '2015-09-13')$count),
-  sum(cran_downloads('RSNNS', from = '2000-01-01', to = '2015-09-13')$count),
-  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = '2015-09-13')$count)
+  sum(cran_downloads('nnet', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('neuralnet', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('RSNNS', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('AMORE', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('nonmlp', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('qrnn', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('frbf', from = '2000-01-01', to = today)$count)
 )
+
 @
 
-Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
-The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.
-An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
+See refs in the RSNNS paper!!
+
+Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}.
+The \code{train} function from the \pkg{caret} package \citep{Kuhn15} can be used to create a neural network model by repeated training with functions from the remaining packages.
+Additional packages without \pkg{NeuralNetTools} methods include....
+
+An additional package, \pkg{FCNN4R} \citep{Klima15}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
 At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.
@@
 A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
-The \code{neuraldat} package provides data following this structure.
-Novel datasets can be preprocessed in the common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
+Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
 
@@
 To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.
-Although useful for viewing the basic structure, the output is visually minimal and does not include extensive options for customization.
+Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.
 
-Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes many options to customize plot aesthetics.
-A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respecitively.
+Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes several options to customize plot aesthetics.
+A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respectively.
 The default settings plot positive weights between layers as black lines and negative weights as grey lines.
-Line thickness is in proportion to relative magnitude of each weight.
+Line thickness is in proportion to absolte magnitude of each weight.
 The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables.
@@
 Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.
+The following shows some examples of basic plots
 

### 2015-09-17 nnt_manu.tex
added refs for neural network packages, need to mention others in the RSNNS paper

@@
 
-Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
-The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.
-An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
+Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}.
+The \code{train} function from the \pkg{caret} package \citep{Kuhn15} can be used to create a neural network model by repeated training with functions from the remaining packages.
+An additional package, \pkg{FCNN4R} \citep{Klima15}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
 At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.

### 2015-09-28 nnt_manu.Rnw
started adding plot figs, clarified all neural net packages in R

@@
-\documentclass[article]{jss}
+\documentclass[article,shortnames]{jss}
 
 \usepackage{acronym}
+\usepackage{cleveref}
 \acrodef{mlp}[mlp]{multilayer perceptron}
@@
 
-An intriguing characteristic of the neural network approach to predictive modelling is the absence of assumptions regarding the distributional characteristics of the response variables.
+An intriguing characteristic of neural networks is the relaxation of assumptions for the distributional characteristics of the response variables.
 Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.
@@
 
-<<eval = F, echo = F>>=
+<<eval = T, echo = F>>=
 # total downloads for neural network packages off CRAN
@@
   sum(cran_downloads('AMORE', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('nonmlp', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('qrnn', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('frbf', from = '2000-01-01', to = today)$count)
+  sum(cran_downloads('monmlp', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('qrnn', from = '2000-01-01', to = today)$count)
 )
 
+tots_nnt <- 100 * sum(tots[1:3])/sum(tots)
+tots_nnet <- 100 * tots[1]/sum(tots)
 @
 
-See refs in the RSNNS paper!!
-
 Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}.
-The \code{train} function from the \pkg{caret} package \citep{Kuhn15} can be used to create a neural network model by repeated training with functions from the remaining packages.
-Additional packages without \pkg{NeuralNetTools} methods include....
-
-An additional package, \pkg{FCNN4R} \citep{Klima15}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
-At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.
-As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.
-In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.
-Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.
-The primary functions in each package also vary in the type of input used to specify the model structure.
-Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.
+Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
+Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
+At the time of writing, the CRAN download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.
+As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
+Methods are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class) objects.
+Additional \code{default} or \code{numeric} methods are available for some of the generic functions.
 
@@
 Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
+The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects (note the syntax differences).
+
+<<eval = T, echo= F, message = F>>=
+library(NeuralNetTools)
+@
+<<results = 'hide'>>=
+# set random seed for initial inputs during training
+set.seed(123)
+
+# mlp object, RSNNS package
+library(RSNNS)
+x <- neuraldat[, c('X1', 'X2', 'X3')]
+y <- neuraldat[, 'Y1']
+mod1 <- mlp(x, y, size = 5)
+
+# nn object, neuralnet package
+library(neuralnet)
+mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
+
+#nnet object, nnet package
+library(nnet)
+mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
+@
 
@@
 
-Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes several options to customize plot aesthetics.
-A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respectively.
+Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a network as a neural interpretation diagram \citep{Ozesmi99} and includes several options to customize plot aesthetics.
+A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to view the thickness and color of the weight connections based on mangnitude and sign, respectively.
 The default settings plot positive weights between layers as black lines and negative weights as grey lines.
-Line thickness is in proportion to absolte magnitude of each weight.
-The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables.
-One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn.
-The output layer is plotted last with nodes labeled as O1 through On.
-Bias nodes connected to the hidden and output layers are also shown.
-Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.
-The following shows some examples of basic plots
+Line thickness is in proportion to absolute magnitude of each weight (\cref{fig:plotnet}).
 
-<<echo = F>>=
-library(NeuralNetTools)
-@
-<<>>=
-# create a model using the nnet package
-# neuraldat is included with NeuralDat
+<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (left) and using the neural interpretation diagram (right).
+Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
+Default options are changed for the example.'>>=
+par(mar = c(0, 0, 0, 0))
+plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
+plotnet(mod3, rel_rsc = 8)
 @
 
-A primary network and a skip layer network can be plotted for nnet models with a skip layer connection.
+A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection.
 The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}.
@@
 
-Pruned networks in RSNNS.
+<<plotnet_skip, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
+The skip layer (left) and primary neural network (right) can be viewed separately with \\code{plotnet}.', results = 'hide'>>=
+# create a model with a skip layer
+modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
+
+# plot
+par(mar = c(0, 0, 0, 0))
+plotnet(modskip, skip = TRUE)
+plotnet(modskip)
+@
+
+The \pkg{RSNNS} package provides several algorithms that can be used to prune connections in a neural network.
+This approach is useful to identify and remove connection weights between layers that do contribute to the predictive performance of the network, which allows for a potential increase in power by increasing the degrees of freedom.
+The \pkg{RSNNS} package implements
 
@@
 
+% note that ref titles need to be in title-case 
 \bibliographystyle{jss}

### 2015-09-28 nnt_manu.tex
started adding plot figs, clarified all neural net packages in R

@@
-\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[article,shortnames]{jss}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \usepackage{acronym}
+\usepackage{cleveref}
 \acrodef{mlp}[mlp]{multilayer perceptron}
@@
 
-Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}.
-The \code{train} function from the \pkg{caret} package \citep{Kuhn15} can be used to create a neural network model by repeated training with functions from the remaining packages.
-An additional package, \pkg{FCNN4R} \citep{Klima15}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.
-At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.
-As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.
-In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.
-Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.
-The primary functions in each package also vary in the type of input used to specify the model structure.
-Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.
+Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}.
+Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
+Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
+At the time of writing, the CRAN download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 93\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
+As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
+Methods are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class) objects.
+Additional \code{default} or \code{numeric} methods are available for some of the generic functions.
 
@@
 Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
+The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects (note the syntax differences).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# set random seed for initial inputs during training}
+\hlkwd{set.seed}\hlstd{(}\hlnum{123}\hlstd{)}
+
+\hlcom{# mlp object, RSNNS package}
+\hlkwd{library}\hlstd{(RSNNS)}
+\hlstd{x} \hlkwb{<-} \hlstd{neuraldat[,} \hlkwd{c}\hlstd{(}\hlstr{'X1'}\hlstd{,} \hlstr{'X2'}\hlstd{,} \hlstr{'X3'}\hlstd{)]}
+\hlstd{y} \hlkwb{<-} \hlstd{neuraldat[,} \hlstr{'Y1'}\hlstd{]}
+\hlstd{mod1} \hlkwb{<-} \hlkwd{mlp}\hlstd{(x, y,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{)}
+
+\hlcom{# nn object, neuralnet package}
+\hlkwd{library}\hlstd{(neuralnet)}
+\hlstd{mod2} \hlkwb{<-} \hlkwd{neuralnet}\hlstd{(Y1} \hlopt{~} \hlstd{X1} \hlopt{+} \hlstd{X2} \hlopt{+} \hlstd{X3,} \hlkwc{data} \hlstd{= neuraldat,} \hlkwc{hidden} \hlstd{=} \hlnum{5}\hlstd{)}
+
+\hlcom{#nnet object, nnet package}
+\hlkwd{library}\hlstd{(nnet)}
+\hlstd{mod3} \hlkwb{<-} \hlkwd{nnet}\hlstd{(Y1} \hlopt{~} \hlstd{X1} \hlopt{+} \hlstd{X2} \hlopt{+} \hlstd{X3,} \hlkwc{data} \hlstd{= neuraldat,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.
-Although useful for viewing the basic structure, the output is visually minimal and does not include extensive options for customization.
+Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.
 
-Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes many options to customize plot aesthetics.
-A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respecitively.
+Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a network as a neural interpretation diagram \citep{Ozesmi99} and includes several options to customize plot aesthetics.
+A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to view the thickness and color of the weight connections based on mangnitude and sign, respectively.
 The default settings plot positive weights between layers as black lines and negative weights as grey lines.
-Line thickness is in proportion to relative magnitude of each weight.
-The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables.
-One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn.
-The output layer is plotted last with nodes labeled as O1 through On.
-Bias nodes connected to the hidden and output layers are also shown.
-Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.
+Line thickness is in proportion to absolute magnitude of each weight (\cref{fig:plotnet}).
 
@@
 \begin{alltt}
-\hlcom{# create a model using the nnet package}
-\hlcom{# neuraldat is included with NeuralDat}
+\hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
+\hlkwd{plotnet}\hlstd{(mod3,} \hlkwc{nid} \hlstd{= F,} \hlkwc{circle_col} \hlstd{=} \hlstr{'grey'}\hlstd{,} \hlkwc{bord_col} \hlstd{=} \hlstr{'grey'}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(mod3,} \hlkwc{rel_rsc} \hlstd{=} \hlnum{8}\hlstd{)}
 \end{alltt}
-\end{kframe}
+\end{kframe}\begin{figure}[!ht]
 
-A primary network and a skip layer network can be plotted for nnet models with a skip layer connection.
+{\centering \includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-1} 
+\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-2}
+
+}
+
+\caption[Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the neural interpretation diagram (right)]{Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the neural interpretation diagram (right).
+Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
+Default options are changed for the example.}\label{fig:plotnet}
+\end{figure}
+
+A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection.
 The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}.
@@
 
-Pruned networks in RSNNS.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# create a model with a skip layer}
+\hlstd{modskip} \hlkwb{<-} \hlkwd{nnet}\hlstd{(Y1} \hlopt{~} \hlstd{X1} \hlopt{+} \hlstd{X2} \hlopt{+} \hlstd{X3,} \hlkwc{data} \hlstd{= neuraldat,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+
+\hlcom{# plot}
+\hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
+\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(modskip)}
+\end{alltt}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-1} 
+\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-2}
+
+}
+
+\caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
+The skip layer (left) and primary neural network (right) can be viewed separately with \code{plotnet}.}\label{fig:plotnet_skip}
+\end{figure}
+
+\end{knitrout}
+
+The \pkg{RSNNS} package provides several algorithms that can be used to prune connections in a neural network.
+This approach is useful to identify and remove connection weights between layers that do contribute to the predictive performance of the network, which allows for a potential increase in power by increasing the degrees of freedom.
+The \pkg{RSNNS} package implements
 

### 2015-09-28 nnt_manu.Rnw
emergency commit power outage!!!

@@
 par(mar = c(0, 0, 0, 0))
-plotnet(modskip, skip = TRUE)
-plotnet(modskip)
+plotnet(modskip, skip = TRUE, rel_rsc = 8)
+plotnet(modskip, rel_rsc = 8)
 @
 
-The \pkg{RSNNS} package provides several algorithms that can be used to prune connections in a neural network.
-This approach is useful to identify and remove connection weights between layers that do contribute to the predictive performance of the network, which allows for a potential increase in power by increasing the degrees of freedom.
-The \pkg{RSNNS} package implements
+The \pkg{RSNNS} package provides several algorithms that can be used to prune connections or nodes in a neural network.
+This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
+Pruning has the benefit of a potential increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
+Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
+The \code{plotnet} function includes options to display pruned connections that are removed using any of the three methods for removing weighted connections.
+
+<<plotprune, eval = F, echo = T, fig.height = 7, fig.width = 7, out.width = '\\textwidth'>>=
+
+# pruned model using code from RSSNS pruning demo
+pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+ no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, init_matrix_value = 1e-6,
+ input_pruning = TRUE, hidden_pruning = TRUE)
+mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+ pruneFuncParams = pruneFuncParams)
+
+plotnet(mod)
+plotnet(mod, prune_col = 'lightblue')
+@
 

### 2015-09-28 nnt_manu.tex
emergency commit power outage!!!

@@
 \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlkwd{plotnet}\hlstd{(modskip)}
+\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rel_rsc} \hlstd{=} \hlnum{8}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{rel_rsc} \hlstd{=} \hlnum{8}\hlstd{)}
 \end{alltt}
@@
 
-The \pkg{RSNNS} package provides several algorithms that can be used to prune connections in a neural network.
-This approach is useful to identify and remove connection weights between layers that do contribute to the predictive performance of the network, which allows for a potential increase in power by increasing the degrees of freedom.
-The \pkg{RSNNS} package implements
+The \pkg{RSNNS} package provides several algorithms that can be used to prune connections or nodes in a neural network.
+This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
+Pruning has the benefit of a potential increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
+Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
+The \code{plotnet} function includes options to display pruned connections that are removed using any of the three methods for removing weighted connections.
+
+\begin{kframe}
+\begin{alltt}
+\hlcom{# pruned model using code from RSSNS pruning demo}
+\hlstd{pruneFuncParams} \hlkwb{<-} \hlkwd{list}\hlstd{(}\hlkwc{max_pr_error_increase} \hlstd{=} \hlnum{10.0}\hlstd{,} \hlkwc{pr_accepted_error} \hlstd{=} \hlnum{1.0}\hlstd{,}
+ \hlkwc{no_of_pr_retrain_cycles} \hlstd{=} \hlnum{1000}\hlstd{,} \hlkwc{min_error_to_stop} \hlstd{=} \hlnum{0.01}\hlstd{,} \hlkwc{init_matrix_value} \hlstd{=} \hlnum{1e-6}\hlstd{,}
+ \hlkwc{input_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{hidden_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlstd{mod} \hlkwb{<-} \hlkwd{mlp}\hlstd{(x, y,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{pruneFunc} \hlstd{=} \hlstr{"OptimalBrainSurgeon"}\hlstd{,}
+ \hlkwc{pruneFuncParams} \hlstd{= pruneFuncParams)}
+
+\hlkwd{plotnet}\hlstd{(mod)}
+\hlkwd{plotnet}\hlstd{(mod,} \hlkwc{prune_col} \hlstd{=} \hlstr{'lightblue'}\hlstd{)}
+\end{alltt}
+\end{kframe}
 

### 2015-09-28 nnt_manu.Rnw
plot section done

@@
 <<results = 'hide'>>=
-# set random seed for initial inputs during training
+# set random seed for initial weights for training
 set.seed(123)
@@
 Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
-The \code{plotnet} function includes options to display pruned connections that are removed using any of the three methods for removing weighted connections.
+The \code{plotnet} function can be used to plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
 
-<<plotprune, eval = F, echo = T, fig.height = 7, fig.width = 7, out.width = '\\textwidth'>>=
-
-# pruned model using code from RSSNS pruning demo
+<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.
+The default plotting behavior of \\code{plotnet} is to omit pruned connections (left), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (right).'>>=
+# pruned model using RSNNS
 pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
- no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, init_matrix_value = 1e-6,
- input_pruning = TRUE, hidden_pruning = TRUE)
+  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
 mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
@@
 
+# plot, pruned connections are omitted (default) or included with prune_col
+par(mar = c(0, 0, 0, 0))
 plotnet(mod)
@@
 \subsection{Evaluating variable importance}
+
+The primary benefit of visualizing a neural network interpretation diagram is the ability to evaluate network archicture and the variation in weighted connections between the layers.
+For most networks,
 

### 2015-09-28 nnt_manu.tex
plot section done

@@
 \begin{alltt}
-\hlcom{# set random seed for initial inputs during training}
+\hlcom{# set random seed for initial weights for training}
 \hlkwd{set.seed}\hlstd{(}\hlnum{123}\hlstd{)}
@@
 Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
-The \code{plotnet} function includes options to display pruned connections that are removed using any of the three methods for removing weighted connections.
+The \code{plotnet} function can be used to plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
 
@@
 \begin{alltt}
-\hlcom{# pruned model using code from RSSNS pruning demo}
+\hlcom{# pruned model using RSNNS}
 \hlstd{pruneFuncParams} \hlkwb{<-} \hlkwd{list}\hlstd{(}\hlkwc{max_pr_error_increase} \hlstd{=} \hlnum{10.0}\hlstd{,} \hlkwc{pr_accepted_error} \hlstd{=} \hlnum{1.0}\hlstd{,}
- \hlkwc{no_of_pr_retrain_cycles} \hlstd{=} \hlnum{1000}\hlstd{,} \hlkwc{min_error_to_stop} \hlstd{=} \hlnum{0.01}\hlstd{,} \hlkwc{init_matrix_value} \hlstd{=} \hlnum{1e-6}\hlstd{,}
- \hlkwc{input_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{hidden_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+  \hlkwc{no_of_pr_retrain_cycles} \hlstd{=} \hlnum{1000}\hlstd{,} \hlkwc{min_error_to_stop} \hlstd{=} \hlnum{0.01}\hlstd{,}
+  \hlkwc{init_matrix_value} \hlstd{=} \hlnum{1e-6}\hlstd{,} \hlkwc{input_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{hidden_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{)}
 \hlstd{mod} \hlkwb{<-} \hlkwd{mlp}\hlstd{(x, y,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{pruneFunc} \hlstd{=} \hlstr{"OptimalBrainSurgeon"}\hlstd{,}
@@
 
+\hlcom{# plot, pruned connections are omitted (default) or included with prune_col}
+\hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
 \hlkwd{plotnet}\hlstd{(mod)}
@@
 \end{alltt}
-\end{kframe}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=0.48\textwidth]{figs/plotprune-1} 
+\includegraphics[width=0.48\textwidth]{figs/plotprune-2}
+
+}
+
+\caption{A pruned neural network using the ``Optimal Brain Surgeon'' algorithm described in \citet{Zell98}.
+The default plotting behavior of \code{plotnet} is to omit pruned connections (left), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (right).}\label{fig:plotprune}
+\end{figure}
 

### 2015-09-29 nnt_manu.Rnw
almost done with var importance

@@
 \usepackage{cleveref}
-\acrodef{mlp}[mlp]{multilayer perceptron}
+\acrodef{mlp}[MLP]{multilayer perceptron}
+\acrodef{nid}[NID]{neural interpretation diagram}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The increasing quantity of information and computational capacity to address relevant research questions has contributed to the growth of data science as a legitimate field of study.
 Data science is a relatively new paradigm of analysis that focuses on the synthesis of unstructured information from multiple sources to identify patterns or trends `born from the data' \citep{Kelling09}.
@@
 Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}).
-Morever, disciplines that have historically been limited by data quantity, such as ecological studies across broad temporal and spatial scales, have also realized the importance of data intensive approaches of analysis given the improved ability to acquire information (e.g., \citet{Swanson15}).
+Morever, disciplines that have historically been limited by data quantity, such as ecological studies across broad temporal and spatial scales, have also realized the importance of data intensive approaches given the increasing use of novel techniques to acquire information (e.g., \citet{Swanson15}).
 Regardless of the discipline, quantitative methods that explicitly focus on inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
@@
 Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in data intensive science.
-The most popular form of neural network is the \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
-This model is typically used to predict the response of one or more variables given co-occurrence of one to many explanatory variables.
+The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
+This model is typically used to predict the response of one or more variables given one to many explanatory variables.
 The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through an iterative training process with the backpropation algorithm.
@@
 \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks, most of which were previously available but not commonly used.
-For example, \citet{Olden02} describe neural interpretation diagrams for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
+For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
 These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.
@@
 
-Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a network as a neural interpretation diagram \citep{Ozesmi99} and includes several options to customize plot aesthetics.
-A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to view the thickness and color of the weight connections based on mangnitude and sign, respectively.
+Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a network as a \acl{nid} \citep{Ozesmi99} and includes several options to customize plot aesthetics.
+A \ac{nid} is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to view the thickness and color of the weight connections based on mangnitude and sign, respectively.
 The default settings plot positive weights between layers as black lines and negative weights as grey lines.
@@
 
-<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (left) and using the neural interpretation diagram (right).
+<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (left) and using the \\acl{nid} (right).
 Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
@@
 plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
-plotnet(mod3, rel_rsc = 8)
+plotnet(mod3)
 @
@@
 par(mar = c(0, 0, 0, 0))
-plotnet(modskip, skip = TRUE, rel_rsc = 8)
-plotnet(modskip, rel_rsc = 8)
+plotnet(modskip, skip = TRUE)
+plotnet(modskip)
 @
@@
 par(mar = c(0, 0, 0, 0))
-plotnet(mod)
-plotnet(mod, prune_col = 'lightblue')
+plotnet(mod, rel_rsc = c(3, 8))
+plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
 @
@@
 
-The primary benefit of visualizing a neural network interpretation diagram is the ability to evaluate network archicture and the variation in weighted connections between the layers.
-For most networks,
+The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in weighted connections between the layers.
+Although useful as a general tool, the \ac{nid} is impractical for evaluating most networks given the amount of weighted connections.
+Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.
+Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's conection weights algorithm \citep{Olden04}.
 
-The \code{garson} function uses Garson's algorithm to evaluate relative variable importance.
-This function identifies the relative importance of explanatory variables for a single response variable by deconstructing the model weights.
-The importance of each variable can be determined by identifying all weighted connections between the layers in the network.
-That is, all weights connecting the specific input node that pass through the hidden layer to the response variable are identified.
-This is repeated for all other explanatory variables until a list of all weights that are specific to each input variable is obtained.
-The connections are tallied for each input node and scaled relative to all other inputs.
-A single value is obtained for each explanatory variable that describes the relationship with the response variable in the model.
-The results indicate relative importance as the absolute magnitude from zero to one.
-The function cannot be used to evaluate the direction of the response.
-Only neural networks with one hidden layer and one output node can be evaluated.
+Garson's algorithm was originally described by \citet{Garson91} and further modified by \citet{Goh95}.
+The \code{garson} function is an implementation of the method described in the appendix of \citet{Goh95}.
+This method identifies the relative importance of each variable as an absolute magnitude by deconstructing all weighted connections between the layers in the network.
+For each input node, all weights connecting an input through the hidden layer to the response variable are identified to return a list of all weights specific to each input variable.
+A summed product of the connections for each input node are then scaled relative to all other inputs.
+A value for each input node indicates relative importance as the absolute magnitude from zero to one.
+The method is limited in that the direction of the response cannot be determined and only neural networks with one hidden layer and one output node can be evaluated.
 
-The \code{olden} function is an alternative and more flexible approach to evaluate variable importance.
-The function calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output neuron and sums the product across all hidden neurons.
-An advantage of this approach is the relative contributions of each connection weight are maintained in terms of both magnitude and sign as compared to Garson's algorithm which only considers the absolute magnitude.
+The \code{olden} function is a more flexible approach to evaluate variable importance using the connection weights algorithm described in \citet{Olden04}.
+This method calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output node and sums the product across all hidden nodes.
+An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign, as compared to Garson's algorithm which only considers absolute magnitude.
 For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect whereas Garson's algorithm may provide misleading results based on the absolute magnitude.
 An additional advantage is that Olden's algorithm is capable of evaluating neural networks with multiple hidden layers and response variables.
-The importance values assigned to each variable are in units that are based directly on the summed product of the connection weights.
-The actual values should only be interpreted based on relative sign and magnitude between explanatory variables.
-Comparisons between different models should not be made.
+The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.
 
-Issues with different indications of variable importance as a model is refit...
+Both functions have similar implementations and require only a model object as input.
+The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model.
+The plot aesthetics are based on internal code from the function but can be changed using conventional syntax for \pkg{ggplot2}.
+The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.
+
+<<plotimp, fig.show = 'hold', fig.height = 2, fig.width = 4, out.width = '0.45\\textwidth', fig.cap = 'Variable importance'>>=
+# garson and olden functions for each model
+garson(mod1)
+olden(mod1)
+garson(mod2)
+olden(mod2)
+garson(mod3)
+olden(mod3)
+@
 
@@
 \section[Conclusions]{Conclusions}
+Issues with different indications of variable importance as a model is refit.
 

### 2015-09-29 nnt_manu.tex
almost done with var importance

@@
 \hlkwd{plotnet}\hlstd{(mod3,} \hlkwc{nid} \hlstd{= F,} \hlkwc{circle_col} \hlstd{=} \hlstr{'grey'}\hlstd{,} \hlkwc{bord_col} \hlstd{=} \hlstr{'grey'}\hlstd{)}
-\hlkwd{plotnet}\hlstd{(mod3,} \hlkwc{rel_rsc} \hlstd{=} \hlnum{8}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(mod3)}
 \end{alltt}
@@
 
-\caption[Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the neural interpretation diagram (right)]{Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the neural interpretation diagram (right).
+\caption[Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the \acl{nid} (right)]{Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the \acl{nid} (right).
 Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
@@
 \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rel_rsc} \hlstd{=} \hlnum{8}\hlstd{)}
-\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{rel_rsc} \hlstd{=} \hlnum{8}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(modskip)}
 \end{alltt}
@@
 \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(mod)}
-\hlkwd{plotnet}\hlstd{(mod,} \hlkwc{prune_col} \hlstd{=} \hlstr{'lightblue'}\hlstd{)}
+\hlkwd{plotnet}\hlstd{(mod,} \hlkwc{rel_rsc} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{8}\hlstd{))}
+\hlkwd{plotnet}\hlstd{(mod,} \hlkwc{prune_col} \hlstd{=} \hlstr{'lightblue'}\hlstd{,} \hlkwc{rel_rsc} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{8}\hlstd{))}
 \end{alltt}
@@
 
-The \code{garson} function uses Garson's algorithm to evaluate relative variable importance.
-This function identifies the relative importance of explanatory variables for a single response variable by deconstructing the model weights.
-The importance of each variable can be determined by identifying all weighted connections between the layers in the network.
-That is, all weights connecting the specific input node that pass through the hidden layer to the response variable are identified.
-This is repeated for all other explanatory variables until a list of all weights that are specific to each input variable is obtained.
-The connections are tallied for each input node and scaled relative to all other inputs.
-A single value is obtained for each explanatory variable that describes the relationship with the response variable in the model.
-The results indicate relative importance as the absolute magnitude from zero to one.
-The function cannot be used to evaluate the direction of the response.
-Only neural networks with one hidden layer and one output node can be evaluated.
+The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in weighted connections between the layers.
+Although useful as a general tool, the \ac{nid} is impractical for evaluating most networks given the amount of weighted connections.
+Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.
+Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's conection weights algorithm \citep{Olden04}.
 
-The \code{olden} function is an alternative and more flexible approach to evaluate variable importance.
-The function calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output neuron and sums the product across all hidden neurons.
-An advantage of this approach is the relative contributions of each connection weight are maintained in terms of both magnitude and sign as compared to Garson's algorithm which only considers the absolute magnitude.
+Garson's algorithm was originally described by \citet{Garson91} and further modified by \citet{Goh95}.
+The \code{garson} function is an implementation of the method described in the appendix of \citet{Goh95}.
+This method identifies the relative importance of each variable as an absolute magnitude by deconstructing all weighted connections between the layers in the network.
+For each input node, all weights connecting an input through the hidden layer to the response variable are identified to return a list of all weights specific to each input variable.
+A summed product of the connections for each input node are then scaled relative to all other inputs.
+A value for each input node indicates relative importance as the absolute magnitude from zero to one.
+The method is limited in that the direction of the response cannot be determined and only neural networks with one hidden layer and one output node can be evaluated.
+
+The \code{olden} function is a more flexible approach to evaluate variable importance using the connection weights algorithm described in \citet{Olden04}.
+This method calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output node and sums the product across all hidden nodes.
+An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign, as compared to Garson's algorithm which only considers absolute magnitude.
 For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect whereas Garson's algorithm may provide misleading results based on the absolute magnitude.
 An additional advantage is that Olden's algorithm is capable of evaluating neural networks with multiple hidden layers and response variables.
-The importance values assigned to each variable are in units that are based directly on the summed product of the connection weights.
-The actual values should only be interpreted based on relative sign and magnitude between explanatory variables.
-Comparisons between different models should not be made.
+The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.
 
-Issues with different indications of variable importance as a model is refit...
+Both functions have similar implementations and require only a model object as input.
+The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model.
+The plot aesthetics are based on internal code from the function but can be changed using conventional syntax for \pkg{ggplot2}.
+The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.
+
+\begin{kframe}
+\begin{alltt}
+\hlcom{# garson and olden functions for each model}
+\hlkwd{garson}\hlstd{(mod1)}
+\hlkwd{olden}\hlstd{(mod1)}
+\hlkwd{garson}\hlstd{(mod2)}
+\hlkwd{olden}\hlstd{(mod2)}
+\hlkwd{garson}\hlstd{(mod3)}
+\hlkwd{olden}\hlstd{(mod3)}
+\end{alltt}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=0.45\textwidth]{figs/plotimp-1} 
+\includegraphics[width=0.45\textwidth]{figs/plotimp-2} 
+\includegraphics[width=0.45\textwidth]{figs/plotimp-3} 
+\includegraphics[width=0.45\textwidth]{figs/plotimp-4} 
+\includegraphics[width=0.45\textwidth]{figs/plotimp-5} 
+\includegraphics[width=0.45\textwidth]{figs/plotimp-6}
+
+}
+
+\caption[Variable importance]{Variable importance}\label{fig:plotimp}
+\end{figure}
 

### 2015-09-30 nnt_manu.Rnw
better figures with subcaptions

@@
 \usepackage{cleveref}
+\usepackage{subfig}
+
 \acrodef{mlp}[MLP]{multilayer perceptron}
@@
 
-<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (left) and using the \\acl{nid} (right).
+<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (\\ref{fig:plotnet1}) and using the \\acl{nid} (\\ref{fig:plotnet2}).
 Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
-Default options are changed for the example.'>>=
+Default options are changed for the example.', fig.subcap = c('', '')>>=
 par(mar = c(0, 0, 0, 0))
@@
 <<plotnet_skip, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
-The skip layer (left) and primary neural network (right) can be viewed separately with \\code{plotnet}.', results = 'hide'>>=
+The skip layer (\\ref{fig:plotnet_skip1}) and primary neural network (\\ref{fig:plotnet_skip2}) can be viewed separately with \\code{plotnet}.', results = 'hide', fig.subcap = c('', '')>>=
 # create a model with a skip layer
@@
 <<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.
-The default plotting behavior of \\code{plotnet} is to omit pruned connections (left), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (right).'>>=
+The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
 # pruned model using RSNNS
@@
 Both functions have similar implementations and require only a model object as input.
-The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model.
+The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).
 The plot aesthetics are based on internal code from the function but can be changed using conventional syntax for \pkg{ggplot2}.
@@
 
-<<plotimp, fig.show = 'hold', fig.height = 2, fig.width = 4, out.width = '0.45\\textwidth', fig.cap = 'Variable importance'>>=
+<<eval = F>>=
 # garson and olden functions for each model
@@
 @
+
+\begin{figure}
+\centering
+\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
+<<plotimp-1, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
+garson(mod1)
+@
+\label{fig:plotimp1}
+}
+\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
+<<plotimp-2, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
+olden(mod1)
+@
+\label{fig:plotimp2}
+}
+
+\subfloat[][Model 2 (\code{nn}), \code{garson}]{
+<<plotimp-3, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
+garson(mod2)
+@
+\label{fig:plotimp3}
+}
+\subfloat[][Model 2 (\code{nn}), \code{olden}]{
+<<plotimp-4, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
+olden(mod2)
+@
+\label{fig:plotimp4}
+}
+
+\subfloat[][Model 3 (\code{nnet}), \code{garson}]{
+<<plotimp-5, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
+garson(mod3)
+@
+\label{fig:plotimp5}
+}
+\subfloat[][Model 3 (\code{nnet}), \code{olden}]{
+<<plotimp-6, fig.height = 2, fig.width =4, out.width = '0.47\\textwidth', echo = F>>=
+olden(mod3)
+@
+\label{fig:plotimp6}
+}
+\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \citealt{Olden04}).
+Garson's algorithm shows importance as absolute values from 0--1 whereas Olden's algorithm preserves sign and magnitude.
+Importance values for Olden's algorithm are directly from the summed product of model weights and should only be evaluated based on the sign and relative magnitudes between variables.}
+\label{fig:plotimp}
+\end{figure}
 

### 2015-09-30 nnt_manu.tex
better figures with subcaptions

@@
 
-{\centering \includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-1} 
-\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-2}
+{\centering \subfloat[\label{fig:plotnet1}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-1} }
+\subfloat[\label{fig:plotnet2}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-2} }
 
@@
 
-\caption[Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the \acl{nid} (right)]{Examples from the \code{plotnet} function showing neural networks as a standard graphic (left) and using the \acl{nid} (right).
+\caption{Examples from the \code{plotnet} function showing neural networks as a standard graphic (\ref{fig:plotnet1}) and using the \acl{nid} (\ref{fig:plotnet2}).
 Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
@@
 
-{\centering \includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-1} 
-\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-2}
+{\centering \subfloat[\label{fig:plotnet_skip1}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-1} }
+\subfloat[\label{fig:plotnet_skip2}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-2} }
 
@@
 \caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
-The skip layer (left) and primary neural network (right) can be viewed separately with \code{plotnet}.}\label{fig:plotnet_skip}
+The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet}.}\label{fig:plotnet_skip}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.48\textwidth]{figs/plotprune-1} 
-\includegraphics[width=0.48\textwidth]{figs/plotprune-2}
+{\centering \subfloat[\label{fig:plotprune1}]{\includegraphics[width=0.48\textwidth]{figs/plotprune-1} }
+\subfloat[\label{fig:plotprune2}]{\includegraphics[width=0.48\textwidth]{figs/plotprune-2} }
 
@@
 \caption{A pruned neural network using the ``Optimal Brain Surgeon'' algorithm described in \citet{Zell98}.
-The default plotting behavior of \code{plotnet} is to omit pruned connections (left), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (right).}\label{fig:plotprune}
+The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}\label{fig:plotprune}
 \end{figure}
@@
 \end{alltt}
-\end{kframe}\begin{figure}[!ht]
+\end{kframe}
 
-{\centering \includegraphics[width=0.45\textwidth]{figs/plotimp-1} 
-\includegraphics[width=0.45\textwidth]{figs/plotimp-2} 
-\includegraphics[width=0.45\textwidth]{figs/plotimp-3} 
-\includegraphics[width=0.45\textwidth]{figs/plotimp-4} 
-\includegraphics[width=0.45\textwidth]{figs/plotimp-5} 
-\includegraphics[width=0.45\textwidth]{figs/plotimp-6}
+\begin{figure}
+\centering
+\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
+
+{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-1-1}
 
@@
 
-\caption[Variable importance]{Variable importance}\label{fig:plotimp}
+\label{fig:plotimp1}
+}
+\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
+
+{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-2-1}
+
+}
+
+\label{fig:plotimp2}
+}
+
+\subfloat[][Model 2 (\code{nn}), \code{garson}]{
+
+{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-3-1}
+
+}
+
+\label{fig:plotimp3}
+}
+\subfloat[][Model 2 (\code{nn}), \code{olden}]{
+
+{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-4-1}
+
+}
+
+\label{fig:plotimp4}
+}
+
+\subfloat[][Model 3 (\code{nnet}), \code{garson}]{
+
+{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-5-1}
+
+}
+
+\label{fig:plotimp5}
+}
+\subfloat[][Model 3 (\code{nnet}), \code{olden}]{
+
+{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-6-1}
+
+}
+
+\label{fig:plotimp6}
+}
+\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \citealt{Olden04}).
+Garson's algorithm shows importance as absolute values from 0--1 whereas Olden's algorithm preserves sign and magnitude.
+Importance values for Olden's algorithm are directly from the summed product of model weights and should only be evaluated based on the sign and relative magnitudes between variables.}
+\label{fig:plotimp}
 \end{figure}

### 2015-09-30 nnt_manu.Rnw
lekprofile section done

@@
 % knitr options
-<<setup,include = F,cache = F>>=
+<<setup, include = F, cache = F>>=
 library(knitr)
@@
 options(replace.assign = TRUE, digits = 1)
+
+# other libs
+library(ggplot2)
 @
@@
 The response variables are also standardized from zero to one.
-A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
+A common approach for data preprocessing prior to creating a neural network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
 Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
@@
 
-The Lek profile method is described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
-The profile method is fairly generic and can be extended to any statistical model in R with a predict method.
+An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
+The profile method is fairly generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method.
 However, it is one of few methods used to evaluate sensitivity in neural networks.
+The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
 
-The profile method can be used to evaluate the effect of explanatory variables by returning a plot of the predicted response across the range of values for each separate variable.
-The original profile method evaluated the effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum).
-This is implemented in in the function by creating a matrix of values for explanatory variables where the number of rows is the number of observations and the number of columns is the number of explanatory variables.
-All explanatory variables are held at their mean (or other constant value) while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
-This matrix (or data frame) is then used to predict values of the response variable from a fitted model object.
+The \code{lekprofile} function evaluates the effect of explanatory variables by returning a plot of the predicted response across the range of values for each variable.
+A critical aspect of the method is the remaining explanatory variables that are held constant when separately evaluating each input variable.
+The \code{lekprofile} function provides two options for determining constant values of unevaluated explanatory variables.
+The first option following the original profile method evaluates effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}).
+This is implemented by creating a matrix of values for explanatory variables where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables.
+All explanatory variables are held at their median (or other constant value), while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
+This matrix is then used to predict values of the response variable from a fitted model object.
 This is repeated for each explanatory variable to obtain all response curves.
-Values passed to \code{split_vals} must range from zero to one to define the quantiles for holding unevaluated explanatory variables.
+Constant values are defined by passing one or more numeric values in the range 0--1 to the \code{group\_vals} that define the quantiles, with the default as minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
 
-An alternative implementation of the profile method is to group the unevaluated explanatory variables using groupings defined by the statistical properties of the data.
-Covariance among predictors may present unlikely scenarios if holding all unevaluated variables at the same level.
-To address this issue, the function provides an option to hold unevalutaed variable at mean values defined by natural clusters in the data.
-kmeans clustering is used on the input data.frame of explanatory variables if the argument passed to \code{split_vals} is an integer value greater than one.
-The centers of the clusters are then used as constant values for the unevaluated variables.
-An arbitrary grouping scheme can also be passed to \code{split_vals} as a data.frame where the user can specify exact values for holding each value constant (see the examples).
-Examples in \citet{Beck14a} show this...
+An alternative implementation of \code{lekprofile} is to group the unevaluated explanatory variables using natural groupings defined by statistical properties of the data.
+Covariance among predictors may present unlikely scenarios if all unevaluated variables at held at the same level (e.g., high values for one variable may be unlikely to co-occur with high values for a second variable).
+To address this issue, a second option is available to hold unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
+Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one.
+The centers (means) of the clusters are then used as constant values for the unevaluated variables.
+\citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.
+Lake clusters were identified given covariance among explanatory variables, such that holding explanatory variables at related values created more ecologically relevant response curves.
+Similar plots can be created with \code{lekprofile} as response curves (\cref{fig:plotlek_sens}) or to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
+For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights in each bar proportional to the constant values.
+<<eval = F>>=
+# lek profile, quantile evaluation
+lekprofile(mod3)
 
-For all plots, the legend with the 'splits' label indicates the colors that correspond to each group.
-The groups describe the values at which unevaluated explanatory variables were held constant, either as specific quantiles, group assignments based on clustering, or in the arbitrary grouping defined by the user.
-The constant values of each explanatory variable for each split can be viewed as a barplot by using \code{split_show = TRUE}.
+# lek profile, quantile groups 
+lekprofile(mod3, group_show = TRUE)
 
-Note that there is no predict method for neuralnet objects from the nn package.
-The lekprofile method for nn objects uses the nnet package to recreate the input model, which is then used for the sensitivity predictions.
-This approach only works for networks with one hidden layer.
+# lek profile, cluster evaluation
+lekprofile(mod3, group_vals = 6)
+
+# lek profile, cluster groups
+lekprofile(mod3, group_vals = 6, group_show = TRUE)
+@
+
+\begin{figure}
+\centering
+\subfloat[][Response with quantiles]{
+<<plotlek_sens1, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+lekprofile(mod3) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+@
+\label{fig:plotlek_sens1}
+}
+
+\subfloat[][Reponse with clusters]{
+<<plotlek_sens2, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+lekprofile(mod3, group_vals = 6) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+@
+\label{fig:plotlek_sens2}
+}s
+\caption{Sensitivity analysis of neural network using the Lek profile method to evaluate the effects of explanatory variables.
+\Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles and \cref{fig:plotlek_sens2} groups by cluster means from {\it k}-means clustering of all explanatory variables.
+Values at which explanatory variables are held constant in each plot are shown in \cref{fig:plotlek_bars1,fig:plotlek_bars2}, respectively.}
+\label{fig:plotlek_sens}
+\end{figure}
+
+%%%%
+\begin{figure}
+\centering
+\subfloat[][Quantile groupings]{
+<<plotlek_bars1, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+lekprofile(mod3, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+@
+\label{fig:plotlek_bars1}
+}
+
+\subfloat[][Cluster groupings]{
+<<plotlek_bars2, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+lekprofile(mod3, group_vals = 6, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+@
+\label{fig:plotlek_bars2}
+}
+\caption{Bar plots for values in each group of unevaluated explanatory variables in \cref{fig:plotlek_sens1,fig:plotlek_sens2}.
+\Cref{fig:plotlek_bars1} shows default quantile groupings set at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
+For example, variables are held at zero for group 1 (i.e., stacked bars with no height) for the minimum value, whereas group 6 holds variables at their maximum of one.
+\Cref{fig:plotlek_bars2} shows the cluster centers for each variable in each group identified from {\it k}-means clustering.}
+\label{fig:plotlek_bars}
+\end{figure}
 

### 2015-09-30 nnt_manu.tex
lekprofile section done

@@
 
-The Lek profile method is described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
-The profile method is fairly generic and can be extended to any statistical model in R with a predict method.
+An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
+The profile method is fairly generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method.
 However, it is one of few methods used to evaluate sensitivity in neural networks.
+The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
 
-The profile method can be used to evaluate the effect of explanatory variables by returning a plot of the predicted response across the range of values for each separate variable.
-The original profile method evaluated the effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum).
-This is implemented in in the function by creating a matrix of values for explanatory variables where the number of rows is the number of observations and the number of columns is the number of explanatory variables.
-All explanatory variables are held at their mean (or other constant value) while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
-This matrix (or data frame) is then used to predict values of the response variable from a fitted model object.
+The \code{lekprofile} function evaluates the effect of explanatory variables by returning a plot of the predicted response across the range of values for each variable.
+A critical aspect of the method is the remaining explanatory variables that are held constant when separately evaluating each input variable.
+The \code{lekprofile} function provides two options for determining constant values of unevaluated explanatory variables.
+The first option following the original profile method evaluates effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}).
+This is implemented by creating a matrix of values for explanatory variables where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables.
+All explanatory variables are held at their median (or other constant value), while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
+This matrix is then used to predict values of the response variable from a fitted model object.
 This is repeated for each explanatory variable to obtain all response curves.
-Values passed to \code{split_vals} must range from zero to one to define the quantiles for holding unevaluated explanatory variables.
+Constant values are defined by passing one or more numeric values in the range 0--1 to the \code{group\_vals} that define the quantiles, with the default as minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
 
-An alternative implementation of the profile method is to group the unevaluated explanatory variables using groupings defined by the statistical properties of the data.
-Covariance among predictors may present unlikely scenarios if holding all unevaluated variables at the same level.
-To address this issue, the function provides an option to hold unevalutaed variable at mean values defined by natural clusters in the data.
-kmeans clustering is used on the input data.frame of explanatory variables if the argument passed to \code{split_vals} is an integer value greater than one.
-The centers of the clusters are then used as constant values for the unevaluated variables.
-An arbitrary grouping scheme can also be passed to \code{split_vals} as a data.frame where the user can specify exact values for holding each value constant (see the examples).
-Examples in \citet{Beck14a} show this...
+An alternative implementation of \code{lekprofile} is to group the unevaluated explanatory variables using natural groupings defined by statistical properties of the data.
+Covariance among predictors may present unlikely scenarios if all unevaluated variables at held at the same level (e.g., high values for one variable may be unlikely to co-occur with high values for a second variable).
+To address this issue, a second option is available to hold unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
+Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one.
+The centers (means) of the clusters are then used as constant values for the unevaluated variables.
+\citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.
+Lake clusters were identified given covariance among explanatory variables, such that holding explanatory variables at related values created more ecologically relevant response curves.
+Similar plots can be created with \code{lekprofile} as response curves (\cref{fig:plotlek_sens}) or to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
+For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights in each bar proportional to the constant values.
+\begin{kframe}
+\begin{alltt}
+\hlcom{# lek profile, quantile evaluation}
+\hlkwd{lekprofile}\hlstd{(mod3)}
 
-For all plots, the legend with the 'splits' label indicates the colors that correspond to each group.
-The groups describe the values at which unevaluated explanatory variables were held constant, either as specific quantiles, group assignments based on clustering, or in the arbitrary grouping defined by the user.
-The constant values of each explanatory variable for each split can be viewed as a barplot by using \code{split_show = TRUE}.
+\hlcom{# lek profile, quantile groups }
+\hlkwd{lekprofile}\hlstd{(mod3,} \hlkwc{group_show} \hlstd{=} \hlnum{TRUE}\hlstd{)}
 
-Note that there is no predict method for neuralnet objects from the nn package.
-The lekprofile method for nn objects uses the nnet package to recreate the input model, which is then used for the sensitivity predictions.
-This approach only works for networks with one hidden layer.
+\hlcom{# lek profile, cluster evaluation}
+\hlkwd{lekprofile}\hlstd{(mod3,} \hlkwc{group_vals} \hlstd{=} \hlnum{6}\hlstd{)}
+
+\hlcom{# lek profile, cluster groups}
+\hlkwd{lekprofile}\hlstd{(mod3,} \hlkwc{group_vals} \hlstd{=} \hlnum{6}\hlstd{,} \hlkwc{group_show} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+\begin{figure}
+\centering
+\subfloat[][Response with quantiles]{
+
+{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_sens1-1}
+
+}
+
+\label{fig:plotlek_sens1}
+}
+
+\subfloat[][Reponse with clusters]{
+
+{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_sens2-1}
+
+}
+
+\label{fig:plotlek_sens2}
+}s
+\caption{Sensitivity analysis of neural network using the Lek profile method to evaluate the effects of explanatory variables.
+\Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles and \cref{fig:plotlek_sens2} groups by cluster means from {\it k}-means clustering of all explanatory variables.
+Values at which explanatory variables are held constant in each plot are shown in \cref{fig:plotlek_bars1,fig:plotlek_bars2}, respectively.}
+\label{fig:plotlek_sens}
+\end{figure}
+
+%%%%
+\begin{figure}
+\centering
+\subfloat[][Quantile groupings]{
+
+{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_bars1-1}
+
+}
+
+\label{fig:plotlek_bars1}
+}
+
+\subfloat[][Cluster groupings]{
+
+{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_bars2-1}
+
+}
+
+\label{fig:plotlek_bars2}
+}
+\caption{Bar plots for values in each group of unevaluated explanatory variables in \cref{fig:plotlek_sens1,fig:plotlek_sens2}.
+\Cref{fig:plotlek_bars1} shows default quantile groupings set at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
+For example, variables are held at zero for group 1 (i.e., stacked bars with no height) for the minimum value, whereas group 6 holds variables at their maximum of one.
+\Cref{fig:plotlek_bars2} shows the cluster centers for each variable in each group identified from {\it k}-means clustering.}
+\label{fig:plotlek_bars}
+\end{figure}
 

### 2015-10-01 nnt_manu.Rnw
messy conclusions

@@
 
+\acrodef{CRAN}[CRAN]{Comprehensive \proglang{R} Archive Network}
 \acrodef{mlp}[MLP]{multilayer perceptron}
@@
 The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) briefly describe similarities and differences between existing neural network packages in \proglang{R}, and 3) describe the theory and application of the primary functions in the \pkg{NeuralNetTools} package.
-The package is currently available on CRAN, whereas the development version is maintained as a GitHub repository.
+The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.
 
@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the CRAN download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
@@
 
-The stable release of \pkg{NeurelNetTools} can be installed from CRAN and the development version can be installed from GitHub:
+The stable release of \pkg{NeurelNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
 
@@
 \subfloat[][Response with quantiles]{
-<<plotlek_sens1, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+<<plotlek_sens1, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
 lekprofile(mod3) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
@@
 
-\subfloat[][Reponse with clusters]{
-<<plotlek_sens2, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+\subfloat[][Response with clusters]{
+<<plotlek_sens2, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
 lekprofile(mod3, group_vals = 6) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
@@
 }s
-\caption{Sensitivity analysis of neural network using the Lek profile method to evaluate the effects of explanatory variables.
+\caption{Sensitivity analysis ofa  neural network using the Lek profile method to evaluate the effects of explanatory variables.
 \Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles and \cref{fig:plotlek_sens2} groups by cluster means from {\it k}-means clustering of all explanatory variables.
@@
 \subfloat[][Quantile groupings]{
-<<plotlek_bars1, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+<<plotlek_bars1, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
 lekprofile(mod3, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
@@
 \subfloat[][Cluster groupings]{
-<<plotlek_bars2, fig.height = 2.5, fig.width = 8, out.width = '0.8\\textwidth', echo = F>>=
+<<plotlek_bars2, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
 lekprofile(mod3, group_vals = 6, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
@@
 
-\section[Future development]{Future development}
+\section[Applied example]{Applied example}
 
-\section[Conclusions]{Conclusions}
 Issues with different indications of variable importance as a model is refit.
@@
 
+\section[Conclusions]{Conclusions}
+
+The \pkg{NeuralNetTools} package provides a simple approach that improves the quality of information obtained from a feed-forward \ac{mlp} neural network.
+Functions in the package can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis \code{lekprofile}.
+Methods are available for the most downloaded \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods will be made available based on future popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
+An additional objective of the package is to dispel the myth that supervised neural networks for prediction are `black boxes' that provide no information about underlying relationships between the variables \citep{Olden02}.
+Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation of the technique has many parallels with traditional modelling techniques that allow insight into causation.
+For example, network weights are partially analagous to intercept and slope parameters in a standard regression model such that the estimates can be used to characterize variable interactions.
+More over, the model fitting process requires the minimization of an objective function such that conventional techniques to evaluate model sensitivity or performance that apply to other models can be used with neural networks.
+Common performance metrics can be used to select an optimal model, whereas functions in \pkg{NeuralNetTools} can provide a practical means of post-hoc assessment.
+
+A more compelling issue is when and how to apply neural networks given alternative methods for data-intensive exploration.
+A common opinion is that the potential for insight is positively related to model complexity and the relaxation of assumptions required for more constrained methods ().
+Indeed, the `neural' component of neural networks is a reference to the network of impulses in the human brain that lead to learning.
+This perception has likely contributed to the popularity of the method, although the appeal may be misinformed in practice depending on the dataset or question of interest.
+Numerous examples have shown that the \ac{mlp} network may provide comparable insight into variable relationships or predicictive performance as more conventional methods.
+As such, the neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques.
+For example, an appropriate method should be chosen based on power given the sample size, expected interactions between variables, distributional forms of the response, and other relevant concerns that are considered during exploratory data analysis.
+The true value of the \pkg{NeuralNetTools} package is a set of analysis tools that can inform a more balanced assessment given the alternative and complementary methods for data-intensive exploration.
+
 \section[Acknowledgments]{Acknowledgments}
+
+I would like to thank Dr.
+Bruce Wilson and Dr.
+Sanford Weisberg of the University of Minnesota for general guidance during the development of this package.
+Contributions and suggestions from online users have also greatly improved the utility of the package.
 

### 2015-10-01 nnt_manu.tex
messy conclusions

@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the CRAN download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 93\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 93\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_sens1-1}
+{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_sens1-1}
 
@@
 
-\subfloat[][Reponse with clusters]{
+\subfloat[][Response with clusters]{
 
-{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_sens2-1}
+{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_sens2-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_bars1-1}
+{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_bars1-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figs/plotlek_bars2-1}
+{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_bars2-1}
 

### 2015-10-02 nnt_manu.Rnw
revising conclusions, started applied example

@@
 
-Issues with different indications of variable importance as a model is refit.
+Although \pkg{NeuralnetTools} provides several methods to extract information from a neural network, it does not provide explicit guidance for developing the initial model.
+A more challenging aspect is understanding the effects of network architecture, use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting during model development.
+A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to introduce a `best practices' approach to neural network development.
+The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive application of neural networks in practice.
+Several texts provide explicit guidelines on developing neural network models \cite[e.g.,][]{Ripley96, Lek00, Maier00}.
+In general, the following should be considered during model development:
+\begin{itemize}
+\item Initial data pre-processing to normalize inputs and standardize response,
+\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, pruning weights or inputs,
+\item Split of dataset into a training and test dataset, e.g., 2:1, 3:1, leave-one-out, etc., 
+\item Initial starting weights for the back-propagation algorithm, and
+\item Criteria for stopping model training iterations, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
+\end{itemize}
 
-A cautionary note about the `optimal network' and reproducibility of results.
-Note that most literature sources suggest variable standardization prior to using a model yet none of the existing packages in R provide this functionality as a default option (verify).
+<<eval = F>>=
+library(nycflights13)
+library(dplyr)
+
+# data preprocessing
+# select UA carrior for month of december and relevant variables
+# normalize and standardize variables
+tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
+  select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
+  mutate_each(funs(scale), -arr_delay) %>% 
+  mutate_each(funs(as.numeric), -arr_delay) %>% 
+  mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
+  data.frame
+
+# create two models
+# random starting weights are different for each
+ex1 <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod)  
+ex2 <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod)
+
+# view olden results 
+olden(ex1)
+olden(ex2)
+@
 
@@
 The \pkg{NeuralNetTools} package provides a simple approach that improves the quality of information obtained from a feed-forward \ac{mlp} neural network.
-Functions in the package can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis \code{lekprofile}.
-Methods are available for the most downloaded \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods will be made available based on future popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
-An additional objective of the package is to dispel the myth that supervised neural networks for prediction are `black boxes' that provide no information about underlying relationships between the variables \citep{Olden02}.
-Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation of the technique has many parallels with traditional modelling techniques that allow insight into causation.
-For example, network weights are partially analagous to intercept and slope parameters in a standard regression model such that the estimates can be used to characterize variable interactions.
-More over, the model fitting process requires the minimization of an objective function such that conventional techniques to evaluate model sensitivity or performance that apply to other models can be used with neural networks.
-Common performance metrics can be used to select an optimal model, whereas functions in \pkg{NeuralNetTools} can provide a practical means of post-hoc assessment.
+Functions in the package can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).
+Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods will be made available based on future popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
+A primary objective of the package is to dispel the myth that supervised neural networks for prediction are `black boxes' that provide no information about underlying relationships between the variables \citep{Paruelo97,Olden02}.
+Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation has many parallels with traditional modelling techniques that can allow insight into causation.
+Moreover, the model fitting process requires the minimization of an objective function such that conventional techniques to evaluate model sensitivity or performance (e.g., cross-validation) can be used with neural networks.
+Functions in \pkg{NeuralNetTools} can facilitate the selection of the optimal network architecture or can be used for post-hoc assessment.
 
 A more compelling issue is when and how to apply neural networks given alternative methods for data-intensive exploration.
-A common opinion is that the potential for insight is positively related to model complexity and the relaxation of assumptions required for more constrained methods ().
-Indeed, the `neural' component of neural networks is a reference to the network of impulses in the human brain that lead to learning.
-This perception has likely contributed to the popularity of the method, although the appeal may be misinformed in practice depending on the dataset or question of interest.
-Numerous examples have shown that the \ac{mlp} network may provide comparable insight into variable relationships or predicictive performance as more conventional methods.
-As such, the neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques.
-For example, an appropriate method should be chosen based on power given the sample size, expected interactions between variables, distributional forms of the response, and other relevant concerns that are considered during exploratory data analysis.
+The popularity of the \ac{mlp} neural network is partially the cause of misperceptions and generalizations of their benefits as modelling tools \citep{Burke97}.
+Indeed, the `neural' component of neural networks is advertised as a mathematical representation of the network of synaptic impulses in the human brain, although the former is of course an over-simplification of the latter.
+Several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.
+As such, the neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques, with particular emphasis on the needs of a given dataset or research question.
+For example, an appropriate method should be chosen based on power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant concerns that are routinely considered during exploratory data analysis.
 The true value of the \pkg{NeuralNetTools} package is a set of analysis tools that can inform a more balanced assessment given the alternative and complementary methods for data-intensive exploration.

### 2015-10-02 nnt_manu.tex
revising conclusions, started applied example

@@
 
-Issues with different indications of variable importance as a model is refit.
+Although \pkg{NeuralnetTools} provides several methods to extract information from a neural network, it does not provide explicit guidance for developing the initial model.
+A more challenging aspect is understanding the effects of network architecture, use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting during model development.
+A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to introduce a `best practices' approach to neural network development.
+The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive application of neural networks in practice.
+Several texts provide explicit guidelines on developing neural network models \cite[e.g.,][]{Ripley96, Lek00, Maier00}.
+In general, the following should be considered during model development:
+\begin{itemize}
+\item Initial data pre-processing to normalize inputs and standardize response,
+\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, pruning weights or inputs,
+\item Split of dataset into a training and test dataset, e.g., 2:1, 3:1, leave-one-out, etc., 
+\item Initial starting weights for the back-propagation algorithm, and
+\item Criteria for stopping model training iterations, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
+\end{itemize}
 
-A cautionary note about the `optimal network' and reproducibility of results.
-Note that most literature sources suggest variable standardization prior to using a model yet none of the existing packages in R provide this functionality as a default option (verify).
+\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(nycflights13)}
+\hlkwd{library}\hlstd{(dplyr)}
+
+\hlcom{# data preprocessing}
+\hlcom{# select UA carrior for month of december and relevant variables}
+\hlcom{# normalize and standardize variables}
+\hlstd{tomod} \hlkwb{<-} \hlkwd{filter}\hlstd{(flights, month} \hlopt{==} \hlnum{12} \hlopt{&} \hlstd{carrier} \hlopt{==} \hlstr{'UA'}\hlstd{)} \hlopt{%>%}
+  \hlkwd{select}\hlstd{(arr_delay, dep_delay, dep_time, arr_time, air_time, distance)} \hlopt{%>%}
+  \hlkwd{mutate_each}\hlstd{(}\hlkwd{funs}\hlstd{(scale),} \hlopt{-}\hlstd{arr_delay)} \hlopt{%>%}
+  \hlkwd{mutate_each}\hlstd{(}\hlkwd{funs}\hlstd{(as.numeric),} \hlopt{-}\hlstd{arr_delay)} \hlopt{%>%}
+  \hlkwd{mutate}\hlstd{(}\hlkwc{arr_delay} \hlstd{= scales}\hlopt{::}\hlkwd{rescale}\hlstd{(arr_delay,} \hlkwc{to} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{1}\hlstd{)))} \hlopt{%>%}
+  \hlstd{data.frame}
+
+\hlcom{# create two models}
+\hlcom{# random starting weights are different for each}
+\hlstd{ex1} \hlkwb{<-} \hlkwd{nnet}\hlstd{(arr_delay} \hlopt{~} \hlstd{.,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{linout} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{data} \hlstd{= tomod)}
+\hlstd{ex2} \hlkwb{<-} \hlkwd{nnet}\hlstd{(arr_delay} \hlopt{~} \hlstd{.,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{linout} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{data} \hlstd{= tomod)}
+
+\hlcom{# view olden results }
+\hlkwd{olden}\hlstd{(ex1)}
+\hlkwd{olden}\hlstd{(ex2)}
+\end{alltt}
+\end{kframe}
 

### 2015-10-11 nnt_manu.Rnw
finished applied example

@@
 # set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
+opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T)
 options(replace.assign = TRUE, digits = 1)
@@
 A sample dataset is also included with \pkg{NeuralNetTools}.
-The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.
+The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.
 The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.
@@
 Although \pkg{NeuralnetTools} provides several methods to extract information from a neural network, it does not provide explicit guidance for developing the initial model.
-A more challenging aspect is understanding the effects of network architecture, use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting during model development.
-A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to introduce a `best practices' approach to neural network development.
+A potentially more challenging aspect of using neural netwworks is understanding the effects of network architecture, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting during model development.
+A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to introduce a `best practices' approach to model development.
 The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive application of neural networks in practice.
@@
 
-<<eval = F>>=
-library(nycflights13)
+A dataset from the \pkg{nycflights13} package \citep{Wickham14b} is used to demonstrate the effects of different training methods on model conclusions.
+This dataset describes flight data for all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
+The example focuses on all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}) at the destination airport.
+Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between desinations (\code{air\_time}), and distance flown (\code{distance}).
+All explanatory variables are normalized and the response variable is standardized to 0--1.
+
+The effects of network architecture and initial starting weights on indications of variable importance were evaluated by creating a large number of models and assessing the uncertainty of the importance measures.
+Models with one, five, or ten hidden nodes and 100 separate models for each node level with a different set of random starting weights were evaluated.
+The relative importance of the variables was saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{code}{}).
+
+Several conclusions about the dataset and effects of network architecture can be made from the results in \cref{fig:flightimp}.
+First, delays in arrival time are consistently negatively related to distance travelled and positively related to departure delays and time in the air.
+That is, flights arrived later than their scheduled destination time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
+Second, the indications of importance varied in the range of magnitude between the models (i.e., one node  varied +/- 1 and the others +/- 3).
+This result suggests that the absolute values of importance only have relevance within a model, whereas only the relative rankings (e.g., least, most important) can be compared between models.
+Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture (i.e., effect of random starting weights).
+This conclusion emphasizes that a single model can often provide misleading information about variable importance such that several fits are often required before results are stabilized.
+Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect conclusions of variable importance and should be equally considered during model development.
+<<echo = F>>=
+# library(nnet)
+# library(NeuralNetTools)
+# library(dplyr)
+# library(tidyr)
+# library(ggplot2)
+# library(nycflights13)
+# 
+# # data preprocessing
+# # select UA carrier for month of december and relevant variables
+# # normalize and standardize variables
+# tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
+#   select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
+#   mutate_each(funs(scale), -arr_delay) %>% 
+#   mutate_each(funs(as.numeric), -arr_delay) %>% 
+#   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
+#   data.frame
+# 
+# # setup initial conditions
+# nodes <- c(1, 5, 10)
+# seeds <- sample(c(1:5000), 100)
+# grids <- expand.grid(nodes, seeds)
+# names(grids) <- c('nodes', 'seeds') 
+# 
+# # iterate through conditions
+# imp_vals <- vector('list', nrow(grids))
+# for(i in 1:nrow(grids)){
+#   
+#   cat(i, '\t')
+#   
+#   # get specific conditions
+#   seed_val <- grids[i, 'seeds']
+#   set.seed(seed_val)
+#   node_val <- grids[i, 'nodes']
+# 
+#   # create model
+#   mod <- nnet(arr_delay ~ ., size = node_val, linout = TRUE, data = tomod, trace = F)  
+# 
+#   # get importance, append to output
+#   imp <- olden(mod, bar_plot = FALSE)
+#   imp_vals[[i]] <- imp
+#   
+# }
+# 
+# # summarize
+# imp_sums <- do.call('cbind', imp_vals)
+# nms <- row.names(imp_sums)
+# imp_sums <- t(imp_sums)
+# names(imp_sums) <- nms
+# row.names(imp_sums) <- 1:nrow(imp_sums)
+# imp_sums <- data.frame(grids, imp_sums) %>% 
+#   gather('variable', 'importance', -nodes, -seeds) %>% 
+#   group_by(nodes, variable) %>% 
+#   do((function(x) {
+#        
+#     out <- quantile(x$importance, c(0.05, 0.5, 0.95))
+#     out <- data.frame(t(out))
+#     names(out) <- c('lo', 'med', 'hi')
+# 
+#     out
+#       
+#   })(.)) %>% 
+#   data.frame
+# 
+# save(imp_sums, file = 'imp_sums.RData')
+
 library(dplyr)
+library(ggplot2)
 
-# data preprocessing
-# select UA carrior for month of december and relevant variables
-# normalize and standardize variables
-tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
-  select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
-  mutate_each(funs(scale), -arr_delay) %>% 
-  mutate_each(funs(as.numeric), -arr_delay) %>% 
-  mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
-  data.frame
+load(file = 'imp_sums.RData')
 
-# create two models
-# random starting weights are different for each
-ex1 <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod)  
-ex2 <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod)
+# subet by nodes so barplots can be ordered by medians
+toplo1 <- filter(imp_sums, nodes == 1) %>% 
+  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+toplo2 <- filter(imp_sums, nodes == 5) %>% 
+  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+toplo3 <- filter(imp_sums, nodes == 10) %>% 
+  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
 
-# view olden results 
-olden(ex1)
-olden(ex2)
+# widths
+wd <- 0.3
+ylims <- c(-3.93, 3.49)
+
+p1 <- ggplot(toplo1, aes(x = variable, y = med, fill = med)) + 
+  geom_bar(stat = 'identity') + 
+  geom_errorbar(aes(ymin = lo, ymax = hi), width = wd) + 
+  theme_bw() +
+  theme(axis.title.x = element_blank(), legend.position = 'none') +
+  scale_y_continuous('Importance', limits = ylims)
+
+p2 <- ggplot(toplo2, aes(x = variable, y = med, fill = med)) + 
+  geom_bar(stat = 'identity') + 
+  geom_errorbar(aes(ymin = lo, ymax = hi), width = wd) + 
+  theme_bw() +
+  theme(axis.title.x = element_blank(), legend.position = 'none') +
+  scale_y_continuous('Importance', limits = ylims)
+
+p3 <- ggplot(toplo3, aes(x = variable, y = med, fill = med)) + 
+  geom_bar(stat = 'identity') + 
+  geom_errorbar(aes(ymin = lo, ymax = hi), width = wd) + 
+  theme_bw() +
+  theme(axis.title.x = element_blank(), legend.position = 'none') +
+  scale_y_continuous('Importance', limits = ylims)
+
 @
+
+\begin{figure}
+\centering
+\subfloat[][Networks with one node]{
+<<flightimp-1, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
+print(p1)
+@
+\label{fig:flightimp1}
+}
+
+\subfloat[][Networks with five nodes]{
+<<flightimp-2, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
+print(p2)
+@
+\label{fig:flightimp2}
+}
+
+\subfloat[][Networks with ten nodes]{
+<<flightimp-3, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
+print(p3)
+@
+\label{fig:flightimp3}
+}
+\caption{Uncertainty in variable importance estimates for three neural networks to evaluate factors related to arrival delays for flights departing New York City.
+Three model types with one, five, and ten nodes were evaluated with 100 models with different starting weights for each type.}
+\label{fig:flightimp}
+\end{figure}
 
@@
 
-I would like to thank Dr.
-Bruce Wilson and Dr.
-Sanford Weisberg of the University of Minnesota for general guidance during the development of this package.
+I would like to thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.
 Contributions and suggestions from online users have also greatly improved the utility of the package.

### 2015-10-11 nnt_manu.tex
finished applied example

@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 93\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
@@
 
-\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(nycflights13)}
-\hlkwd{library}\hlstd{(dplyr)}
+A dataset from the \pkg{nycflights13} package \citep{Wickham14b} is used to demonstrate the effects of different training methods on model conclusions.
+This dataset describes flight data for all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
+The example focuses on all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}) at the destination airport.
+Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between desinations (\code{air\_time}), and distance flown (\code{distance}).
+All explanatory variables are normalized and the response variable is standardized to 0--1.
 
-\hlcom{# data preprocessing}
-\hlcom{# select UA carrior for month of december and relevant variables}
-\hlcom{# normalize and standardize variables}
-\hlstd{tomod} \hlkwb{<-} \hlkwd{filter}\hlstd{(flights, month} \hlopt{==} \hlnum{12} \hlopt{&} \hlstd{carrier} \hlopt{==} \hlstr{'UA'}\hlstd{)} \hlopt{%>%}
-  \hlkwd{select}\hlstd{(arr_delay, dep_delay, dep_time, arr_time, air_time, distance)} \hlopt{%>%}
-  \hlkwd{mutate_each}\hlstd{(}\hlkwd{funs}\hlstd{(scale),} \hlopt{-}\hlstd{arr_delay)} \hlopt{%>%}
-  \hlkwd{mutate_each}\hlstd{(}\hlkwd{funs}\hlstd{(as.numeric),} \hlopt{-}\hlstd{arr_delay)} \hlopt{%>%}
-  \hlkwd{mutate}\hlstd{(}\hlkwc{arr_delay} \hlstd{= scales}\hlopt{::}\hlkwd{rescale}\hlstd{(arr_delay,} \hlkwc{to} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{1}\hlstd{)))} \hlopt{%>%}
-  \hlstd{data.frame}
+The effects of network architecture and initial starting weights on indications of variable importance were evaluated by creating a large number of models and assessing the uncertainty of the importance measures.
+Models with one, five, or ten hidden nodes and 100 separate models for each node level with a different set of random starting weights were evaluated.
+The relative importance of the variables was saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{code}{}).
 
-\hlcom{# create two models}
-\hlcom{# random starting weights are different for each}
-\hlstd{ex1} \hlkwb{<-} \hlkwd{nnet}\hlstd{(arr_delay} \hlopt{~} \hlstd{.,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{linout} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{data} \hlstd{= tomod)}
-\hlstd{ex2} \hlkwb{<-} \hlkwd{nnet}\hlstd{(arr_delay} \hlopt{~} \hlstd{.,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{linout} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{data} \hlstd{= tomod)}
+Several conclusions about the dataset and effects of network architecture can be made from the results in \cref{fig:flightimp}.
+First, delays in arrival time are consistently negatively related to distance travelled and positively related to departure delays and time in the air.
+That is, flights arrived later than their scheduled destination time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
+Second, the indications of importance varied in the range of magnitude between the models (i.e., one node  varied +/- 1 and the others +/- 3).
+This result suggests that the absolute values of importance only have relevance within a model, whereas only the relative rankings (e.g., least, most important) can be compared between models.
+Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture (i.e., effect of random starting weights).
+This conclusion emphasizes that a single model can often provide misleading information about variable importance such that several fits are often required before results are stabilized.
+Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect conclusions of variable importance and should be equally considered during model development.
 
-\hlcom{# view olden results }
-\hlkwd{olden}\hlstd{(ex1)}
-\hlkwd{olden}\hlstd{(ex2)}
-\end{alltt}
-\end{kframe}
+\begin{figure}
+\centering
+\subfloat[][Networks with one node]{
+
+{\centering \includegraphics[width=0.95\textwidth]{figs/flightimp-1-1}
+
+}
+
+\label{fig:flightimp1}
+}
+
+\subfloat[][Networks with five nodes]{
+
+{\centering \includegraphics[width=0.95\textwidth]{figs/flightimp-2-1}
+
+}
+
+\label{fig:flightimp2}
+}
+
+\subfloat[][Networks with ten nodes]{
+
+{\centering \includegraphics[width=0.95\textwidth]{figs/flightimp-3-1}
+
+}
+
+\label{fig:flightimp3}
+}
+\caption{Uncertainty in variable importance estimates for three neural networks to evaluate factors related to arrival delays for flights departing New York City.
+Three model types with one, five, and ten nodes were evaluated with 100 models with different starting weights for each type.}
+\label{fig:flightimp}
+\end{figure}
 

### 2015-10-11 flightimp.R

@@ -1,0 +1,95 @@
+library(nnet)
+library(NeuralNetTools)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(nycflights13)
+
+# data preprocessing
+# select UA carrier for month of december and relevant variables
+# normalize and standardize variables
+tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
+  select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
+  mutate_each(funs(scale), -arr_delay) %>% 
+  mutate_each(funs(as.numeric), -arr_delay) %>% 
+  mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
+  data.frame
+
+# setup initial conditions
+nodes <- c(1, 5, 10)
+seeds <- sample(c(1:5000), 100)
+grids <- expand.grid(nodes, seeds)
+names(grids) <- c('nodes', 'seeds') 
+
+# iterate through conditions
+imp_vals <- vector('list', nrow(grids))
+for(i in 1:nrow(grids)){
+  
+  cat(i, '\t')
+  
+  # get specific conditions
+  seed_val <- grids[i, 'seeds']
+  set.seed(seed_val)
+  node_val <- grids[i, 'nodes']
+
+  # create model
+  mod <- nnet(arr_delay ~ ., size = node_val, linout = TRUE, data = tomod, trace = F)  
+
+  # get importance, append to output
+  imp <- olden(mod, bar_plot = FALSE)
+  imp_vals[[i]] <- imp
+  
+}
+
+# summarize
+imp_sums <- do.call('cbind', imp_vals)
+nms <- row.names(imp_sums)
+imp_sums <- t(imp_sums)
+names(imp_sums) <- nms
+row.names(imp_sums) <- 1:nrow(imp_sums)
+imp_sums <- data.frame(grids, imp_sums) %>% 
+  gather('variable', 'importance', -nodes, -seeds) %>% 
+  group_by(nodes, variable) %>% 
+  do((function(x) {
+       
+    out <- quantile(x$importance, c(0.05, 0.5, 0.95))
+    out <- data.frame(t(out))
+    names(out) <- c('lo', 'med', 'hi')
+
+    out
+      
+  })(.)) %>% 
+  data.frame
+
+# subet by nodes so barplots can be ordered by medians
+toplo1 <- filter(imp_sums, nodes == 1) %>% 
[…]

### 2015-10-11 nnt_manu.Rnw
link to flight imp code

@@
 Models with one, five, or ten hidden nodes and 100 separate models for each node level with a different set of random starting weights were evaluated.
-The relative importance of the variables was saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{code}{}).
+The relative importance of the variables was saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
 

### 2015-10-13 nnt_manu.Rnw
abstract and code chunk formats

@@
 \usepackage{subfig}
+\usepackage{framed}
 
@@
 \Abstract{
-Functions within this package can be used for the interpretation of neural network models created in \proglang{R}, including functions to plot a neural network interpretation diagram, evaluation of variable importance, and a sensitivity analysis of input variables.
+The supervised neural network has been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.
+A common criticism of the method is the inability to characterize relationships among variables from a fitted model.
+Several methods proposed in recent years to `illuminate the black-box' have not been made available in an open-source programming environment.
+This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.
+Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.
+Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.
+Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.
 }
@@
   Gulf Breeze, Florida, 32561, USA\\
+  Telephone: +1/850/934-2480\\
   E-mail: \email{beck.marcus@epa.gov}
 }
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/512/507-7103
-%% Fax: +43/512/507-2851
 
@@
 # set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T)
-options(replace.assign = TRUE, digits = 1)
+opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T, background = 'white', highlight = F, prompt = TRUE)
+options(replace.assign = TRUE, digits = 1, prompt="R> ")
 
@@
 
-Data science is a relatively new paradigm of analysis that focuses on the synthesis of unstructured information from multiple sources to identify patterns or trends `born from the data' \citep{Kelling09}.
+Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Kelling09}.
 A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
 Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data.
-Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}).
-Morever, disciplines that have historically been limited by data quantity, such as ecological studies across broad temporal and spatial scales, have also realized the importance of data intensive approaches given the increasing use of novel techniques to acquire information (e.g., \citet{Swanson15}).
-Regardless of the discipline, quantitative methods that explicitly focus on inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
+Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}), whereas disciplines that have historically been limited by data quantity have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information (e.g., \citet{Swanson15}).
+Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
 
@@
 Functions provided by the package are those previously described in \citep{Olden02} but have not been available in an open-source programming environment.
-The reach of the package is all-inclusive such that generic functions were developed using S3 methods for all neural network object classes available in \proglang{R}.
-The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) briefly describe similarities and differences between existing neural network packages in \proglang{R}, and 3) describe the theory and application of the primary functions in the \pkg{NeuralNetTools} package.
+The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network functions available in \proglang{R}.
+The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide a context for appropriate use of neural networks and \pkg{NeuralNetTools} in data exploration using an applied example.
 The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.
@@
 The process is repeated until the chosen error function is minimized.
-Formulaically, a generic \ac{mlp} neural network can be represented as \citep{Ripley96}:
+Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
 \begin{equation}
@@
 
-The stable release of \pkg{NeurelNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
+The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
 
-<<eval = F>>=
+<<eval = F, echo = T>>=
 # install from CRAN
@@
 A sample dataset is also included with \pkg{NeuralNetTools}.
-The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.
+The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).
 The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.
@@
 Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods will be made available based on future popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
-A primary objective of the package is to dispel the myth that supervised neural networks for prediction are `black boxes' that provide no information about underlying relationships between the variables \citep{Paruelo97,Olden02}.
+A primary objective of the package is to dispel the myth that supervised neural networks are `black boxes' that provide no information about underlying relationships between the variables \citep{Paruelo97,Olden02}.
 Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation has many parallels with traditional modelling techniques that can allow insight into causation.
@@
 
-I would like to thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.
+I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.
 Contributions and suggestions from online users have also greatly improved the utility of the package.
+Funding for this project was supported in part by an Interdisciplinary Doctoral Fellowship provided by the Graduate School at the University of Minnesota to M.
+Beck.
 

### 2015-10-13 nnt_manu.tex
abstract and code chunk formats

@@
 
-The stable release of \pkg{NeurelNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
+The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
 
-\begin{kframe}
-\begin{alltt}
-\hlcom{# install from CRAN}
-\hlkwd{install.packages}\hlstd{(}\hlstr{'NeuralNetTools'}\hlstd{)}
-\hlkwd{library}\hlstd{(NeuralNetTools)}
-
-\hlcom{# install from GitHub}
-\hlkwd{install.package}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
-\hlkwd{library}\hlstd{(devtools)}
-\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/NeuralNetTools'}\hlstd{)}
-\hlkwd{library}\hlstd{(NeuralNetTools)}
-\end{alltt}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> # install from CRAN
+R> install.packages('NeuralNetTools')
+R> library(NeuralNetTools)
+R> 
+R> # install from GitHub
+R> install.package('devtools')
+R> library(devtools)
+R> install_github('fawda123/NeuralNetTools')
+R> library(NeuralNetTools)
+\end{verbatim}
 \end{kframe}
+\end{knitrout}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# set random seed for initial weights for training}
-\hlkwd{set.seed}\hlstd{(}\hlnum{123}\hlstd{)}
-
-\hlcom{# mlp object, RSNNS package}
-\hlkwd{library}\hlstd{(RSNNS)}
-\hlstd{x} \hlkwb{<-} \hlstd{neuraldat[,} \hlkwd{c}\hlstd{(}\hlstr{'X1'}\hlstd{,} \hlstr{'X2'}\hlstd{,} \hlstr{'X3'}\hlstd{)]}
-\hlstd{y} \hlkwb{<-} \hlstd{neuraldat[,} \hlstr{'Y1'}\hlstd{]}
-\hlstd{mod1} \hlkwb{<-} \hlkwd{mlp}\hlstd{(x, y,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{)}
-
-\hlcom{# nn object, neuralnet package}
-\hlkwd{library}\hlstd{(neuralnet)}
-\hlstd{mod2} \hlkwb{<-} \hlkwd{neuralnet}\hlstd{(Y1} \hlopt{~} \hlstd{X1} \hlopt{+} \hlstd{X2} \hlopt{+} \hlstd{X3,} \hlkwc{data} \hlstd{= neuraldat,} \hlkwc{hidden} \hlstd{=} \hlnum{5}\hlstd{)}
-
-\hlcom{#nnet object, nnet package}
-\hlkwd{library}\hlstd{(nnet)}
-\hlstd{mod3} \hlkwb{<-} \hlkwd{nnet}\hlstd{(Y1} \hlopt{~} \hlstd{X1} \hlopt{+} \hlstd{X2} \hlopt{+} \hlstd{X3,} \hlkwc{data} \hlstd{= neuraldat,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{)}
-\end{alltt}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> # set random seed for initial weights for training
+R> set.seed(123)
+R> 
+R> # mlp object, RSNNS package
+R> library(RSNNS)
+R> x <- neuraldat[, c('X1', 'X2', 'X3')]
+R> y <- neuraldat[, 'Y1']
+R> mod1 <- mlp(x, y, size = 5)
+R> 
+R> # nn object, neuralnet package
+R> library(neuralnet)
+R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
+R> 
+R> #nnet object, nnet package
+R> library(nnet)
+R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
+\end{verbatim}
 \end{kframe}
@@
 
-\begin{kframe}
-\begin{alltt}
-\hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(mod3,} \hlkwc{nid} \hlstd{= F,} \hlkwc{circle_col} \hlstd{=} \hlstr{'grey'}\hlstd{,} \hlkwc{bord_col} \hlstd{=} \hlstr{'grey'}\hlstd{)}
-\hlkwd{plotnet}\hlstd{(mod3)}
-\end{alltt}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> par(mar = c(0, 0, 0, 0))
+R> plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
+R> plotnet(mod3)
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 
+\end{knitrout}
+
 A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection.
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# create a model with a skip layer}
-\hlstd{modskip} \hlkwb{<-} \hlkwd{nnet}\hlstd{(Y1} \hlopt{~} \hlstd{X1} \hlopt{+} \hlstd{X2} \hlopt{+} \hlstd{X3,} \hlkwc{data} \hlstd{= neuraldat,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-
-\hlcom{# plot}
-\hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(modskip,} \hlkwc{skip} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlkwd{plotnet}\hlstd{(modskip)}
-\end{alltt}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> # create a model with a skip layer
+R> modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
+R> 
+R> # plot
+R> par(mar = c(0, 0, 0, 0))
+R> plotnet(modskip, skip = TRUE)
+R> plotnet(modskip)
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 
-\begin{kframe}
-\begin{alltt}
-\hlcom{# pruned model using RSNNS}
-\hlstd{pruneFuncParams} \hlkwb{<-} \hlkwd{list}\hlstd{(}\hlkwc{max_pr_error_increase} \hlstd{=} \hlnum{10.0}\hlstd{,} \hlkwc{pr_accepted_error} \hlstd{=} \hlnum{1.0}\hlstd{,}
-  \hlkwc{no_of_pr_retrain_cycles} \hlstd{=} \hlnum{1000}\hlstd{,} \hlkwc{min_error_to_stop} \hlstd{=} \hlnum{0.01}\hlstd{,}
-  \hlkwc{init_matrix_value} \hlstd{=} \hlnum{1e-6}\hlstd{,} \hlkwc{input_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{hidden_pruning} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlstd{mod} \hlkwb{<-} \hlkwd{mlp}\hlstd{(x, y,} \hlkwc{size} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{pruneFunc} \hlstd{=} \hlstr{"OptimalBrainSurgeon"}\hlstd{,}
- \hlkwc{pruneFuncParams} \hlstd{= pruneFuncParams)}
-
-\hlcom{# plot, pruned connections are omitted (default) or included with prune_col}
-\hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(mod,} \hlkwc{rel_rsc} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{8}\hlstd{))}
-\hlkwd{plotnet}\hlstd{(mod,} \hlkwc{prune_col} \hlstd{=} \hlstr{'lightblue'}\hlstd{,} \hlkwc{rel_rsc} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{8}\hlstd{))}
-\end{alltt}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> # pruned model using RSNNS
+R> pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
++   no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
++   init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
+R> mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
++  pruneFuncParams = pruneFuncParams)
+R> 
+R> # plot, pruned connections are omitted (default) or included with prune_col
+R> par(mar = c(0, 0, 0, 0))
+R> plotnet(mod, rel_rsc = c(3, 8))
+R> plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \end{figure}
+
+\end{knitrout}
 
@@
 
-\begin{kframe}
-\begin{alltt}
-\hlcom{# garson and olden functions for each model}
-\hlkwd{garson}\hlstd{(mod1)}
-\hlkwd{olden}\hlstd{(mod1)}
-\hlkwd{garson}\hlstd{(mod2)}
-\hlkwd{olden}\hlstd{(mod2)}
-\hlkwd{garson}\hlstd{(mod3)}
-\hlkwd{olden}\hlstd{(mod3)}
-\end{alltt}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> # garson and olden functions for each model
+R> garson(mod1)
+R> olden(mod1)
+R> garson(mod2)
+R> olden(mod2)
+R> garson(mod3)
+R> olden(mod3)
+\end{verbatim}
 \end{kframe}
+\end{knitrout}
 
@@
 \subfloat[][Model 1 (\code{mlp}), \code{garson}]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotimp1}
@@
 \subfloat[][Model 1 (\code{mlp}), \code{olden}]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotimp2}
@@
 \subfloat[][Model 2 (\code{nn}), \code{garson}]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotimp3}
@@
 \subfloat[][Model 2 (\code{nn}), \code{olden}]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotimp4}
@@
 \subfloat[][Model 3 (\code{nnet}), \code{garson}]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotimp5}
@@
 \subfloat[][Model 3 (\code{nnet}), \code{olden}]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotimp6}
@@
 For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights in each bar proportional to the constant values.
-\begin{kframe}
-\begin{alltt}
-\hlcom{# lek profile, quantile evaluation}
-\hlkwd{lekprofile}\hlstd{(mod3)}
-
-\hlcom{# lek profile, quantile groups }
-\hlkwd{lekprofile}\hlstd{(mod3,} \hlkwc{group_show} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-
-\hlcom{# lek profile, cluster evaluation}
-\hlkwd{lekprofile}\hlstd{(mod3,} \hlkwc{group_vals} \hlstd{=} \hlnum{6}\hlstd{)}
-
-\hlcom{# lek profile, cluster groups}
-\hlkwd{lekprofile}\hlstd{(mod3,} \hlkwc{group_vals} \hlstd{=} \hlnum{6}\hlstd{,} \hlkwc{group_show} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\end{alltt}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> # lek profile, quantile evaluation
+R> lekprofile(mod3)
+R> 
+R> # lek profile, quantile groups 
+R> lekprofile(mod3, group_show = TRUE)
+R> 
+R> # lek profile, cluster evaluation
+R> lekprofile(mod3, group_vals = 6)
+R> 
+R> # lek profile, cluster groups
+R> lekprofile(mod3, group_vals = 6, group_show = TRUE)
+\end{verbatim}
 \end{kframe}
+\end{knitrout}
 
@@
 \subfloat[][Response with quantiles]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotlek_sens1}
@@
 \subfloat[][Response with clusters]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotlek_sens2}
@@
 \subfloat[][Quantile groupings]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotlek_bars1}
@@
 \subfloat[][Cluster groupings]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:plotlek_bars2}
@@
 \subfloat[][Networks with one node]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:flightimp1}
@@
 \subfloat[][Networks with five nodes]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:flightimp2}
@@
 \subfloat[][Networks with ten nodes]{
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
@@
 
+\end{knitrout}
 \label{fig:flightimp3}

### 2015-10-15 nnt_manu.Rnw
edits to draft before friendly review

@@
 \Abstract{
-The supervised neural network has been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.
+Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.
 A common criticism of the method is the inability to characterize relationships among variables from a fitted model.
-Several methods proposed in recent years to `illuminate the black-box' have not been made available in an open-source programming environment.
+Although several methods proposed in recent years to `illuminate the black-box', they have not been made available in an open-source programming environment.
 This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.
 Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.
-Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
 The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.
@@
 
-Statistical methods that have been used to support data exploration for inductive analysis are numerous \citep{Jain00}.
-A common theme among these methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns in the data with minimal human intervention.
+Statistical methods that have been used to support data exploration are numerous \citep{Jain00}.
+A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.
 Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.
-Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in data intensive science.
+Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in emerging fields focused on data exploration.
 The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
 This model is typically used to predict the response of one or more variables given one to many explanatory variables.
-The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through an iterative training process with the backpropation algorithm.
-Conceptually, the \ac{mlp} is nothing more than a hyper-parameterized non-linear model that can fit a smooth function to any dataset with almost non-existent residual error \citep{Hornik91}.
+The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropation algorithm.
+Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.
 
-An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of critical model information.
-Information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
+An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of model information.
+Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
 As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
-\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks, most of which were previously available but not commonly used.
+\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks.
+Many of these methods were previously described but not commonly used.
 For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
 These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.
-In essence, \ac{mlp} neural networks were developed for prediction but methods described in \citep{Olden02} leverage these models to describe data signals.
-Increasing the accesibility of these diagnostic tools will have value for exploratory analysis in data science.
+Although \ac{mlp} neural networks were developed for prediction, methods described in \citet{Olden02} leverage these models to describe data signals.
+Increasing the accesibility of these diagnostic tools will have value for exploratory data analysis and may also inform causal inference.
 
 This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.
-Functions provided by the package are those previously described in \citep{Olden02} but have not been available in an open-source programming environment.
-The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network functions available in \proglang{R}.
-The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide a context for appropriate use of neural networks and \pkg{NeuralNetTools} in data exploration using an applied example.
+Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.
+The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.
+The objectives are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.
 The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.
@@
 
-An intriguing characteristic of neural networks is the relaxation of assumptions for the distributional characteristics of the response variables.
-Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.
 The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.
-Information travels in one direction where a set of values for one to many variables in the input layer propagates through one or more hidden layers to the resulting layer of the response variables.
+Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables.
 `Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.
@@
 The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.
-Training is commonly achieved using the backpropagation algorithm described in detail in \citep{Rumelhart86}.
-This algorithm identifies the optimal weighting scheme between layers through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.
+Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.
+This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.
 The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.
 The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.
-The process is repeated until the chosen error function is minimized.
+The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.
 Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
@@
 \end{equation}
-where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the respective weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node as the information progresses through the network.
+where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node.
 
@@
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
-Methods are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class) objects.
+Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.
 Additional \code{default} or \code{numeric} methods are available for some of the generic functions.
@@
 
-\pkg{NeuralNetTools} includes four main functions that were developed following techniques described in \citet{Olden02} and references therein.
-These functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.
-Most of the functions require the extraction of model weights in a common format for each of the neural network object classes in \proglang{R}.
+\pkg{NeuralNetTools} includes four main functions that were developed following similar techniques in \citet{Olden02} and references therein.
+Functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.
+Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.
 The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.
-A two-element \code{list} is returned with the first element describing the structure of the network and the second element as a named list of weight values for the input model.
-The function is used internally within the main package functions but may be useful for comparing networks of different object classes.
+A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.
+The function is used internally within the main functions but may be useful for comparing networks of different classes.
 
-A sample dataset is also included with \pkg{NeuralNetTools}.
-The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).
+A common approach for data preprocessing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
+A sample dataset that follows this format is included with \pkg{NeuralNetTools}.
+The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and five columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).
 The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.
 The response variables are also standardized from zero to one.
-A common approach for data preprocessing prior to creating a neural network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
-Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
-The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects (note the syntax differences).
+Additional datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
+The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.
 
@@
 
-The number of existing functions in \pkg{R} to view neural networks is minimal.
+Existing plot functions in \proglang{R} to view neural networks is minimal.
 Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
-To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.
+To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in \proglang{R}.
 Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.
 
-Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a network as a \acl{nid} \citep{Ozesmi99} and includes several options to customize plot aesthetics.
-A \ac{nid} is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to view the thickness and color of the weight connections based on mangnitude and sign, respectively.
-The default settings plot positive weights between layers as black lines and negative weights as grey lines.
+Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid} \citep[\acs{nid},][]{Ozesmi99} and includes several options to customize aesthetics.
+A \ac{nid} is a modification to the standard conceptual illustration of the \ac{mlp} network that changes the thickness and color of the weight connections based on magnitude and sign, respectively.
+Positive weights between layers are shown as black lines and negative weights as grey lines.
 Line thickness is in proportion to absolute magnitude of each weight (\cref{fig:plotnet}).
@@
 <<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (\\ref{fig:plotnet1}) and using the \\acl{nid} (\\ref{fig:plotnet2}).
-Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
-Default options are changed for the example.', fig.subcap = c('', '')>>=
+Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).', fig.subcap = c('', '')>>=
 par(mar = c(0, 0, 0, 0))
@@
 
-A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection.
-The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}.
+A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection \citep{Ripley96}.
+Models with skip layers include additional connections from the input to output layers that bypass the hidden layer.
+The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}.
 If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights.
-Viewing both plots is recommended to see which network has larger relative weights.
-Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0}) will include bias connections to the output layer, whereas these are not included in the plot of the skip layer if size is greater than zero.
+Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.
 
 <<plotnet_skip, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
-The skip layer (\\ref{fig:plotnet_skip1}) and primary neural network (\\ref{fig:plotnet_skip2}) can be viewed separately with \\code{plotnet}.', results = 'hide', fig.subcap = c('', '')>>=
+The skip layer (\\ref{fig:plotnet_skip1}) and primary neural network (\\ref{fig:plotnet_skip2}) can be viewed separately with \\code{plotnet} by using \\code{skip = TRUE} or \\code{skip = FALSE}.', results = 'hide', fig.subcap = c('', '')>>=
 # create a model with a skip layer
@@
 
-The \pkg{RSNNS} package provides several algorithms that can be used to prune connections or nodes in a neural network.
+The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.
 This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
-Pruning has the benefit of a potential increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
+Pruning has additional benefits of an increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
 Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
-The \code{plotnet} function can be used to plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
+The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
 
-<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.
+<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.
 The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
@@
 
-The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in weighted connections between the layers.
-Although useful as a general tool, the \ac{nid} is impractical for evaluating most networks given the amount of weighted connections.
+The primary benefits of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in connections between the layers.
+Although useful as a general tool, the \ac{nid} can be difficult to interpret given the amount of weighted connections in most networks.
 Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.
@@
 Garson's algorithm was originally described by \citet{Garson91} and further modified by \citet{Goh95}.
-The \code{garson} function is an implementation of the method described in the appendix of \citet{Goh95}.
-This method identifies the relative importance of each variable as an absolute magnitude by deconstructing all weighted connections between the layers in the network.
+The \code{garson} function is an implementation of the method described in the appendix of \citet{Goh95} that identifies the relative importance of each variable as an absolute magnitude.
 For each input node, all weights connecting an input through the hidden layer to the response variable are identified to return a list of all weights specific to each input variable.
@@
 
-The \code{olden} function is a more flexible approach to evaluate variable importance using the connection weights algorithm described in \citet{Olden04}.
-This method calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output node and sums the product across all hidden nodes.
-An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign, as compared to Garson's algorithm which only considers absolute magnitude.
-For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect whereas Garson's algorithm may provide misleading results based on the absolute magnitude.
-An additional advantage is that Olden's algorithm is capable of evaluating neural networks with multiple hidden layers and response variables.
+The \code{olden} function is a more flexible approach to evaluate variable importance using the connection weights algorithm \citep{Olden04}.
+This method calculates importance as the summed product of the raw input-hidden and hidden-output connection weights between each input and output node.
+An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign.
+For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect, whereas Garson's algorithm may provide different results based on the absolute magnitude.
+An additional advantage is that Olden's algorithm can evaluate neural networks with multiple hidden layers and response variables.
 The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.
@@
 The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).
-The plot aesthetics are based on internal code from the function but can be changed using conventional syntax for \pkg{ggplot2}.
+The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.
 The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.
@@
 }
-\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \citealt{Olden04}).
-Garson's algorithm shows importance as absolute values from 0--1 whereas Olden's algorithm preserves sign and magnitude.
-Importance values for Olden's algorithm are directly from the summed product of model weights and should only be evaluated based on the sign and relative magnitudes between variables.}
+\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \cref{fig:plotimp1,fig:plotimp3,fig:plotimp5}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \cref{fig:plotimp2,fig:plotimp4,fig:plotimp6}, \citealt{Olden04}).
+Garson's algorithm shows importance as absolute values from 0--1, whereas Olden's algorithm preserves sign and magnitude.
+Importance values for Olden's algorithm are from the summed product of model weights and are not rescaled.}
 \label{fig:plotimp}
@@
 An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
-The profile method is fairly generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method.
+The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
+The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method.
 However, it is one of few methods used to evaluate sensitivity in neural networks.
-The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
 
-The \code{lekprofile} function evaluates the effect of explanatory variables by returning a plot of the predicted response across the range of values for each variable.
-A critical aspect of the method is the remaining explanatory variables that are held constant when separately evaluating each input variable.
-The \code{lekprofile} function provides two options for determining constant values of unevaluated explanatory variables.
-The first option following the original profile method evaluates effects of each variable while holding the remaining expalanatory variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}).
-This is implemented by creating a matrix of values for explanatory variables where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables.
-All explanatory variables are held at their median (or other constant value), while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
+The \code{lekprofile} function evaluates the effects of input variables by returning a plot of model predictions across the range of values for each variable.
+The remaining explanatory variables are held constant when evaluating the effects of each input variable.
+The \code{lekprofile} function provides two options for setting constant values of unevaluated explanatory variables.
+The first option follows the original profile method by holding unevaluated variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}).
+This is implemented by creating a matrix where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables.
+All explanatory variables are held at a chosen constant value (e.g., median), while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
 This matrix is then used to predict values of the response variable from a fitted model object.
 This is repeated for each explanatory variable to obtain all response curves.
-Constant values are defined by passing one or more numeric values in the range 0--1 to the \code{group\_vals} that define the quantiles, with the default as minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
+Constant values are set in \code{lekprofile} by passing one or more values in the range 0--1 to the \code{group\_vals} argument.
+The default holds variables at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles (i.e., \code{group\_vals = c(0, 0.2, 0.4, 0.6, 0.8, 1)}).
 
-An alternative implementation of \code{lekprofile} is to group the unevaluated explanatory variables using natural groupings defined by statistical properties of the data.
-Covariance among predictors may present unlikely scenarios if all unevaluated variables at held at the same level (e.g., high values for one variable may be unlikely to co-occur with high values for a second variable).
-To address this issue, a second option is available to hold unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
+A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data.
+Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable).
+The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
 Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one.
@@
 \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.
-Lake clusters were identified given covariance among explanatory variables, such that holding explanatory variables at related values created more ecologically relevant response curves.
-Similar plots can be created with \code{lekprofile} as response curves (\cref{fig:plotlek_sens}) or to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
-For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights in each bar proportional to the constant values.
+Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.
+Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
+For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
 <<eval = F>>=
@@
 }s
-\caption{Sensitivity analysis ofa  neural network using the Lek profile method to evaluate the effects of explanatory variables.
-\Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles and \cref{fig:plotlek_sens2} groups by cluster means from {\it k}-means clustering of all explanatory variables.
-Values at which explanatory variables are held constant in each plot are shown in \cref{fig:plotlek_bars1,fig:plotlek_bars2}, respectively.}
+\caption{Sensitivity analysis of a neural network using the Lek profile method to evaluate the effects of explanatory variables.
+\Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles (minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles) and \cref{fig:plotlek_sens2} groups by cluster means (six groups).
+Values at which explanatory variables are held constant for each group are shown in \cref{fig:plotlek_bars1,fig:plotlek_bars2}.}
 \label{fig:plotlek_sens}
@@
 }
-\caption{Bar plots for values in each group of unevaluated explanatory variables in \cref{fig:plotlek_sens1,fig:plotlek_sens2}.
+\caption{Bar plots for values of unevaluated explanatory variables in each group in \cref{fig:plotlek_sens1,fig:plotlek_sens2}.
 \Cref{fig:plotlek_bars1} shows default quantile groupings set at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
-For example, variables are held at zero for group 1 (i.e., stacked bars with no height) for the minimum value, whereas group 6 holds variables at their maximum of one.
-\Cref{fig:plotlek_bars2} shows the cluster centers for each variable in each group identified from {\it k}-means clustering.}
+For example, variables are held at negative values for group 1 (i.e., stacked bars with negative heights) for the minimum value, whereas group 6 holds variables at their maximum (largest positive heights).
+\Cref{fig:plotlek_bars2} shows the cluster centers for each variable in each group.
+Groups in \cref{fig:plotlek_bars2} are random because the input variables are from a standard normal distribution.}
 \label{fig:plotlek_bars}
@@
 
-Although \pkg{NeuralnetTools} provides several methods to extract information from a neural network, it does not provide explicit guidance for developing the initial model.
-A potentially more challenging aspect of using neural netwworks is understanding the effects of network architecture, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting during model development.
-A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to introduce a `best practices' approach to model development.
-The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive application of neural networks in practice.
-Several texts provide explicit guidelines on developing neural network models \cite[e.g.,][]{Ripley96, Lek00, Maier00}.
-In general, the following should be considered during model development:
+Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.
+A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.
+A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.
+The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.
+In general, the following should be considered during initial development \cite{Ripley96, Lek00, Maier00}:
 \begin{itemize}
-\item Initial data pre-processing to normalize inputs and standardize response,
+\item Initial data pre-processing to normalize inputs, standardize response, or assess influence of outliers,
 \item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, pruning weights or inputs,
-\item Split of dataset into a training and test dataset, e.g., 2:1, 3:1, leave-one-out, etc., 
+\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
 \item Initial starting weights for the back-propagation algorithm, and
-\item Criteria for stopping model training iterations, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
+\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
 \end{itemize}
 
-A dataset from the \pkg{nycflights13} package \citep{Wickham14b} is used to demonstrate the effects of different training methods on model conclusions.
-This dataset describes flight data for all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
-The example focuses on all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}) at the destination airport.
-Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between desinations (\code{air\_time}), and distance flown (\code{distance}).
-All explanatory variables are normalized and the response variable is standardized to 0--1.
+A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions.
+This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
+The example uses all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.
+Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between desinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
+All explanatory variables were normalized and the response variable was standardized to 0--1.
 
-The effects of network architecture and initial starting weights on indications of variable importance were evaluated by creating a large number of models and assessing the uncertainty of the importance measures.
-Models with one, five, or ten hidden nodes and 100 separate models for each node level with a different set of random starting weights were evaluated.
-The relative importance of the variables was saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
+The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.
+Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.
+Each model had a random set of starting weights for the first training iteration.
+Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
 
-Several conclusions about the dataset and effects of network architecture can be made from the results in \cref{fig:flightimp}.
-First, delays in arrival time are consistently negatively related to distance travelled and positively related to departure delays and time in the air.
-That is, flights arrived later than their scheduled destination time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
-Second, the indications of importance varied in the range of magnitude between the models (i.e., one node  varied +/- 1 and the others +/- 3).
-This result suggests that the absolute values of importance only have relevance within a model, whereas only the relative rankings (e.g., least, most important) can be compared between models.
-Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture (i.e., effect of random starting weights).
-This conclusion emphasizes that a single model can often provide misleading information about variable importance such that several fits are often required before results are stabilized.
-Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect conclusions of variable importance and should be equally considered during model development.
+Several conclusions can be made from \cref{fig:flightimp}.
+First, delays in arrival time are negatively related to distance and positively related to departure delays and travel air.
+That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
+No conclusions can be made for the other variables because the bounds of uncertainty included zero.
+Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).
+This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.
+Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.
+This suggests that a single model can provide misleading information such that several models may be required to improve the certainty of conclusions.
+Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
 <<echo = F>>=
@@
 
-The \pkg{NeuralNetTools} package provides a simple approach that improves the quality of information obtained from a feed-forward \ac{mlp} neural network.
-Functions in the package can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).
-Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods will be made available based on future popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
-A primary objective of the package is to dispel the myth that supervised neural networks are `black boxes' that provide no information about underlying relationships between the variables \citep{Paruelo97,Olden02}.
-Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation has many parallels with traditional modelling techniques that can allow insight into causation.
-Moreover, the model fitting process requires the minimization of an objective function such that conventional techniques to evaluate model sensitivity or performance (e.g., cross-validation) can be used with neural networks.
-Functions in \pkg{NeuralNetTools} can facilitate the selection of the optimal network architecture or can be used for post-hoc assessment.
+The \pkg{NeuralNetTools} package provides a simple approach to improve the quality of information obtained from a feed-forward \ac{mlp} neural network.
+Functions can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).
+Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods could be added based on popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
 
-A more compelling issue is when and how to apply neural networks given alternative methods for data-intensive exploration.
-The popularity of the \ac{mlp} neural network is partially the cause of misperceptions and generalizations of their benefits as modelling tools \citep{Burke97}.
-Indeed, the `neural' component of neural networks is advertised as a mathematical representation of the network of synaptic impulses in the human brain, although the former is of course an over-simplification of the latter.
-Several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.
-As such, the neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques, with particular emphasis on the needs of a given dataset or research question.
-For example, an appropriate method should be chosen based on power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant concerns that are routinely considered during exploratory data analysis.
-The true value of the \pkg{NeuralNetTools} package is a set of analysis tools that can inform a more balanced assessment given the alternative and complementary methods for data-intensive exploration.
+A primary objective of the package is to address the myth that supervised neural networks are `black boxes' that provide no information about underlying relationships between variables \citep{Paruelo97,Olden02}.
+Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation has many parallels with traditional statistical techniques that can allow insight into causation \citep{Cheng94}.
+Moreover, the model fitting process minimizes error using a standard objective function such that conventional techniques to evaluate model sensitivity or performance (e.g., cross-validation) can be used with neural networks.
+As such, functions in \pkg{NeuralNetTools} can facilitate the selection of the optimal network architecture or can be used for post-hoc assessment.
+
+A more compelling issue is when and how to apply neural networks given alternative methods of analysis.
+The popularity of the \ac{mlp} neural network is partially the cause of generalizations and misperceptions of their benefits as modelling tools \citep{Burke97}.
+Perhaps an overstatement, the `neural' component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.
+Additionally, several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.
+A neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques, with particular emphasis on the specific needs of a dataset or research question.
+Considerations for choosing a method may include power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant considerations of exploratory data analysis \citep{Zuur10}.
+\pkg{NeuralNetTools} provides analysis tools that can inform a more balanced assessment given alternative and complementary methods for data exploration.
 

### 2015-10-15 nnt_manu.tex
edits to draft before friendly review

@@
 \caption{Examples from the \code{plotnet} function showing neural networks as a standard graphic (\ref{fig:plotnet1}) and using the \acl{nid} (\ref{fig:plotnet2}).
-Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).
-Default options are changed for the example.}\label{fig:plotnet}
+Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).}\label{fig:plotnet}
 \end{figure}
@@
 
-A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection.
-The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}.
+A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection \citep{Ripley96}.
+Models with skip layers include additional connections from the input to output layers that bypass the hidden layer.
+The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}.
 If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights.
-Viewing both plots is recommended to see which network has larger relative weights.
-Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0}) will include bias connections to the output layer, whereas these are not included in the plot of the skip layer if size is greater than zero.
+Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.
 
@@
 \caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
-The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet}.}\label{fig:plotnet_skip}
+The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet} by using \code{skip = TRUE} or \code{skip = FALSE}.}\label{fig:plotnet_skip}
 \end{figure}
@@
 
-The \pkg{RSNNS} package provides several algorithms that can be used to prune connections or nodes in a neural network.
+The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.
 This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
-Pruning has the benefit of a potential increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
+Pruning has additional benefits of an increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
 Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
-The \code{plotnet} function can be used to plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
+The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
 
@@
 
-\caption{A pruned neural network using the ``Optimal Brain Surgeon'' algorithm described in \citet{Zell98}.
+\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the ``Optimal Brain Surgeon'' algorithm described in \citet{Zell98}.
 The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}\label{fig:plotprune}
@@
 \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.
-Lake clusters were identified given covariance among explanatory variables, such that holding explanatory variables at related values created more ecologically relevant response curves.
-Similar plots can be created with \code{lekprofile} as response curves (\cref{fig:plotlek_sens}) or to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
-For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights in each bar proportional to the constant values.
+Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.
+Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
+For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
 \begin{knitrout}
@@
 
-Although \pkg{NeuralnetTools} provides several methods to extract information from a neural network, it does not provide explicit guidance for developing the initial model.
-A potentially more challenging aspect of using neural netwworks is understanding the effects of network architecture, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting during model development.
-A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to introduce a `best practices' approach to model development.
-The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive application of neural networks in practice.
-Several texts provide explicit guidelines on developing neural network models \cite[e.g.,][]{Ripley96, Lek00, Maier00}.
-In general, the following should be considered during model development:
+Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.
+A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.
+A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.
+The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.
+In general, the following should be considered during initial development \cite{Ripley96, Lek00, Maier00}:
 \begin{itemize}
-\item Initial data pre-processing to normalize inputs and standardize response,
+\item Initial data pre-processing to normalize inputs, standardize response, or assess influence of outliers,
 \item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, pruning weights or inputs,
-\item Split of dataset into a training and test dataset, e.g., 2:1, 3:1, leave-one-out, etc., 
+\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
 \item Initial starting weights for the back-propagation algorithm, and
-\item Criteria for stopping model training iterations, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
+\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
 \end{itemize}
 
-A dataset from the \pkg{nycflights13} package \citep{Wickham14b} is used to demonstrate the effects of different training methods on model conclusions.
-This dataset describes flight data for all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
-The example focuses on all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}) at the destination airport.
-Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between desinations (\code{air\_time}), and distance flown (\code{distance}).
-All explanatory variables are normalized and the response variable is standardized to 0--1.
+A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions.
+This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
+The example uses all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.
+Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between desinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
+All explanatory variables were normalized and the response variable was standardized to 0--1.
 
-The effects of network architecture and initial starting weights on indications of variable importance were evaluated by creating a large number of models and assessing the uncertainty of the importance measures.
-Models with one, five, or ten hidden nodes and 100 separate models for each node level with a different set of random starting weights were evaluated.
-The relative importance of the variables was saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
+The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.
+Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.
+Each model had a random set of starting weights for the first training iteration.
+Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
 
-Several conclusions about the dataset and effects of network architecture can be made from the results in \cref{fig:flightimp}.
-First, delays in arrival time are consistently negatively related to distance travelled and positively related to departure delays and time in the air.
-That is, flights arrived later than their scheduled destination time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
-Second, the indications of importance varied in the range of magnitude between the models (i.e., one node  varied +/- 1 and the others +/- 3).
-This result suggests that the absolute values of importance only have relevance within a model, whereas only the relative rankings (e.g., least, most important) can be compared between models.
-Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture (i.e., effect of random starting weights).
-This conclusion emphasizes that a single model can often provide misleading information about variable importance such that several fits are often required before results are stabilized.
-Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect conclusions of variable importance and should be equally considered during model development.
+Several conclusions can be made from \cref{fig:flightimp}.
+First, delays in arrival time are negatively related to distance and positively related to departure delays and travel air.
+That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
+No conclusions can be made for the other variables because the bounds of uncertainty included zero.
+Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).
+This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.
+Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.
+This suggests that a single model can provide misleading information such that several models may be required to improve the certainty of conclusions.
+Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
 

### 2015-10-19 nnt_manu.Rnw
Sehan edits

@@
 A common criticism of the method is the inability to characterize relationships among variables from a fitted model.
-Although several methods proposed in recent years to `illuminate the black-box', they have not been made available in an open-source programming environment.
+Although several methods have been proposed to `illuminate the black-box', they have not been made available in an open-source programming environment.
 This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.
@@
 The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.
-Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
+Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
 \begin{equation}
-y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
+y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
 \end{equation}
-where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node.
+where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.
 
@@
 
-A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection \citep{Ripley96}.
-Models with skip layers include additional connections from the input to output layers that bypass the hidden layer.
+A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection (\cref{fig:plotnet_skip}).
+Models with skip layers include additional connections from the input to output layers that bypass the hidden layer \citep{Ripley96}.
 The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}.
@@
 
-\begin{figure}
+\begin{figure}[!h]
 \centering
@@
 \begin{itemize}
-\item Initial data pre-processing to normalize inputs, standardize response, or assess influence of outliers,
-\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, pruning weights or inputs,
+\item Initial data pre-processing to normalize inputs, standardize response, and assess influence of outliers,
+\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, and pruning weights or inputs,
 \item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
@@
 I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.
+Thanks to Sehan Lee for reviewing an earlier draft.
 Contributions and suggestions from online users have also greatly improved the utility of the package.

### 2015-10-19 nnt_manu.tex
Sehan edits

@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 79\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.

### 2015-11-30 nnt_manu.Rnw
manual bib formatting
revisions done from Marc's comments

@@
 A common criticism of the method is the inability to characterize relationships among variables from a fitted model.
-Although several methods have been proposed to `illuminate the black-box', they have not been made available in an open-source programming environment.
+Although several methods have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.
 This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.
@@
 
-Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Kelling09}.
+Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.
 A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
 Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data.
-Fields of research characterized by high-throughput data have a strong foundation in computationally-intensive methods of analysis (e.g., \citet{Saeys07}), whereas disciplines that have historically been limited by data quantity have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information (e.g., \citet{Swanson15}).
+Fields of research characterized by high-throughput data (e.g., bioinformatics, \citet{Saeys07}) have a strong foundation in computationally-intensive methods of analysis, whereas disciplines that have historically been limited by data quantity (e.g., field ecology, \citet{Swanson15}) have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information.
 Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
 
-Statistical methods that have been used to support data exploration are numerous \citep{Jain00}.
+Statistical methods that have been used to support data exploration are numerous \citep{Jain00,Recknagel06,Zuur10}.
 A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.
@@
 This model is typically used to predict the response of one or more variables given one to many explanatory variables.
-The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropation algorithm.
+The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropagation algorithm.
 Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.
@@
 Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
-As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
+As such, a common criticism is that neural networks are `black boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
 \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks.
@@
 Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.
-The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.
-The objectives are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.
+The reach of the package is extensive in that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.
+The objectives of this article are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.
 The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.
@@
 Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables.
-`Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.
-Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.
-`Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.
+Hidden layers between the input and response layers are key components of a neural network that mediate the transfer of information.
+Just as the input and response layers are composed of variables or nodes, each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.
+Bias layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.
 
-Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.
+Training a neural network model requires identifying the optimal weights that define the connections between the model layers.
 The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.
@@
 Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
-Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
+Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
 At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.
@@
 Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.
-The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.
+The \code{neuralweights} function can be used to retrieve model weights for any of the model classes described above.
 A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.
@@
 
-Existing plot functions in \proglang{R} to view neural networks is minimal.
+Existing plot functions in \proglang{R} to view neural networks are minimal.
 Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
@@
 
-<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.
+<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the `Optimal Brain Surgeon\' algorithm described in \\citet{Zell98}.
 The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
@@
 The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.
-In general, the following should be considered during initial development \cite{Ripley96, Lek00, Maier00}:
+In general, the following should be considered during initial development \citep{Ripley96, Lek00, Maier00}:
 \begin{itemize}
@@
 
-A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions.
+A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).
 This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
-The example uses all flights from the `UA' carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.
+The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.
 Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between desinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
@@
 Each model had a random set of starting weights for the first training iteration.
-Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (get \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
+Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
 
@@
 The popularity of the \ac{mlp} neural network is partially the cause of generalizations and misperceptions of their benefits as modelling tools \citep{Burke97}.
-Perhaps an overstatement, the `neural' component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.
+Perhaps an overstatement, the neural component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.
 Additionally, several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.
@@
 I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.
-Thanks to Sehan Lee for reviewing an earlier draft.
+Thanks to Sehan Lee and Marc Weber for reviewing an earlier draft.
 Contributions and suggestions from online users have also greatly improved the utility of the package.
@@
 % note that ref titles need to be in title-case 
-\bibliographystyle{jss}
-\bibliography{refs}
+% \bibliographystyle{jss}
+% \bibliography{refs}
+% this file is bbl edited by hand after running bibtex
+\input{CopyOfnnt_manu.bbl}
 

### 2015-11-30 nnt_manu.tex
manual bib formatting
revisions done from Marc's comments

@@
 Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
-Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
+Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
 At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 79\%.
@@
 
-\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the ``Optimal Brain Surgeon'' algorithm described in \citet{Zell98}.
+\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.
 The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}\label{fig:plotprune}

### 2015-11-30 nnt_manu.Rnw
contact update

@@
 %% almost as usual
-\author{Marcus W. Beck\\Oak Ridge Institute for Science and Education\\US Environmental Protection Agency}
+\author{Marcus W. Beck\\US Environmental Protection Agency}
 \title{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}
@@
   Marcus W. Beck\\
-  Oak Ridge Institude for Science and Education\\
   US Environmental Protection Agency\\

### 2015-12-08 nnt_manu.R
manu code added

@@ -1,0 +1,170 @@
+## # install from CRAN
+## install.packages('NeuralNetTools')
+## library(NeuralNetTools)
+## 
+## # install from GitHub
+## install.package('devtools')
+## library(devtools)
+## install_github('fawda123/NeuralNetTools')
+## library(NeuralNetTools)
+library(NeuralNetTools)
+
+# mlp object, RSNNS package
+library(RSNNS)
+x <- neuraldat[, c('X1', 'X2', 'X3')]
+y <- neuraldat[, 'Y1']
+mod1 <- mlp(x, y, size = 5)
+
+# nn object, neuralnet package
+library(neuralnet)
+mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
+
+#nnet object, nnet package
+library(nnet)
+mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
+par(mar = c(0, 0, 0, 0))
+plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
+plotnet(mod3)
+# create a model with a skip layer
+modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
+
+# plot
+par(mar = c(0, 0, 0, 0))
+plotnet(modskip, skip = TRUE)
+plotnet(modskip)
+# pruned model using RSNNS
+pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
+mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+ pruneFuncParams = pruneFuncParams)
+
+# plot, pruned connections are omitted (default) or included with prune_col
+par(mar = c(0, 0, 0, 0))
+plotnet(mod, rel_rsc = c(3, 8))
+plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
+
+# garson and olden functions for each model
+garson(mod1)
+olden(mod1)
+garson(mod2)
+olden(mod2)
+garson(mod3)
+olden(mod3)
+
+# lek profile, quantile evaluation
+lekprofile(mod3)
+
+# lek profile, quantile groups
[…]

### 2015-12-08 cover_letter.tex
cover letter

\documentclass[a4paper,12pt]{article} 
\usepackage{times}
\usepackage[top=1in,bottom=1in,left=1in,right=1in,nohead]{geometry}
\usepackage{graphics}
\usepackage{hyperref}

\begin{document}
\renewcommand{\rmdefault}{ptm}
\pagestyle{empty}

\setlength{\parindent}{0mm} 
\setlength{\parskip}{5mm}

\begin{flushright}
\today
\end{flushright}

\emph{To:}\newline
Bettina Gr\"un, Edzer Pebesma, Achim Zeileis\newline
Editors-in-Chief\newline
Journal of Statistical Software

\emph{From:}\newline
Marcus W. Beck\newline
US Environmental Protection Agency\newline
beck.marcus@epa.gov, 850-934-2480\vspace{0.1in}

Enclosed please find my manuscript, entitled `NeuralNetTools: Visualization and Analysis Tools for Neural Networks', to be considered as a research article in the Journal of Statistical Software.

Supervised neural networks are statistical models that have been used to support prediction and exploratory analysis in data-rich environments.  A common criticism of these methods is that they represent `black-boxes' that do not describe any information about relationships among variables.  Methods to quantify information from a fitted neural network have been previously described in the literature but have not been available in a statistical computing environment.  This article describe the NeuralNetTools package for R that was developed to better quantify information provided by a neural network, including methods for plotting using a neural network interpretation diagram, estimates of variable importance, and a sensitivity analysis.  The package is exhaustive in scope by including methods for the most commonly used neural network objects in R.

Accompanying this submission is the manuscript PDF created with the JSS \LaTeX\ template, an R script of the manuscript code, and links to the package source code at the bottom of this letter.  Please note that the current version of the package on CRAN was submitted this year on Dec. 1 for minor revisions related to manuscript content.  The original package was first uploaded to CRAN on December 16\textsuperscript{th}, 2014.  Accordingly, the package has been used by the R community for a sufficient amount of time to address bugs or issues (see \href{https://github.com/fawda123/NeuralNetTools/issues?q=is%3Aissue+is%3Aclosed}{here}) in the early releases.

I greatly appreciate the opportunity to publish my work in JSS.  Please feel free to contact me directly should additional information be needed.

\vspace{0.1in}

\hspace{4.5in}Respectfully,

\hspace{4.5in}Dr. Marcus W. Beck

{\it Package sources}

Github (dev): \url{https://github.com/fawda123/NeuralNetTools}

CRAN (stable): \url{https://cran.r-project.org/web/packages/NeuralNetTools/index.html}

\end{document}

### 2015-12-14 nnt_manu.Rnw
cover letter date change, type in abstract

@@
 Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.
-Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
+Methods are provided for objects from many of the common neural network packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
 The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.

### 2016-01-12 nnt_manu.Rnw
everything sweave, no more knitr

@@
 
-% knitr options
-<<setup, include = F, cache = F>>=
-library(knitr)
-# set global chunk options
-opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T, background = 'white', highlight = F, prompt = TRUE)
-options(replace.assign = TRUE, digits = 1, prompt="R> ")
+% options and libs
+<<echo = FALSE>>=
+# opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T, background = 'white', highlight = F, prompt = TRUE)
+options(digits = 1, prompt = 'R> ')
 
@@
 % get online bib file
-<<echo = FALSE, cache = FALSE>>=
+<<echo = FALSE>>=
 refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
@@
 \begin{document}
+\SweaveOpts{concordance=FALSE}
 
@@
 
-<<eval = T, echo= F, message = F>>=
+<<eval = T, echo = F, message = F>>=
 library(NeuralNetTools)
 @
-<<results = 'hide'>>=
+<<results = hide>>=
 # set random seed for initial weights for training
@@
 
-<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (\\ref{fig:plotnet1}) and using the \\acl{nid} (\\ref{fig:plotnet2}).
-Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).', fig.subcap = c('', '')>>=
+<<echo = FALSE, results = hide>>=
+pdf('figs/plotnet.pdf', height = 5.5, width = 5.5, family = 'serif')
 par(mar = c(0, 0, 0, 0))
@@
 plotnet(mod3)
+dev.off()
 @
+\begin{figure}[!ht]
+\subfloat[][]{
+\includegraphics[width=7.5cm,page=1]{figs/plotnet.pdf}
+\label{fig:plotnet1}
+}
+\subfloat[][]{
+\includegraphics[width=7.5cm, page=2]{figs/plotnet.pdf}
+\label{fig:plotnet2}
+}
+\caption{Examples from the \code{plotnet} function showing neural networks as a standard graphic (\ref{fig:plotnet1}) and using the \acl{nid} (\ref{fig:plotnet2}).
+Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).}
+\label{fig:plotnet}
+\end{figure}
 
@@
 
-<<plotnet_skip, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
-The skip layer (\\ref{fig:plotnet_skip1}) and primary neural network (\\ref{fig:plotnet_skip2}) can be viewed separately with \\code{plotnet} by using \\code{skip = TRUE} or \\code{skip = FALSE}.', results = 'hide', fig.subcap = c('', '')>>=
+<<echo = FALSE, results = hide>>=
 # create a model with a skip layer
@@
 # plot
+pdf('figs/plotnet_skip.pdf', height = 5.5, width = 5.5, family = 'serif')
 par(mar = c(0, 0, 0, 0))
@@
 plotnet(modskip)
+dev.off()
 @
+\begin{figure}[!ht]
+\subfloat[][]{
+\includegraphics[width=7.5cm,page=1]{figs/plotnet_skip.pdf}
+\label{fig:plotnet_skip1}
+}
+\subfloat[][]{
+\includegraphics[width=7.5cm, page=2]{figs/plotnet_skip.pdf}
+\label{fig:plotnet_skip2}
+}
+\caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
+The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet} by using \code{skip = TRUE} or \code{skip = FALSE}.}
+\label{fig:plotnet_skip}
+\end{figure}
 
@@
 
-<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the `Optimal Brain Surgeon\' algorithm described in \\citet{Zell98}.
-The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
+<<eval = F, echo = F, results = hide>>=
+# pruned model using RSNNS
+pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
+mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+ pruneFuncParams = pruneFuncParams)
+
+# plot, pruned connections are omitted (default) or included with prune_col
+pdf('figs/plotprune.pdf', height = 7, width = 7.7, family = 'serif')
+par(mar = c(0, 0, 0, 0))
+plotnet(mod, rel_rsc = c(3, 8))
+plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
+dev.off()
+@
+<<eval = F, echo = T, results = hide>>=
 # pruned model using RSNNS
@@
 @
+\begin{figure}[!ht]
+\subfloat[][]{
+\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
+\label{fig:plotprune1}
+}
+\subfloat[][]{
+\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
+\label{fig:plotprune2}
+}
+\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.
+The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
+\label{fig:plotprune}
+\end{figure}
 
@@
 @
-
+<<results = hide, echo = FALSE>>=
+pdf('figs/plotimp.pdf', family = 'serif', height = 2, width = 4)
+garson(mod1)
+olden(mod1)
+garson(mod2)
+olden(mod2)
+garson(mod3)
+olden(mod3)
+dev.off()
+@
 \begin{figure}[!h]
@@
 \subfloat[][Model 1 (\code{mlp}), \code{garson}]{
-<<plotimp-1, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
-garson(mod1)
-@
+\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
 \label{fig:plotimp1}
@@
 \subfloat[][Model 1 (\code{mlp}), \code{olden}]{
-<<plotimp-2, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
-olden(mod1)
-@
+\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
 \label{fig:plotimp2}
@@
 \subfloat[][Model 2 (\code{nn}), \code{garson}]{
-<<plotimp-3, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
-garson(mod2)
-@
+\includegraphics[width = 0.47\textwidth, page = 3]{figs/plotimp.pdf}
 \label{fig:plotimp3}
@@
 \subfloat[][Model 2 (\code{nn}), \code{olden}]{
-<<plotimp-4, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
-olden(mod2)
-@
+\includegraphics[width = 0.47\textwidth, page = 4]{figs/plotimp.pdf}
 \label{fig:plotimp4}
@@
 \subfloat[][Model 3 (\code{nnet}), \code{garson}]{
-<<plotimp-5, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
-garson(mod3)
-@
+\includegraphics[width = 0.47\textwidth, page = 5]{figs/plotimp.pdf}
 \label{fig:plotimp5}
@@
 \subfloat[][Model 3 (\code{nnet}), \code{olden}]{
-<<plotimp-6, fig.height = 2, fig.width =4, out.width = '0.47\\textwidth', echo = F>>=
-olden(mod3)
-@
+\includegraphics[width = 0.47\textwidth, page = 6]{figs/plotimp.pdf}
 \label{fig:plotimp6}
@@
 @
+<<echo = F, results = hide>>=
+pdf('figs/plotlek_sens.pdf', height = 2.5, width = 8, family = 'serif')
+lekprofile(mod3) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+lekprofile(mod3, group_vals = 6) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+dev.off()
 
-\begin{figure}
+pdf('figs/plotlek_bars.pdf', height = 2.5, width = 8, family = 'serif')
+lekprofile(mod3, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+lekprofile(mod3, group_vals = 6, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
+dev.off()
+@
+\begin{figure}[!ht]
 \centering
 \subfloat[][Response with quantiles]{
-<<plotlek_sens1, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
-lekprofile(mod3) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
-@
+\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
 \label{fig:plotlek_sens1}
@@
 \subfloat[][Response with clusters]{
-<<plotlek_sens2, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
-lekprofile(mod3, group_vals = 6) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
-@
+\includegraphics[width = 0.9\textwidth, page = 2]{figs/plotlek_sens.pdf}
 \label{fig:plotlek_sens2}
-}s
+}
 \caption{Sensitivity analysis of a neural network using the Lek profile method to evaluate the effects of explanatory variables.
@@
 \subfloat[][Quantile groupings]{
-<<plotlek_bars1, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
-lekprofile(mod3, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
-@
+\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_bars.pdf}
 \label{fig:plotlek_bars1}
@@
 \subfloat[][Cluster groupings]{
-<<plotlek_bars2, fig.height = 2.5, fig.width = 8, out.width = '0.9\\textwidth', echo = F>>=
-lekprofile(mod3, group_vals = 6, group_show = TRUE) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
-@
+\includegraphics[width = 0.9\textwidth, page = 2]{figs/plotlek_bars.pdf}
 \label{fig:plotlek_bars2}
@@
 Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
-<<echo = F>>=
+<<echo = F, results = hide>>=
 # library(nnet)
@@
 
+pdf('figs/flightimp.pdf', height = 3, width = 8, family = 'serif')
+print(p1)
+print(p2)
+print(p3)
+dev.off()
 @
-
 \begin{figure}[!h]
@@
 \subfloat[][Networks with one node]{
-<<flightimp-1, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
-print(p1)
-@
+\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
 \label{fig:flightimp1}
@@
 \subfloat[][Networks with five nodes]{
-<<flightimp-2, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
-print(p2)
-@
+\includegraphics[width = 0.95\textwidth, page = 2]{figs/flightimp.pdf}
 \label{fig:flightimp2}
@@
 \subfloat[][Networks with ten nodes]{
-<<flightimp-3, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
-print(p3)
-@
+\includegraphics[width = 0.95\textwidth, page = 3]{figs/flightimp.pdf}
 \label{fig:flightimp3}

### 2016-01-12 nnt_manu.tex
everything sweave, no more knitr

@@
-\documentclass[article,shortnames]{jss}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass[article,shortnames]{jss}
 
@@
 
-% knitr options
-
+% options and libs
 % get online bib file
@@
 %% need no \usepackage{Sweave.sty}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 79\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 80\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 R> # install from CRAN
@@
 R> library(NeuralNetTools)
-R> 
 R> # install from GitHub
@@
 R> library(NeuralNetTools)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 R> # set random seed for initial weights for training
 R> set.seed(123)
-R> 
 R> # mlp object, RSNNS package
@@
 R> mod1 <- mlp(x, y, size = 5)
-R> 
 R> # nn object, neuralnet package
@@
 R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
-R> 
 R> #nnet object, nnet package
@@
 R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-R> par(mar = c(0, 0, 0, 0))
-R> plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
-R> plotnet(mod3)
-\end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \subfloat[\label{fig:plotnet1}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-1} }
-\subfloat[\label{fig:plotnet2}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet-2} }
-
+\begin{figure}[!ht]
+\subfloat[][]{
+\includegraphics[width=7.5cm,page=1]{figs/plotnet.pdf}
+\label{fig:plotnet1}
 }
-
+\subfloat[][]{
+\includegraphics[width=7.5cm, page=2]{figs/plotnet.pdf}
+\label{fig:plotnet2}
+}
 \caption{Examples from the \code{plotnet} function showing neural networks as a standard graphic (\ref{fig:plotnet1}) and using the \acl{nid} (\ref{fig:plotnet2}).
-Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).}\label{fig:plotnet}
+Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).}
+\label{fig:plotnet}
 \end{figure}
-
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-R> # create a model with a skip layer
-R> modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
-R> 
-R> # plot
-R> par(mar = c(0, 0, 0, 0))
-R> plotnet(modskip, skip = TRUE)
-R> plotnet(modskip)
-\end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \subfloat[\label{fig:plotnet_skip1}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-1} }
-\subfloat[\label{fig:plotnet_skip2}]{\includegraphics[width=7.5cm,height=7.5cm]{figs/plotnet_skip-2} }
-
+\begin{figure}[!ht]
+\subfloat[][]{
+\includegraphics[width=7.5cm,page=1]{figs/plotnet_skip.pdf}
+\label{fig:plotnet_skip1}
 }
-
+\subfloat[][]{
+\includegraphics[width=7.5cm, page=2]{figs/plotnet_skip.pdf}
+\label{fig:plotnet_skip2}
+}
 \caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.
-The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet} by using \code{skip = TRUE} or \code{skip = FALSE}.}\label{fig:plotnet_skip}
+The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet} by using \code{skip = TRUE} or \code{skip = FALSE}.}
+\label{fig:plotnet_skip}
 \end{figure}
-
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 R> # pruned model using RSNNS
@@
 +  pruneFuncParams = pruneFuncParams)
-R> 
 R> # plot, pruned connections are omitted (default) or included with prune_col
@@
 R> plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
-\end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \subfloat[\label{fig:plotprune1}]{\includegraphics[width=0.48\textwidth]{figs/plotprune-1} }
-\subfloat[\label{fig:plotprune2}]{\includegraphics[width=0.48\textwidth]{figs/plotprune-2} }
-
+\end{Sinput}
+\end{Schunk}
+\begin{figure}[!ht]
+\subfloat[][]{
+\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
+\label{fig:plotprune1}
 }
-
+\subfloat[][]{
+\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
+\label{fig:plotprune2}
+}
 \caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.
-The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}\label{fig:plotprune}
+The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
+\label{fig:plotprune}
 \end{figure}
-
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 R> # garson and olden functions for each model
@@
 R> olden(mod3)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
+\end{Sinput}
+\end{Schunk}
 \begin{figure}[!h]
@@
 \subfloat[][Model 1 (\code{mlp}), \code{garson}]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-1-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
 \label{fig:plotimp1}
@@
 \subfloat[][Model 1 (\code{mlp}), \code{olden}]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-2-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
 \label{fig:plotimp2}
@@
 \subfloat[][Model 2 (\code{nn}), \code{garson}]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-3-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.47\textwidth, page = 3]{figs/plotimp.pdf}
 \label{fig:plotimp3}
@@
 \subfloat[][Model 2 (\code{nn}), \code{olden}]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-4-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.47\textwidth, page = 4]{figs/plotimp.pdf}
 \label{fig:plotimp4}
@@
 \subfloat[][Model 3 (\code{nnet}), \code{garson}]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-5-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.47\textwidth, page = 5]{figs/plotimp.pdf}
 \label{fig:plotimp5}
@@
 \subfloat[][Model 3 (\code{nnet}), \code{olden}]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.47\textwidth]{figs/plotimp-6-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.47\textwidth, page = 6]{figs/plotimp.pdf}
 \label{fig:plotimp6}
@@
 For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 R> # lek profile, quantile evaluation
 R> lekprofile(mod3)
-R> 
 R> # lek profile, quantile groups 
 R> lekprofile(mod3, group_show = TRUE)
-R> 
 R> # lek profile, cluster evaluation
 R> lekprofile(mod3, group_vals = 6)
-R> 
 R> # lek profile, cluster groups
 R> lekprofile(mod3, group_vals = 6, group_show = TRUE)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\begin{figure}
+\end{Sinput}
+\end{Schunk}
+\begin{figure}[!ht]
 \centering
 \subfloat[][Response with quantiles]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_sens1-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
 \label{fig:plotlek_sens1}
@@
 \subfloat[][Response with clusters]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_sens2-1}
-
+\includegraphics[width = 0.9\textwidth, page = 2]{figs/plotlek_sens.pdf}
+\label{fig:plotlek_sens2}
 }
-
-\end{knitrout}
-\label{fig:plotlek_sens2}
-}s
 \caption{Sensitivity analysis of a neural network using the Lek profile method to evaluate the effects of explanatory variables.
@@
 \subfloat[][Quantile groupings]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_bars1-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_bars.pdf}
 \label{fig:plotlek_bars1}
@@
 \subfloat[][Cluster groupings]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.9\textwidth]{figs/plotlek_bars2-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.9\textwidth, page = 2]{figs/plotlek_bars.pdf}
 \label{fig:plotlek_bars2}
@@
 Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
-
 \begin{figure}[!h]
@@
 \subfloat[][Networks with one node]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.95\textwidth]{figs/flightimp-1-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
 \label{fig:flightimp1}
@@
 \subfloat[][Networks with five nodes]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.95\textwidth]{figs/flightimp-2-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.95\textwidth, page = 2]{figs/flightimp.pdf}
 \label{fig:flightimp2}
@@
 \subfloat[][Networks with ten nodes]{
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
-
-{\centering \includegraphics[width=0.95\textwidth]{figs/flightimp-3-1}
-
-}
-
-\end{knitrout}
+\includegraphics[width = 0.95\textwidth, page = 3]{figs/flightimp.pdf}
 \label{fig:flightimp3}

### 2016-01-12 nnt_manu.Rnw
no comments in code chunks

@@
 
-The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
+The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:
 
 <<eval = F, echo = T>>=
-# install from CRAN
 install.packages('NeuralNetTools')
-library(NeuralNetTools)
-
-# install from GitHub
-install.package('devtools')
-library(devtools)
-install_github('fawda123/NeuralNetTools')
 library(NeuralNetTools)
@@
 <<results = hide>>=
-# set random seed for initial weights for training
 set.seed(123)
-
-# mlp object, RSNNS package
 library(RSNNS)
@@
 mod1 <- mlp(x, y, size = 5)
-
-# nn object, neuralnet package
 library(neuralnet)
 mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
-
-#nnet object, nnet package
 library(nnet)
@@
 <<eval = F, echo = T, results = hide>>=
-# pruned model using RSNNS
 pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
@@
  pruneFuncParams = pruneFuncParams)
-
-# plot, pruned connections are omitted (default) or included with prune_col
-par(mar = c(0, 0, 0, 0))
 plotnet(mod, rel_rsc = c(3, 8))
@@
 The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.
-
+Variable importance shown in \Cref{fig:plotimp} was estimated for each model:
 <<eval = F>>=
-# garson and olden functions for each model
 garson(mod1)
@@
 Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
-For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
+For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
+Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
 <<eval = F>>=
-# lek profile, quantile evaluation
 lekprofile(mod3)
-
-# lek profile, quantile groups 
 lekprofile(mod3, group_show = TRUE)
-
-# lek profile, cluster evaluation
 lekprofile(mod3, group_vals = 6)
-
-# lek profile, cluster groups
 lekprofile(mod3, group_vals = 6, group_show = TRUE)

### 2016-01-12 nnt_manu.tex
no comments in code chunks

@@
 
-The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:
+The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:
 
@@
 \begin{Sinput}
-R> # install from CRAN
 R> install.packages('NeuralNetTools')
-R> library(NeuralNetTools)
-R> # install from GitHub
-R> install.package('devtools')
-R> library(devtools)
-R> install_github('fawda123/NeuralNetTools')
 R> library(NeuralNetTools)
@@
 \begin{Sinput}
-R> # set random seed for initial weights for training
 R> set.seed(123)
-R> # mlp object, RSNNS package
 R> library(RSNNS)
@@
 R> mod1 <- mlp(x, y, size = 5)
-R> # nn object, neuralnet package
 R> library(neuralnet)
 R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
-R> #nnet object, nnet package
 R> library(nnet)
@@
 \begin{Sinput}
-R> # pruned model using RSNNS
 R> pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
@@
 +  pruneFuncParams = pruneFuncParams)
-R> # plot, pruned connections are omitted (default) or included with prune_col
-R> par(mar = c(0, 0, 0, 0))
 R> plotnet(mod, rel_rsc = c(3, 8))
@@
 The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.
-
+Variable importance shown in \Cref{fig:plotimp} was estimated for each model:
 \begin{Schunk}
 \begin{Sinput}
-R> # garson and olden functions for each model
 R> garson(mod1)
@@
 Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
-For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
+For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
+Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
 \begin{Schunk}
 \begin{Sinput}
-R> # lek profile, quantile evaluation
 R> lekprofile(mod3)
-R> # lek profile, quantile groups 
 R> lekprofile(mod3, group_show = TRUE)
-R> # lek profile, cluster evaluation
 R> lekprofile(mod3, group_vals = 6)
-R> # lek profile, cluster groups
 R> lekprofile(mod3, group_vals = 6, group_show = TRUE)

### 2016-01-12 nnt_manu.Rnw
prescreening revision done

@@
 \acrodef{nid}[NID]{neural interpretation diagram}
+
+\crefname{figure}{Figure}{Figures}
 
@@
 }
-\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
+\Keywords{neural networks, \code{plotnet}, sensitivity, variable importance, \proglang{R}}
 \Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
@@
 <<echo = FALSE>>=
-# opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, cache = T, background = 'white', highlight = F, prompt = TRUE)
 options(digits = 1, prompt = 'R> ')
@@
 # other libs
-library(ggplot2)
+library("ggplot2")
 @
@@
 
-An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of model information.
+An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but complicates the extraction of model information.
 Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
 As such, a common criticism is that neural networks are `black boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
-\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks.
+\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information about variable relationships from neural networks.
 Many of these methods were previously described but not commonly used.
@@
 
-\section[Theoretical foundation]{Theoretical foundation and existing R packages}
+\section[Theoretical foundation]{Theoretical foundation and existing \proglang{R} packages}
 
@@
 The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.
-Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.
+Training is commonly achieved using the backpropagation algorithm described in \citet{Rumelhart86}.
 This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.
 The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.
-The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.
+The weights are then changed through a backpropagation step that begins by changing weights in the output layer and then the remaining hidden layers.
 The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.
-Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
+A fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
 \begin{equation}
@@
 \end{equation}
-where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.
+where the estimated value $k$ of the response variable $y$ is a sum of products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.
 
@@
 # make sure to change cdate
-library(cranlogs)
+library("cranlogs")
 today <- as.character(as.Date(Sys.time()))
@@
 Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
-Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
+Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
 At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
-Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.
+Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class).
 Additional \code{default} or \code{numeric} methods are available for some of the generic functions.
@@
 <<eval = F, echo = T>>=
-install.packages('NeuralNetTools')
-library(NeuralNetTools)
+install.packages("NeuralNetTools")
+library("NeuralNetTools")
 @
@@
 
-A common approach for data preprocessing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
+A common approach for data pre-processing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
 A sample dataset that follows this format is included with \pkg{NeuralNetTools}.
@@
 The response variables are also standardized from zero to one.
-Additional datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
+Additional datasets can be pre-processed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
 The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.
@@
 <<eval = T, echo = F, message = F>>=
-library(NeuralNetTools)
+library("NeuralNetTools")
 @
@@
 set.seed(123)
-library(RSNNS)
+library("RSNNS")
 x <- neuraldat[, c('X1', 'X2', 'X3')]
@@
 mod1 <- mlp(x, y, size = 5)
-library(neuralnet)
+library("neuralnet")
 mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
-library(nnet)
+library("nnet")
 mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
@@
 Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid} \citep[\acs{nid},][]{Ozesmi99} and includes several options to customize aesthetics.
-A \ac{nid} is a modification to the standard conceptual illustration of the \ac{mlp} network that changes the thickness and color of the weight connections based on magnitude and sign, respectively.
+A \ac{nid} is a modification of the standard conceptual illustration of the \ac{mlp} network that changes the thickness and color of the weight connections based on magnitude and sign, respectively.
 Positive weights between layers are shown as black lines and negative weights as grey lines.
-Line thickness is in proportion to absolute magnitude of each weight (\cref{fig:plotnet}).
+Line thickness is proportional to the absolute magnitude of each weight (\cref{fig:plotnet}).
 
@@
 This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
-Pruning has additional benefits of an increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.
-Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
+Pruning has the additional benefit of increasing power by increasing the degrees of freedom or the identification of input nodes with low importance.
+Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
 The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
@@
 
-The primary benefits of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in connections between the layers.
+The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in connections between the layers.
 Although useful as a general tool, the \ac{nid} can be difficult to interpret given the amount of weighted connections in most networks.
@@
 For each input node, all weights connecting an input through the hidden layer to the response variable are identified to return a list of all weights specific to each input variable.
-A summed product of the connections for each input node are then scaled relative to all other inputs.
+Summed products of the connections for each input node are then scaled relative to all other inputs.
 A value for each input node indicates relative importance as the absolute magnitude from zero to one.
@@
 An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign.
-For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect, whereas Garson's algorithm may provide different results based on the absolute magnitude.
-An additional advantage is that Olden's algorithm can evaluate neural networks with multiple hidden layers and response variables.
+For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect, whereas \code{garson} may provide different results based on the absolute magnitude.
+An additional advantage is that the \code{olden} function can evaluate neural networks with multiple hidden layers and response variables.
 The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.
@@
 
-An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}.
+An alternative approach to evaluate variable interactions in a neural network is the Lek profile method \citep{Lek96,Gevrey03}.
 The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
@@
 This is implemented by creating a matrix where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables.
-All explanatory variables are held at a chosen constant value (e.g., median), while the variable of interest is sequenced from its minimum to maximum value across the range of observations.
+All explanatory variables are held constant (e.g., median) while the variable of interest is sequenced from its minimum to maximum.
 This matrix is then used to predict values of the response variable from a fitted model object.
@@
 Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable).
-The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
-Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one.
-The centers (means) of the clusters are then used as constant values for the unevaluated variables.
+The second option holds unevaluated variables at means defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
+Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer greater than one.
+The centers (means) of the clusters are then used as constants for the unevaluated variables.
 \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.
-Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.
+Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more interpretable response curves.
 Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
@@
 \item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
-\item Initial starting weights for the back-propagation algorithm, and
+\item Initial starting weights for the backpropagation algorithm, and
 \item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
@@
 The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.
-Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between desinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
+Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
 All explanatory variables were normalized and the response variable was standardized to 0--1.
@@
 Several conclusions can be made from \cref{fig:flightimp}.
-First, delays in arrival time are negatively related to distance and positively related to departure delays and travel air.
+First, delays in arrival time are negatively related to distance and positively related to departure delays and air time.
 That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
@@
 Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.
-This suggests that a single model can provide misleading information such that several models may be required to improve the certainty of conclusions.
-Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
+This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.
+Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
 <<echo = F, results = hide>>=
-# library(nnet)
-# library(NeuralNetTools)
-# library(dplyr)
-# library(tidyr)
-# library(ggplot2)
-# library(nycflights13)
+# library("nnet")
+# library("NeuralNetTools")
+# library("dplyr")
+# library("tidyr")
+# library("ggplot2")
+# library("nycflights13")
 # 
@@
 
-library(dplyr)
-library(ggplot2)
+library("dplyr")
+library("ggplot2")
 
@@
 
-A primary objective of the package is to address the myth that supervised neural networks are `black boxes' that provide no information about underlying relationships between variables \citep{Paruelo97,Olden02}.
-Although neural networks are considered complex relative to more conventional approaches, the theoretical foundation has many parallels with traditional statistical techniques that can allow insight into causation \citep{Cheng94}.
+A primary objective of the package is to address the concern that supervised neural networks are `black boxes' that provide no information about underlying relationships between variables \citep{Paruelo97,Olden02}.
+Although neural networks are considered relatively complex statistical models, the theoretical foundation has many parallels with simpler statistical techniques that provide for evaluation of variable important \citep{Cheng94}.
 Moreover, the model fitting process minimizes error using a standard objective function such that conventional techniques to evaluate model sensitivity or performance (e.g., cross-validation) can be used with neural networks.
@@
 
-A more compelling issue is when and how to apply neural networks given alternative methods of analysis.
-The popularity of the \ac{mlp} neural network is partially the cause of generalizations and misperceptions of their benefits as modelling tools \citep{Burke97}.
+Another important issue is determining when and how to apply neural networks given availability of alternative methods of analysis.
+The popularity of the \ac{mlp} neural network is partly to blame  for the generalizations and misperceptions about their benefits as modelling tools \citep{Burke97}.
 Perhaps an overstatement, the neural component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.
@@
 Considerations for choosing a method may include power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant considerations of exploratory data analysis \citep{Zuur10}.
-\pkg{NeuralNetTools} provides analysis tools that can inform a more balanced assessment given alternative and complementary methods for data exploration.
+\pkg{NeuralNetTools} provides analysis tools that can inform evaluation and selection from among several alternative methods for exploratory data analysis.
 

### 2016-01-12 nnt_manu.tex
prescreening revision done

@@
 Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
-Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust back-propagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
+Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
 At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 80\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
-Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.
+Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class).
 Additional \code{default} or \code{numeric} methods are available for some of the generic functions.
@@
 \begin{Sinput}
-R> install.packages('NeuralNetTools')
-R> library(NeuralNetTools)
+R> install.packages("NeuralNetTools")
+R> library("NeuralNetTools")
 \end{Sinput}
@@
 R> set.seed(123)
-R> library(RSNNS)
+R> library("RSNNS")
 R> x <- neuraldat[, c('X1', 'X2', 'X3')]
@@
 R> mod1 <- mlp(x, y, size = 5)
-R> library(neuralnet)
+R> library("neuralnet")
 R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
-R> library(nnet)
+R> library("nnet")
 R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
@@
 Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.
-This suggests that a single model can provide misleading information such that several models may be required to improve the certainty of conclusions.
-Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be equally considered during model development.
+This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.
+Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
 \begin{figure}[!h]

### 2016-01-12 nnt_manu.R

@@ -1,50 +1,34 @@
-## # install from CRAN
-## install.packages('NeuralNetTools')
-## library(NeuralNetTools)
-## 
-## # install from GitHub
-## install.package('devtools')
-## library(devtools)
-## install_github('fawda123/NeuralNetTools')
-## library(NeuralNetTools)
-library(NeuralNetTools)
+## install.packages("NeuralNetTools")
+library("NeuralNetTools")
 
-# mlp object, RSNNS package
-library(RSNNS)
+set.seed(123)
+library("RSNNS")
 x <- neuraldat[, c('X1', 'X2', 'X3')]
 y <- neuraldat[, 'Y1']
 mod1 <- mlp(x, y, size = 5)
-
-# nn object, neuralnet package
-library(neuralnet)
+library("neuralnet")
 mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
-
-#nnet object, nnet package
-library(nnet)
+library("nnet")
 mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
+
 par(mar = c(0, 0, 0, 0))
 plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
 plotnet(mod3)
-# create a model with a skip layer
+
 modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
 
-# plot
 par(mar = c(0, 0, 0, 0))
 plotnet(modskip, skip = TRUE)
 plotnet(modskip)
-# pruned model using RSNNS
+
 pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
-  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01,
   init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
 mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
  pruneFuncParams = pruneFuncParams)
-
-# plot, pruned connections are omitted (default) or included with prune_col
-par(mar = c(0, 0, 0, 0))
 plotnet(mod, rel_rsc = c(3, 8))
 plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
 
-# garson and olden functions for each model
 garson(mod1)
 olden(mod1)
[…]

### 2016-11-22 nnt_manu.tex
commit before jss revision

@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 80\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 95\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.

### 2016-11-22 review_response.Rnw
revision draft done, need to proof and fix applied example

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{indentfirst}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[final]{animate}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{array}
\usepackage{subfig} 
\usepackage[noae]{Sweave}
\usepackage{cleveref}
\usepackage[figureposition=bottom]{caption}
\usepackage{paralist}
\usepackage{acronym}
\usepackage{outlines}
\usepackage{amsmath}

%acronyms
% \acrodef{}{}

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
options(replace.assign=TRUE,width=90,digits=1)
@

\setlength{\parskip}{5mm}   
\setlength{\parindent}{0in}

\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}

\begin{document}
\SweaveOpts{concordance=TRUE}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to reviewer comments ``NeuralNetTools: Visualization and Analysis Tools for Neural Networks'' by M. W. Beck.}

{\it I thank the reviewer for providing thoughtful comments on my manuscript and the NeuralNetTools package.  My response to these comments are shown in italics.  Page and paragraph numbers refer to the original manuscript.}

Reviewer C:

The manuscript identifies the need for open-source tools for visualizing the structure of a multilayer perceptron (MLP) neural network and for visualizing variable importance.  The author provides an intuitive background of neural networks and a brief description of previous publications that develop tools for visualizing and understanding the MLP neural network.  The supplied code works (mostly) as advertised (see below for discrepancies).  Overall the paper is well written with a logical flow.  The simple examples adequately demonstrate the tools.  However, in my opinion the applied example does not fully draw-out the additional interpretation that is claimed by the NeuralNetTools package.  Furthermore, there is limited usefulness in visualizing a neural network when there are more than about 10 input variables.  (Although the same problem applies to visualizing a tree-based model when there are many terminal nodes.)  Still I believe this tool could be useful for data sets with a small number of variables.

{\it I certainly agree that plotting a neural network with many variables is impractical, but as noted, this is an issue with any plotting method for large structures.  I believe the main benefit of the package is in the supporting functions, e.g., \texttt{garson}, \texttt{lekprofile}, \texttt{olden}, that use the structure of the model (weights) and predictions to better understand relationships between variables. The \texttt{plotnet} function has importance as a basic visualization tool that informs additional analysis. Our hope is that this is clear from the existing text in the manuscript that immediately follows the plot section (i.e., first paragraph, sec. 3.2).  I have added some text to the conclusions to re-emphasize this issue. First paraggraph of sec. 5: ``Although visualizing a neural network with \texttt{plotnet} is impractical for large models, the remaining functions can simplify model complexity to identify important relationships between variables.''

The applied example section has also been revised to provide a more comprehensive demonstration of the utility of the package.  Please see the specific response below.
}

I would recommend tentatively accepting the manuscript with the author addressing the following items:

$\bullet$ I would like to see a different applied example (or the current example changed) that gives a better practical demonstration of the ability to visualize a network and the corresponding interpretation.

{\it Content was added to the section that shows the interpretation of a neural network model using functions in the package.  Figure 7 was added to show the results from a simple model with an interpretation provided in the text.  Most of the addition to the text was included in the following paragraph (but see the applied example section for the entire revision):

``Figure 7 shows the information about arrival delays that can be obtained with the functions in \textbf{NeuralNetTools}. The NID (7a) shows the model structure and can be used develop a general characterization of the relationships between variables.  For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays. Figures 7b and 7c provide more quantitative descriptions using information from both the NID and model predictions.  Figure 7b shows variable importance using the \texttt{garson} and \texttt{olden} algorithms.  The \texttt{garson} function suggests time between destinations (\texttt{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \texttt{olden} method.  However, the \texttt{garson} function shows arrival time as the second most important variable, whereas the \texttt{olden} function shows this variable as having a weak negative relationship with arrival time. This discrepancy is explained below. Finally, results from the \texttt{lekprofile} function (7c) confirm those in 7b, with the addition of non-linear responses that vary by different groupings of the data.  Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group five that generally had early arrival times and long departure delays.''    
}

$\bullet$ Abstract:  please clarify what is meant by ``unstructured datasets.''

{\it This was changed to ``...among multiple variables''.}

$\bullet$ Introduction:  I do not agree that ``Data-intensive analysis is a relatively new research approach...''

{\it The first sentence was revised: ``A common objective of data-intensive analysis is the synthesis of unstructured information to identify patterns or trends `born from the data'.''}

$\bullet$ Page 2, last paragraph:  NeuralNetTools does not ``improve the breadth and quality of information obtained from the MLP neural network.''  Instead, the package provides tools to better understand the MLP neural network.

{\it The sentence was revised: ''...that was developed to better understand information obtained from the MLP neural network.''}

$\bullet$ Page 6:  I am not familiar with the concepts of power and degrees of freedom with neural networks.  Please clarify the sentence, ``Pruning has additional benefits...''

{\it  These sentences were clarified to emphasize that pruning can aid in 1) the interpretation of relationships between variables by removing unimportant connections, and 2) successive model fitting by reducing the number of weights that are changed during model training: ``In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.  This reduces the number of free parameters (weights) that are  estimated by the model optimization routine, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error (i.e., model identifiability, Ellenius and Groth 2000).'' }

$\bullet$ Page 8, Section 3.3:  It is not clear to me how the Lek profile is used to evaluate interactions between variables.

{\it The \texttt{lekprofile} function evaluates the change in the response variable across the range of values for a given explanatory variable while holding the remaining explanatory variables constant.  I have considered these plots conceptually similar to traditional interaction plots (i.e., what is the effect of an explanatory variable at different values of another explanatory variable), but I suppose there is not a 1:1 correspondence. The \texttt{lekprofile} function is more appropriately used to evaluate the shape of the response for a given explanatory variable, as opposed to the importance of variables as for the other functions. I've replaced `interactions' with `relationships' as this description is more appropriate.}

$\bullet$ Figure 6:  I was not able to reproduce this plot.  In addition, when I generated this plot, the bars for two of the groups contain zero (see Groups 1 and 2) - that is, the bar is greater than zero for part of the bar and is less than zero for the other part of the bar.   I thought these bars would be bounded above or below by zero—perhaps I'm not understanding this plot.

{\it This was an issue with stacking of non-positive values in ggplot bar plots.  The situation has been resolved.  Additionally, the default position of the bar plots is now \texttt{`dodge'} (side-by-side for each group), as passed to \texttt{ggplot2::geom\_bar}.  This was implemented in the development branch on the GitHub page and will be pushed in the next release to CRAN.}

$\bullet$ Section 4:  When I ran the code provided in the git repository, I could not reproduce Figure 7 (see attached Figure7.png file).  I added the following line before the toplo1 line to make this work:

\texttt{imp\_sums\$variable}

{\it Thanks, this was fixed.}

\end{document}

### 2016-11-22 nnt_manu.Rnw
revision draft done, need to proof and fix applied example

@@
 \Abstract{
-Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.
-A common criticism of the method is the inability to characterize relationships among variables from a fitted model.
-Although several methods have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.
+Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns among multiple variables.
+A common criticism of these methods is the inability to characterize relationships among variables from a fitted model.
+Although several techniques have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.
 This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.
@@
 
-Data-intensive analysis is a relatively new research approach that focuses on the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.
-A central theme is the focus on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
+A common objective of data-intensive analysis is the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.
+Analysis is primarily focused on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
 Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data.
@@
 
-This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.
+This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to better understand information obtained from the \ac{mlp} neural network.
 Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.
@@
 This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
-Pruning has the additional benefit of increasing power by increasing the degrees of freedom or the identification of input nodes with low importance.
+In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.
+This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}.
 Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
@@
 
-<<eval = F, echo = F, results = hide>>=
+<<eval = T, echo = F, results = hide>>=
 # pruned model using RSNNS
@@
 
-An alternative approach to evaluate variable interactions in a neural network is the Lek profile method \citep{Lek96,Gevrey03}.
+An alternative approach to evaluate variable relationships in a neural network is the Lek profile method \citep{Lek96,Gevrey03}.
 The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
@@
 
-A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).
+A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate 1) use of the functions in \pkg{NeuralNetTools} to gain additional insight into relationships among variables, and 2) effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).
 This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
@@
 Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
-All explanatory variables were normalized and the response variable was standardized to 0--1.
 
-The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.
+First, the apropriate month and airline carrier were selected, all explanatory variables were normalized, and the response variable was standardized to 0--1.
+<<eval = T>>=
+library("nycflights13")
+library("dplyr")
+tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
+  select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
+  mutate_each(funs(scale), -arr_delay) %>% 
+  mutate_each(funs(as.numeric), -arr_delay) %>% 
+  mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
+  data.frame
+@
+Then, a standard \ac{mlp} with five hidden nodes was created with the \pkg{nnet} package to model the effects of selected variables on arrival delays.
+The entire dataset is used for the example but separate training and validation datasets should be used in practice.
+<<eval = F, eval = T, results = hide>>=
+library(nnet)
+mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
+@
+The default output is limited to structural information about the model and methods for model predictions (see \code{str(mod)} and \code{?predict.nnet}).
+Using functions from \pkg{NeuralNetTools}, a more comprehensive understanding of the relationships between the variables is illustrated.
+<<eval = F>>=
+plotnet(mod)
+garson(mod)
+olden(mod)
+lekprofile(mod, group_vals = 5)
+lekprofile(mod, group_vals = 5, group_show = TRUE)
+@
+<<echo = F, results = hide>>=
+library(nnet)
+mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)
+
+pdf('figs/flightplots1.pdf', height = 4, width = 5.5, family = 'serif')
+par(mar = c(0, 0, 0, 0))
+plotnet(mod, cex_val = 0.7, line_stag = 0.02, circle_cex = 4)
+gridExtra::grid.arrange(
+  garson(mod),
+  olden(mod),
+  ncol = 1
+  )
+dev.off()
+
+p1 <- lekprofile(mod, group_vals = 5)
+p2 <- lekprofile(mod, group_vals = 5, group_show = TRUE)
+pA <- ggplot_gtable(ggplot_build(p1))
+pB <- ggplot_gtable(ggplot_build(p2))
+maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])
+pA$widths[2:3] <- maxWidth
+pB$widths[2:3] <- maxWidth
+
+pdf('figs/flightplots2.pdf', height = 4.5, width = 8, family = 'serif')
+gridExtra::grid.arrange(
+  pA, 
+  pB,
+  ncol = 1
+)
+dev.off()
+@
+
+\Cref{fig:flightplots} shows the information about arrival delays that can be obtained with the functions in \pkg{NeuralNetTools}.
+The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used develop a general characterization of the relationships between variables.
+For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays.
+\Cref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more quantitative descriptions using information from both the \ac{nid} and model predictions.
+\Cref{fig:flightplotsb} shows variable importance using the \code{garson} and \code{olden} algorithms.
+The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.
+However, the \code{garson} function shows arrival time as the second most important variable, whereas the \code{olden} function shows this variable as having a weak negative relationship with arrival time.
+This discrepancy is explained below.
+Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.
+Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group four that generally had early arrival times and long departure delays.
+
+%%%%
+\begin{figure}
+\centering
+\subfloat[][\code{plotnet}]{
+\includegraphics[width = 0.5\textwidth, page = 1]{figs/flightplots1.pdf}
+\label{fig:flightplotsa}
+}
+\subfloat[][\code{garson} (top) and \code{olden} (bottom)]{
+\includegraphics[width = 0.5\textwidth, page = 2]{figs/flightplots1.pdf}
+\label{fig:flightplotsb}
+}
+
+\subfloat[][\code{lekprofile}]{
+\includegraphics[width = 1.05\textwidth, page = 1]{figs/flightplots2.pdf}
+\label{fig:flightplotsc}
+}
+
+\caption{Results from a simple \ac{mlp} model of arrival delay for December airline flights versus departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between destinations (\code{air\_time}), and distance flown (\code{distance}).
+The three plots show the \ac{nid} from \code{plotnet} (\ref{fig:flightplotsa}), variable importance with \code{garson} and \code{olden} (\ref{fig:flightplotsb}), and sensitivity analysis with variable groupings from \code{lekprofile} (\ref{fig:flightplotsc}).
+Interpretations are provided in the text.}
+\label{fig:flightplots}
+\end{figure}
+
+A second analysis was needed to show the effects of network architecture and initial starting weights on uncertainty in estimates of variable importance.
 Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.
@@
 
-Several conclusions can be made from \cref{fig:flightimp}.
-First, delays in arrival time are negatively related to distance and positively related to departure delays and air time.
+Several conclusions from \cref{fig:flightimp} provide further information to interpret the trends in \cref{fig:flightplots}.
+First, consistent relationships can be identified such that delays in arrival time were negatively related to distance and positively related to departure delays and air time.
 That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
@@
 Functions can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).
+Although visualizing a neural network with \code{plotnet} is impractical for large models, the remaining functions can simplify model complexity to identify important relationships between variables.
 Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods could be added based on popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).

### 2016-11-22 nnt_manu.tex
revision draft done, need to proof and fix applied example

@@
 Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
-All explanatory variables were normalized and the response variable was standardized to 0--1.
 
-The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.
+First, the apropriate month and airline carrier were selected, all explanatory variables were normalized, and the response variable was standardized to 0--1.
+\begin{Schunk}
+\begin{Sinput}
+R> library("nycflights13")
+R> library("dplyr")
+R> tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
++   select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
++   mutate_each(funs(scale), -arr_delay) %>% 
++   mutate_each(funs(as.numeric), -arr_delay) %>% 
++   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
++   data.frame
+\end{Sinput}
+\end{Schunk}
+Then, a standard \ac{mlp} with five hidden nodes was created with the \pkg{nnet} package to model the effects of selected variables on arrival delays.
+The entire dataset is used for the example but separate training and validation datasets should be used in practice.
+\begin{Schunk}
+\begin{Sinput}
+R> library(nnet)
+R> mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
+\end{Sinput}
+\end{Schunk}
+The default output is limited to structural information about the model and methods for model predictions (see \code{str(mod)} and \code{?predict.nnet}).
+Using functions from \pkg{NeuralNetTools}, a more comprehensive understanding of the relationships between the variables is illustrated.
+\begin{Schunk}
+\begin{Sinput}
+R> plotnet(mod)
+R> garson(mod)
+R> olden(mod)
+R> lekprofile(mod, group_vals = 5)
+R> lekprofile(mod, group_vals = 5, group_show = TRUE)
+\end{Sinput}
+\end{Schunk}
+
+\Cref{fig:flightplots} shows the information about arrival delays that can be obtained with the functions in \pkg{NeuralNetTools}.
+The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used develop a general characterization of the relationships between variables.
+For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays.
+\Cref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more quantitative descriptions using information from both the \ac{nid} and model predictions.
+\Cref{fig:flightplotsb} shows variable importance using the \code{garson} and \code{olden} algorithms.
+The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.
+However, the \code{garson} function shows arrival time as the second most important variable, whereas the \code{olden} function shows this variable as having a weak negative relationship with arrival time.
+This discrepancy is explained below.
+Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.
+Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group four that generally had early arrival times and long departure delays.
+
+%%%%
+\begin{figure}
+\centering
+\subfloat[][\code{plotnet}]{
+\includegraphics[width = 0.5\textwidth, page = 1]{figs/flightplots1.pdf}
+\label{fig:flightplotsa}
+}
+\subfloat[][\code{garson} (top) and \code{olden} (bottom)]{
+\includegraphics[width = 0.5\textwidth, page = 2]{figs/flightplots1.pdf}
+\label{fig:flightplotsb}
+}
+
+\subfloat[][\code{lekprofile}]{
+\includegraphics[width = 1.05\textwidth, page = 1]{figs/flightplots2.pdf}
+\label{fig:flightplotsc}
+}
+
+\caption{Results from a simple \ac{mlp} model of arrival delay for December airline flights versus departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between destinations (\code{air\_time}), and distance flown (\code{distance}).
+The three plots show the \ac{nid} from \code{plotnet} (\ref{fig:flightplotsa}), variable importance with \code{garson} and \code{olden} (\ref{fig:flightplotsb}), and sensitivity analysis with variable groupings from \code{lekprofile} (\ref{fig:flightplotsc}).
+Interpretations are provided in the text.}
+\label{fig:flightplots}
+\end{figure}
+
+A second analysis was needed to show the effects of network architecture and initial starting weights on uncertainty in estimates of variable importance.
 Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.

### 2016-11-22 flightimp.R

@@ -1,9 +1,10 @@
-library(nnet)
-library(NeuralNetTools)
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(nycflights13)
+
+library("nnet")
+library("NeuralNetTools")
+library("dplyr")
+library("tidyr")
+library("ggplot2")
+library("nycflights13")
 
 # data preprocessing
 # select UA carrier for month of december and relevant variables
@@ -14,6 +15,19 @@ tomod <- filter(flights, month == 12 & carrier == 'UA') %>%
   mutate_each(funs(as.numeric), -arr_delay) %>% 
   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
   data.frame
+
+# example model
+mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
+
+# plots of the example model
+plotnet(mod)
+garson(mod)
+olden(mod)
+lekprofile(mod, group_vals = 5)
+lekprofile(mod, group_vals = 5, group_show = TRUE)
+
+##
+# test the effects of model structure and starting conditions on variable importance
 
 # setup initial conditions
 nodes <- c(1, 5, 10)
@@ -63,15 +77,15 @@ imp_sums <- data.frame(grids, imp_sums) %>%
 
 # subet by nodes so barplots can be ordered by medians
 toplo1 <- filter(imp_sums, nodes == 1) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 toplo2 <- filter(imp_sums, nodes == 5) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 toplo3 <- filter(imp_sums, nodes == 10) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 
 # widths
 wd <- 0.3
-ylims <- c(-3.93, 3.49)
+ylims <- with(imp_sums, c(min(lo), max(hi)))
 
 p1 <- ggplot(toplo1, aes(x = variable, y = med, fill = med)) + 
   geom_bar(stat = 'identity') + 
[…]

### 2016-11-22 nnt_manu.R

@@ -58,6 +58,19 @@ tomod <- filter(flights, month == 12 & carrier == 'UA') %>%
   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
   data.frame
 
+# example model
+mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
+
+# plots of the example model
+plotnet(mod)
+garson(mod)
+olden(mod)
+lekprofile(mod, group_vals = 5)
+lekprofile(mod, group_vals = 5, group_show = TRUE)
+
+##
+# test the effects of model structure and starting conditions on variable importance
+
 # setup initial conditions
 nodes <- c(1, 5, 10)
 seeds <- sample(c(1:5000), 100)
@@ -106,15 +119,15 @@ imp_sums <- data.frame(grids, imp_sums) %>%
 
 # subet by nodes so barplots can be ordered by medians
 toplo1 <- filter(imp_sums, nodes == 1) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 toplo2 <- filter(imp_sums, nodes == 5) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 toplo3 <- filter(imp_sums, nodes == 10) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 
 # widths
 wd <- 0.3
-ylims <- c(-3.93, 3.49)
+ylims <- with(imp_sums, c(min(lo), max(hi)))
 
 p1 <- ggplot(toplo1, aes(x = variable, y = med, fill = med)) + 
   geom_bar(stat = 'identity') + 

### 2016-11-28 review_response.Rnw
revision resubmitted to JSS

@@
 
-{\it I thank the reviewer for providing thoughtful comments on my manuscript and the NeuralNetTools package.
-My response to these comments are shown in italics.
-Page and paragraph numbers refer to the original manuscript.}
+{\it I thank the reviewer for providing thoughtful comments on my manuscript and the NeuralNetTools package.  My response to these comments are shown in italics.}
 
@@
 I have added some text to the conclusions to re-emphasize this issue.
-First paraggraph of sec.
+First paragraph of sec.
 5: ``Although visualizing a neural network with \texttt{plotnet} is impractical for large models, the remaining functions can simplify model complexity to identify important relationships between variables.''
@@
 ``Figure 7 shows the information about arrival delays that can be obtained with the functions in \textbf{NeuralNetTools}.
-The NID (7a) shows the model structure and can be used develop a general characterization of the relationships between variables.
+The NID (7a) shows the model structure and can be used as a general characterization of the relationships between variables.
 For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays.
-Figures 7b and 7c provide more quantitative descriptions using information from both the NID and model predictions.
+Similarly, a large positive weight is observed between I3 and H1, suggesting arrival time is positively associated with arrival delays.
+However, interpreting individual connection weights between layers is challenging.
+Figures 7b amd 7c provide more quantitative descriptions using information from both the NID and model predictions.
 Figure 7b shows variable importance using the \texttt{garson} and \texttt{olden} algorithms.
 The \texttt{garson} function suggests time between destinations (\texttt{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \texttt{olden} method.
-However, the \texttt{garson} function shows arrival time as the second most important variable, whereas the \texttt{olden} function shows this variable as having a weak negative relationship with arrival time.
-This discrepancy is explained below.
+However, the \texttt{garson} function shows arrival time (\texttt{arr\_time}) as the third most important variable for arrival delays, whereas this is shown as the strongest positive association by the \texttt{olden} function.
+Similar discrepancies between the two methods are observed for other variables, which are explained below.
 Finally, results from the \texttt{lekprofile} function (7c) confirm those in 7b, with the addition of non-linear responses that vary by different groupings of the data.
-Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group five that generally had early arrival times and long departure delays.''
+Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group four that generally had early arrival times and long departure delays.''
 }
@@
 {\it  These sentences were clarified to emphasize that pruning can aid in 1) the interpretation of relationships between variables by removing unimportant connections, and 2) successive model fitting by reducing the number of weights that are changed during model training: ``In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.
-This reduces the number of free parameters (weights) that are  estimated by the model optimization routine, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error (i.e., model identifiability, Ellenius and Groth 2000).'' }
+This reduces the number of free parameters (weights) that are  estimated by the model optimization routine, increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error (i.e., model identifiability, Ellenius and Groth 2000).'' }
 
@@
 Additionally, the default position of the bar plots is now \texttt{`dodge'} (side-by-side for each group), as passed to \texttt{ggplot2::geom\_bar}.
-This was implemented in the development branch on the GitHub page and will be pushed in the next release to CRAN.}
+This has been changed in the latest version on CRAN.}
 

### 2016-11-28 review_response.tex
revision draft done, need to proof and fix applied example
revision resubmitted to JSS

--- 2016-11-22 review_response.Rnw
@@
 %knitr options
-<<setup, cache = F, echo = F>>=
-library(knitr)
-# set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
-options(replace.assign=TRUE,width=90,digits=1)
-@
 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{review_response-concordance}
 \raggedright

### 2016-11-28 cover_letter.tex
package license to cover letter
revision resubmitted to JSS

@@
 \pagestyle{empty} 
+\small
 
@@
 
+\vfill
 \emph{To:}\newline
@@
 
-Enclosed please find my manuscript, entitled `NeuralNetTools: Visualization and Analysis Tools for Neural Networks', to be considered as a research article in the Journal of Statistical Software.
+Enclosed please find my revised manuscript, entitled `NeuralNetTools: Visualization and Analysis Tools for Neural Networks'.  This revision follows a minor acceptence as a research article in JSS.
 
-Supervised neural networks are statistical models that have been used to support prediction and exploratory analysis in data-rich environments.
-A common criticism of these methods is that they represent `black-boxes' that do not describe any information about relationships among variables.
-Methods to quantify information from a fitted neural network have been previously described in the literature but have not been available in a statistical computing environment.
-This article describe the NeuralNetTools package for R that was developed to better quantify information provided by a neural network, including methods for plotting using a neural network interpretation diagram, estimates of variable importance, and a sensitivity analysis.
-The package is exhaustive in scope by including methods for the most commonly used neural network objects in R.
+I appreciate the insightful comments provided by the reviewer and I have made every effort to address the concerns.
+In particular, I have improved the applied example by demonstrating the interpretation of a simple neural network model using functions from the package.
+This includes addition to the text and Figure 7.
+I have also fixed any remaining issues with reproducibility in the manuscript code and supplementary material.
+Please see all of my responses to the reviewer comments included with my revision.
 
-Accompanying this submission is the manuscript PDF created with the JSS \LaTeX\ template, an R script of the manuscript code, and links to the package source code at the bottom of this letter.
-Please note that the current version of the package on CRAN was submitted this year on Dec.
-1 for minor revisions related to manuscript content.
-The original package was first uploaded to CRAN on December 16\textsuperscript{th}, 2014.
-Accordingly, the package has been used by the R community for a sufficient amount of time to address bugs or issues (see \href{https://github.com/fawda123/NeuralNetTools/issues?q=is%3Aissue+is%3Aclosed}{here}) in the early releases.
+This revision includes a manuscript PDF created with the JSS \LaTeX\ template (nnt\_manu.pdf), my response to the reviewer comments (review\_response.pdf), an R script of the manuscript code (nnt\_manu.R), a .tex file for the manuscript text (nnt\_manu.tex), a .bbl file with bibliographic information (refs.bbl), all pdf files used for figures (in fig folder), and supporting JSS \LaTeX\ template files.
+Links to the package source code are at the bottom of this letter.
+The package is licensed uncer CC0.
 
@@
 
+\vfill
 \end{document}

### 2016-11-28 nnt_manu.Rnw
revision resubmitted to JSS

@@
 In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.
-This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}.
+This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}.
 Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
@@
 \Cref{fig:flightplots} shows the information about arrival delays that can be obtained with the functions in \pkg{NeuralNetTools}.
-The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used develop a general characterization of the relationships between variables.
+The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used as a general characterization of the relationships between variables.
 For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays.
+Similarly, a large positive weight is observed between I3 and H1, suggesting arrival time is positively associated with arrival delays.
+However, interpreting individual connection weights between layers is challenging.
 \Cref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more quantitative descriptions using information from both the \ac{nid} and model predictions.
@@
 The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.
-However, the \code{garson} function shows arrival time as the second most important variable, whereas the \code{olden} function shows this variable as having a weak negative relationship with arrival time.
-This discrepancy is explained below.
+However, the \code{garson} function shows arrival time (\code{arr\_time}) as the third most important variable for arrival delays, whereas this is shown as the strongest positive association by the \code{olden} function.
+Similar discrepancies between the two methods are observed for other variables, which are explained below.
 Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.

### 2016-11-28 resubmit/nnt_manu.tex
revision resubmitted to JSS

--- 2016-11-22 nnt_manu.tex
@@
 The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.
-However, the \code{garson} function shows arrival time as the second most important variable, whereas the \code{olden} function shows this variable as having a weak negative relationship with arrival time.
-This discrepancy is explained below.
+However, the \code{garson} function shows arrival time (\code{arr\_time}) as the third most important variable for arrival delays, whereas this is ranked highest by the \code{olden} function.
+Similar discrepancies between the two methods are observed for other variables, which are explained below.
 Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.
@@
 % note that ref titles need to be in title-case 
-% \bibliographystyle{jss}
-% \bibliography{refs}
-% this file is bbl edited by hand after running bibtex
-\input{CopyOfnnt_manu.bbl}
+\input{refs.bbl}
 

### 2016-11-28 resubmit/nnt_manu.R

@@ -1,0 +1,155 @@
+## install.packages("NeuralNetTools")
+library("NeuralNetTools")
+
+set.seed(123)
+library("RSNNS")
+x <- neuraldat[, c('X1', 'X2', 'X3')]
+y <- neuraldat[, 'Y1']
+mod1 <- mlp(x, y, size = 5)
+library("neuralnet")
+mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
+library("nnet")
+mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
+
+par(mar = c(0, 0, 0, 0))
+plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
+plotnet(mod3)
+
+modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
+
+par(mar = c(0, 0, 0, 0))
+plotnet(modskip, skip = TRUE)
+plotnet(modskip)
+
+pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01,
+  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
+mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+ pruneFuncParams = pruneFuncParams)
+plotnet(mod, rel_rsc = c(3, 8))
+plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
+
+garson(mod1)
+olden(mod1)
+garson(mod2)
+olden(mod2)
+garson(mod3)
+olden(mod3)
+
+lekprofile(mod3)
+lekprofile(mod3, group_show = TRUE)
+lekprofile(mod3, group_vals = 6)
+lekprofile(mod3, group_vals = 6, group_show = TRUE)
+
+library("nnet")
+library("NeuralNetTools")
+library("dplyr")
+library("tidyr")
+library("ggplot2")
+library("nycflights13")
+
+# data preprocessing
+# select UA carrier for month of december and relevant variables
+# normalize and standardize variables
+tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
+  select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
[…]

### 2018-05-03 final_JSS_edits/paper.tex
working on final JSS edits

--- 2016-11-28 resubmit/nnt_manu.tex
@@
-\documentclass[article,shortnames]{jss}
+\documentclass[article]{jss}
+\usepackage{thumbpdf} 
+\graphicspath{{Figures/}}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{Marcus W. Beck\\US Environmental Protection Agency}
 
-%% almost as usual
-\author{Marcus W. Beck\\US Environmental Protection Agency}
 \title{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Marcus W. Beck} %% comma-separated
-\Plaintitle{NeuralNetTools: Visualization and Analysis Tools for Neural Networks} %% without formatting
+\Plainauthor{Marcus W. Beck} 
+\Plaintitle{NeuralNetTools: Visualization and Analysis Tools for Neural Networks}
+
 \Shorttitle{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}
 
-%% an abstract and keywords
-\Abstract{
-Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns among multiple variables.
-A common criticism of these methods is the inability to characterize relationships among variables from a fitted model.
-Although several techniques have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.
-This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.
-Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.
-Methods are provided for objects from many of the common neural network packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.
-The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.
-Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.
-}
-\Keywords{neural networks, \code{plotnet}, sensitivity, variable importance, \proglang{R}}
-\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
-%% at least one keyword must be supplied
+\Abstract{ Supervised neural networks have been applied as a machine
+  learning technique to identify and predict emergent patterns among
+  multiple variables.  A common criticism of these methods is the
+  inability to characterize relationships among variables from a
+  fitted model. Although several techniques have been proposed to
+  ``illuminate the black box'', they have not been made available in an
+  open-source programming environment.  This article describes the
+  \pkg{NeuralNetTools} package that can be used for the interpretation
+  of supervised neural network models created in \proglang{R}.
+  Functions in the package can be used to visualize a model using a
+  neural network interpretation diagram, evaluate variable importance
+  by disaggregating the model weights, and perform a sensitivity
+  analysis of the response variables to changes in the input
+  variables.  Methods are provided for objects from many of the common
+  neural network packages in \proglang{R}, including \pkg{caret},
+  \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides
+  a brief overview of the theoretical foundation of neural networks, a
+  description of the package structure and functions, and an applied
+  example to provide a context for model development with
+  \pkg{NeuralNetTools}.  Overall, the package provides a toolset for
+  neural networks that complements existing quantitative techniques
+  for data-intensive exploration.  }
 
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
+\Keywords{neural networks, \code{plotnet}, sensitivity, variable
+  importance, \proglang{R}}
+
+\Plainkeywords{neural networks, plotnet, sensitivity, variable
+  importance, R}
+
 %% \Volume{50}
@@
 
-%% The address of (at least) one author should be given
-%% in the following format:
 \Address{
@@
 
-% options and libs
-% get online bib file
-
-%% need no \usepackage{Sweave.sty}
-
 \begin{document}
-
-%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
 
@@
 
-A common objective of data-intensive analysis is the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.
-Analysis is primarily focused on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.
-Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data.
-Fields of research characterized by high-throughput data (e.g., bioinformatics, \citet{Saeys07}) have a strong foundation in computationally-intensive methods of analysis, whereas disciplines that have historically been limited by data quantity (e.g., field ecology, \citet{Swanson15}) have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information.
-Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.
+A common objective of data-intensive analysis is the synthesis of
+unstructured information to identify patterns or trends ``born from the
+data'' \citep{Bell09,Kelling09,Michener12}.  Analysis is primarily
+focused on data exploration and prediction as compared to
+hypothesis-testing using domain-specific methods for scientific
+exploration \citep{Kell03}.  Demand for quantitative toolsets to
+address challenges in data-rich environments has increased drastically
+with the advancement of techniques for rapid acquisition of
+data. Fields of research characterized by high-throughput data (e.g.,
+bioinformatics; \citealt{Saeys07}) have a strong foundation in
+computationally-intensive methods of analysis, whereas disciplines
+that have historically been limited by data quantity (e.g., field
+ecology; \citealt{Swanson15}) have also realized the importance of
+quantitative toolsets given the development of novel techniques to
+acquire information.  Quantitative methods that facilitate inductive
+reasoning can serve a complementary role to conventional,
+hypothesis-driven approaches to scientific discovery \citep{Kell03}.
 
-Statistical methods that have been used to support data exploration are numerous \citep{Jain00,Recknagel06,Zuur10}.
-A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.
-Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.
-Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in emerging fields focused on data exploration.
-The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.
-This model is typically used to predict the response of one or more variables given one to many explanatory variables.
-The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropagation algorithm.
-Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.
+Statistical methods that have been used to support data exploration
+are numerous \citep{Jain00,Recknagel06,Zuur10}.  A common theme among
+data intensive methods is the use of machine-learning algorithms where
+the primary objective is to identify emergent patterns with minimal
+human intervention.  Neural networks, in particular, are designed to
+mimic the neuronal structure of the human brain by ``learning'' inherent
+data structures through adaptive algorithms
+\citep{Rumelhart86,Ripley96}.  Although the conceptual model was
+introduced several decades ago \citep{McCulloch43}, neural networks
+have had a central role in emerging fields focused on data
+exploration.  The most popular form of neural network is the
+feed-forward \ac{mlp} trained using the backpropagation algorithm
+\citep{Rumelhart86}.  This model is typically used to predict the
+response of one or more variables given one to many explanatory
+variables.  The hallmark feature of the \ac{mlp} is the
+characterization of relationships using an arbitrary number of
+parameters (i.e., the hidden layer) that are chosen through iterative
+training with the backpropagation algorithm.  Conceptually, the
+\ac{mlp} is a hyper-parameterized non-linear model that can fit a
+smooth function to any dataset with minimal residual error
+\citep{Hornik91}.
 
-An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but complicates the extraction of model information.
-Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network.
-As such, a common criticism is that neural networks are `black boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.
-\citet{Olden02} provide a rebuttal to this concern by describing methods to extract information about variable relationships from neural networks.
-Many of these methods were previously described but not commonly used.
-For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.
-These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.
-Although \ac{mlp} neural networks were developed for prediction, methods described in \citet{Olden02} leverage these models to describe data signals.
-Increasing the accesibility of these diagnostic tools will have value for exploratory data analysis and may also inform causal inference.
+An arbitrarily large number of parameters to fit a neural network
+provides obvious predictive advantages, but complicates the extraction
+of model information.  Diagnostic information such as variable
+importance or model sensitivity are necessary aspects of exploratory
+data analysis that are not easily obtained from a neural network. As
+such, a common criticism is that neural networks are ``black boxes''
+that offer minimal insight into relationships among variables
+\citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to
+this concern by describing methods to extract information about
+variable relationships from neural networks.  Many of these methods
+were previously described but not commonly used.  For example,
+\citet{Olden02} describe the \ac{nid} for plotting \citep{Ozesmi99},
+the Garson algorithm for variable importance \citep{Garson91}, and the
+profile method for sensitivity analysis \citep{Lek96}.  These
+quantitative tools ``illuminate the black box'' by disaggregating the
+network parameters to characterize relationships between variables
+that are described by the model.  Although \ac{mlp} neural networks
+were developed for prediction, methods described in \citet{Olden02}
+leverage these models to describe data signals.  Increasing the
+accessibility of these diagnostic tools will have value for exploratory
+data analysis and may also inform causal inference.
 
-This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to better understand information obtained from the \ac{mlp} neural network.
-Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.
-The reach of the package is extensive in that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.
-The objectives of this article are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.
-The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.
+This article describes the \pkg{NeuralNetTools} package
+\citep{NeuralNetTools} for \proglang{R} \citep{R} that was developed
+to better understand information obtained from the \ac{mlp} neural
+network.  Functions provided by the package are those described in
+\citet{Olden02} but have not been previously available in an
+open-source programming environment.  The reach of the package is
+extensive in that generic functions were developed for model objects
+from the most popular neural network packages available in
+\proglang{R}.  The objectives of this article are to 1) provide an
+overview of the statistical foundation of the \ac{mlp} network, 2)
+describe the theory and application of the main functions in the
+\pkg{NeuralNetTools} package, and 3) provide an applied example using
+neural networks and \pkg{NeuralNetTools} in data exploration.  The
+current released package version is available from the \ac{CRAN}
+(CRAN) at \url{https://CRAN.R-project.org/package=NeuralNetTools},
+whereas the development version is maintained as a GitHub repository.
 
-\section[Theoretical foundation]{Theoretical foundation and existing \proglang{R} packages}
+\section[Theoretical foundation and existing R packages]{Theoretical foundation and existing \proglang{R} packages}
 
-The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.
-Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables.
-Hidden layers between the input and response layers are key components of a neural network that mediate the transfer of information.
-Just as the input and response layers are composed of variables or nodes, each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.
-Bias layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.
+The typical \ac{mlp} network is composed of multiple layers that
+define the transfer of information between input and response layers.
+Information travels in one direction where a set of values for
+variables in the input layer propagates through one or more hidden
+layers to the final layer of the response variables. Hidden layers
+between the input and response layers are key components of a neural
+network that mediate the transfer of information.  Just as the input
+and response layers are composed of variables or nodes, each hidden
+layer is composed of nodes with weighted connections that define the
+strength of information flow between layers.  Bias layers connected to
+hidden and response layers may also be used that are analogous to
+intercept terms in a standard regression model.
 
-Training a neural network model requires identifying the optimal weights that define the connections between the model layers.
-The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.
-Training is commonly achieved using the backpropagation algorithm described in \citet{Rumelhart86}.
-This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.
-The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.
-The weights are then changed through a backpropagation step that begins by changing weights in the output layer and then the remaining hidden layers.
-The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.
-A fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
+Training a neural network model requires identifying the optimal
+weights that define the connections between the model layers.  The
+optimal weights are those that minimize prediction error for a test
+dataset that is independent of the training dataset.  Training is
+commonly achieved using the backpropagation algorithm described in
+\citet{Rumelhart86}.  This algorithm identifies the optimal weighting
+scheme through an iterative process where weights are gradually
+changed through a forward- and backward-propagation process
+\citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an
+arbitrary weighting scheme to the connections in the network, followed
+by estimating the output in the response variable through the
+forward-propagation of information through the network, and finally
+calculating the difference between the predicted and actual value of
+the response.  The weights are then changed through a backpropagation
+step that begins by changing weights in the output layer and then the
+remaining hidden layers.  The process is repeated until the chosen
+error function is minimized, as in standard model-fitting techniques
+for regression \citep{Cheng94}.  A fitted \ac{mlp} neural network can
+be represented as \citep{Bishop95,Venables02}:
 \begin{equation}
-y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
+y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right),
 \end{equation}
-where the estimated value $k$ of the response variable $y$ is a sum of products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.
+where the estimated value of the response variable $y_k$ is a sum of
+products between the respective weights $w$ for $i$ input variables
+$x$ and $h$ hidden nodes, mediated by the activation functions $f_h$
+and $f_o$ for each hidden and output node.
 
-Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}.
-Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.
-Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 95\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
-As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.
-Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class).
-Additional \code{default} or \code{numeric} methods are available for some of the generic functions.
+Methods in \pkg{NeuralNetTools} were written for several \proglang{R}
+packages that can be used to create \ac{mlp} neural networks:
+\pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and
+\pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed
+for neural network objects created with the \code{train} function from
+the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R}
+packages that can create \ac{mlp} neural networks include \pkg{AMORE}
+that implements the ``TAO-robust backpropagation algorithm'' for model
+fitting \citep{Castejon14}, \pkg{FCNN4R} that provides an \proglang{R}
+interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15},
+\pkg{monmlp} for networks with partial monotonicity constraints
+\citep{Cannon15}, and \pkg{qrnn} for quantile regression neural
+networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN}
+download logs \citep{Csardi15} showed that the \proglang{R} packages
+with methods in \pkg{NeuralNetTools} included 95\% of all downloads
+for the available \ac{mlp} packages, with \pkg{nnet} accounting for
+over 78\%.  As such, methods have not been included in
+\pkg{NeuralNetTools} for the remaining packages, although further
+development of \pkg{NeuralNetTools} could include additional methods
+based on popularity.  Methods for each function are currently
+available for `\code{mlp}' (\pkg{RSNNS}), `\code{nn}'
+(\pkg{neuralnet}), `\code{nnet}' (\pkg{nnet}), and `\code{train}'
+(\pkg{caret}; only if the object also inherits from the `\code{nnet}'
+class) objects.  Additional default methods or methods for class
+`\code{numeric}' are available for some of the generic functions.
 
@@
 The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
+%
+\pkg{NeuralNetTools} includes four main functions that were developed
+following similar techniques in \citet{Olden02} and references
+therein.  The functions include \code{plotnet} to plot a neural
+network interpretation diagram, \code{garson} and \code{olden} to
+evaluate variable importance, and \code{lekprofile} for a sensitivity
+analysis of neural network response to input variables.  Most of the
+functions require the extraction of model weights in a common format
+for the neural network object classes in \proglang{R}.  The
+\code{neuralweights} function can be used to retrieve model weights
+for any of the model classes described above.  A two-element
+\code{list} is returned with the first element describing the
+structure of the network (number of nodes in the input, hidden, and
+output layers) and the second element as a named list of model
+weights.  The function is used internally within the main functions
+but may also be useful for comparing networks of different classes.
 
-\pkg{NeuralNetTools} includes four main functions that were developed following similar techniques in \citet{Olden02} and references therein.
-Functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.
-Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.
-The \code{neuralweights} function can be used to retrieve model weights for any of the model classes described above.
-A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.
-The function is used internally within the main functions but may be useful for comparing networks of different classes.
-
-A common approach for data pre-processing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.
-A sample dataset that follows this format is included with \pkg{NeuralNetTools}.
-The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and five columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).
-The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.
-The response variables are also standardized from zero to one.
-Additional datasets can be pre-processed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.
-The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.
-
+A common approach for data pre-processing is to normalize the input
+variables and to standardize the response variables
+\citep{Lek00,Olden02}.  A sample dataset that follows this format is
+included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a
+simple \code{data.frame} with 2000 rows of observations and five
+columns for two response variables (\code{Y1} and \code{Y2}) and three
+input variables (\code{X1}, \code{X2}, and \code{X3}).  The input
+variables are random observations from a standard normal distribution
+and the response variables are linear combinations of the input
+variables with additional random components.  The response variables
+are also scaled from zero to one.  Variables in additional datasets can be
+pre-processed to this common format using the \code{scale} function
+from \pkg{base} to center and scale input variables (i.e., z-scores) 
+and the \code{rescale} function from \pkg{scales} to scale response variables 
+from zero to one.  The examples below use three models created from the 
+\code{neuraldat} dataset and include `\code{mlp}' (\pkg{RSNNS}), `\code{nn}' 
+(\pkg{RSNNS}), and `\code{nnet}' (\pkg{nnet}) objects.
+%
 \begin{Schunk}
@@
 R> library("RSNNS")
-R> x <- neuraldat[, c('X1', 'X2', 'X3')]
-R> y <- neuraldat[, 'Y1']
+R> x <- neuraldat[, c("X1", "X2", "X3")]
+R> y <- neuraldat[, "Y1"]
 R> mod1 <- mlp(x, y, size = 5)
@@
 
-Existing plot functions in \proglang{R} to view neural networks are minimal.
-Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance.
-To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in \proglang{R}.
-Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.
+Existing plot functions in \proglang{R} to view neural networks are
+minimal.  Such tools have practical use for visualizing network
+architecture and connections between layers that mediate variable
+importance. To our knowledge, only the \pkg{neuralnet} and
+\pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks
+in \proglang{R}.  Although useful for viewing the basic structure, the
+output is minimal and does not include extensive options for
+customization.
 
-Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid} \citep[\acs{nid},][]{Ozesmi99} and includes several options to customize aesthetics.
-A \ac{nid} is a modification of the standard conceptual illustration of the \ac{mlp} network that changes the thickness and color of the weight connections based on magnitude and sign, respectively.
-Positive weights between layers are shown as black lines and negative weights as grey lines.
-Line thickness is proportional to the absolute magnitude of each weight (\cref{fig:plotnet}).
+The \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid}
+\citep[\acs{nid};][]{Ozesmi99} and includes several options to
+customize aesthetics. A \ac{nid} is a modification of the standard
+conceptual illustration of the \ac{mlp} network that changes the
+thickness and color of the weight connections based on magnitude and
+sign, respectively.  Positive weights between layers are shown as
+black lines and negative weights as gray lines. Line thickness is
+proportional to the absolute magnitude of each weight
+(\cref{fig:plotnet}).
 
-\begin{figure}[!ht]
+\begin{figure}[t!]
 \subfloat[][]{
-\includegraphics[width=7.5cm,page=1]{figs/plotnet.pdf}
+\includegraphics[width=7.5cm,page=1, trim = 0 20 0 15, clip]{plotnet.pdf}
 \label{fig:plotnet1}
@@
 \subfloat[][]{
-\includegraphics[width=7.5cm, page=2]{figs/plotnet.pdf}
+\includegraphics[width=7.5cm, page=2, trim = 0 20 0 15, clip]{plotnet.pdf}
 \label{fig:plotnet2}
@@
 
-A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection (\cref{fig:plotnet_skip}).
-Models with skip layers include additional connections from the input to output layers that bypass the hidden layer \citep{Ripley96}.
-The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}.
-If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights.
-Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.
+A primary and skip layer network can also be plotted for `\code{nnet}'
+models with a skip layer connection (\cref{fig:plotnet_skip}). Models
+with skip layers include additional connections from the input to
+output layers that bypass the hidden layer \citep{Ripley96}.  The
+default behavior of \code{plotnet} is to plot the primary network,
+whereas the skip layer can be viewed separately with \code{skip =
+  TRUE}. If \code{nid = TRUE}, the line widths for both the primary
+and skip layer plots are relative to all weights. Plotting a network
+with only a skip layer (i.e., no hidden layer, \code{size = 0} in
+\code{nnet}) will include bias connections to the output layer,
+whereas these are included only in the primary plot if size is greater
+than zero.
 
-\begin{figure}[!ht]
+\begin{figure}[t!]
 \subfloat[][]{
-\includegraphics[width=7.5cm,page=1]{figs/plotnet_skip.pdf}
+\includegraphics[width=7.5cm,page=1, trim = 0 20 0 15, clip]{plotnet_skip.pdf}
 \label{fig:plotnet_skip1}
@@
 \subfloat[][]{
-\includegraphics[width=7.5cm, page=2]{figs/plotnet_skip.pdf}
+\includegraphics[width=7.5cm, page=2, trim = 0 20 0 15, clip]{plotnet_skip.pdf}
 \label{fig:plotnet_skip2}
@@
 
-The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.
-This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.
-In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.
-This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}.
-Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.
-The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).
-
+The \pkg{RSNNS} package provides algorithms to prune connections or
+nodes in a neural network \citep{Bergmeir12}.  This approach can
+remove connection weights between layers or input nodes that do not
+contribute to the predictive performance of the network.  In addition
+to visualizing connections in the network that are not important,
+connections that are pruned can be removed in successive model
+fitting.  This reduces the number of free parameters (weights) that
+are estimated by the model optimization algorithm, increasing the
+likelihood of convergence to an estimable numeric solution for the
+remaining connection weights that minimizes prediction error
+\citep[i.e., model identifiability;][]{Ellenius00}. Algorithms in
+\pkg{RSNNS} for weight pruning include magnitude-based pruning,
+optimal brain damage, and optimal brain surgeon, whereas algorithms
+for node pruning include skeletonization and the non-contributing
+units method \citep{Zell98}.  The \code{plotnet} function can plot a
+pruned neural network, with options to omit or display the pruned
+connections (\cref{fig:plotprune}).
+%
 \begin{Schunk}
 \begin{Sinput}
-R> pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
-+   no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
-+   init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
+R> pruneFuncParams <- list(max_pr_error_increase = 10.0,
++    pr_accepted_error = 1.0, no_of_pr_retrain_cycles = 1000, 
++    min_error_to_stop = 0.01, init_matrix_value = 1e-6,  
++    input_pruning = TRUE, hidden_pruning = TRUE)
 R> mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
@@
 R> plotnet(mod, rel_rsc = c(3, 8))
-R> plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
+R> plotnet(mod, prune_col = "lightblue", rel_rsc = c(3, 8))
 \end{Sinput}
 \end{Schunk}
-\begin{figure}[!ht]
+%
+\begin{figure}[t!]
 \subfloat[][]{
-\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
+\includegraphics[width=0.48\textwidth, page=1, trim = 0 20 0 65, clip]{plotprune.pdf}
 \label{fig:plotprune1}
@@
 \subfloat[][]{
-\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
+\includegraphics[width=0.48\textwidth, page=2, trim = 0 20 0 65, clip]{plotprune.pdf}
 \label{fig:plotprune2}
 }
-\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.
-The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
+\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12}
+  using the ``optimal brain surgeon'' algorithm described in
+  \citet{Zell98}.  The default plotting behavior of \code{plotnet} is
+  to omit pruned connections (\ref{fig:plotprune1}), whereas they can
+  be viewed as dashed lines by including the \code{prune\_col}
+  argument (\ref{fig:plotprune2}).}
 \label{fig:plotprune}
@@
 Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.
-Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's conection weights algorithm \citep{Olden04}.
+Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's connection weights algorithm \citep{Olden04}.
 
@@
 An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign.
-For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect, whereas \code{garson} may provide different results based on the absolute magnitude.
+For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a canceling effect, whereas \code{garson} may provide different results based on the absolute magnitude.
 An additional advantage is that the \code{olden} function can evaluate neural networks with multiple hidden layers and response variables.
-The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.
+The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0 to 1.
 
-Both functions have similar implementations and require only a model object as input.
-The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).
-The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.
-The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.
-Variable importance shown in \Cref{fig:plotimp} was estimated for each model:
+Both functions have similar implementations and require only a model
+object as input.  The default output is a \pkg{ggplot2} bar plot
+\citep[i.e., \code{geom\_bar};][]{Wickham09} that shows the relative
+importance of each input variable in the model (\cref{fig:plotimp}).
+The plot aesthetics are based on internal code that can be changed
+using conventional syntax for \pkg{ggplot2} applied to the output
+object.  The importance values can also be returned as a
+\code{data.frame} if \code{bar\_plot = FALSE}.  Variable importance
+shown in \Cref{fig:plotimp} is estimated for each model using:
+%
 \begin{Schunk}
@@
 \end{Schunk}
-\begin{figure}[!h]
+%
+\begin{figure}[t!]
 \centering
-\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
-\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
+\subfloat[][Model 1 (`\code{mlp}'), \code{garson}]{
+\includegraphics[width = 0.47\textwidth, page = 1]{plotimp.pdf}
 \label{fig:plotimp1}
 }
-\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
-\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
+\subfloat[][Model 1 (`\code{mlp}'), \code{olden}]{
+\includegraphics[width = 0.47\textwidth, page = 2]{plotimp.pdf}
 \label{fig:plotimp2}
@@
 
-\subfloat[][Model 2 (\code{nn}), \code{garson}]{
-\includegraphics[width = 0.47\textwidth, page = 3]{figs/plotimp.pdf}
+\subfloat[][Model 2 (`\code{nn}'), \code{garson}]{
+\includegraphics[width = 0.47\textwidth, page = 3]{plotimp.pdf}
 \label{fig:plotimp3}
 }
-\subfloat[][Model 2 (\code{nn}), \code{olden}]{
-\includegraphics[width = 0.47\textwidth, page = 4]{figs/plotimp.pdf}
+\subfloat[][Model 2 (`\code{nn}'), \code{olden}]{
+\includegraphics[width = 0.47\textwidth, page = 4]{plotimp.pdf}
 \label{fig:plotimp4}
@@
 
-\subfloat[][Model 3 (\code{nnet}), \code{garson}]{
-\includegraphics[width = 0.47\textwidth, page = 5]{figs/plotimp.pdf}
+\subfloat[][Model 3 (`\code{nnet}'), \code{garson}]{
+\includegraphics[width = 0.47\textwidth, page = 5]{plotimp.pdf}
 \label{fig:plotimp5}
 }
-\subfloat[][Model 3 (\code{nnet}), \code{olden}]{
-\includegraphics[width = 0.47\textwidth, page = 6]{figs/plotimp.pdf}
+\subfloat[][Model 3 (`\code{nnet}'), \code{olden}]{
+\includegraphics[width = 0.47\textwidth, page = 6]{plotimp.pdf}
 \label{fig:plotimp6}
 }
-\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \cref{fig:plotimp1,fig:plotimp3,fig:plotimp5}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \cref{fig:plotimp2,fig:plotimp4,fig:plotimp6}, \citealt{Olden04}).
-Garson's algorithm shows importance as absolute values from 0--1, whereas Olden's algorithm preserves sign and magnitude.
-Importance values for Olden's algorithm are from the summed product of model weights and are not rescaled.}
+\caption{Variable importance for three models using Garson's algorithm
+  for relative importance (\code{garson},
+  \cref{fig:plotimp1,fig:plotimp3,fig:plotimp5};
+  \citealt{Garson91,Goh95}) and Olden's connection weights algorithm
+  (\code{olden}, \cref{fig:plotimp2,fig:plotimp4,fig:plotimp6};
+  \citealt{Olden04}).  Garson's algorithm shows importance as absolute
+  values from 0--1, whereas Olden's algorithm preserves sign and
+  magnitude.  Importance values for Olden's algorithm are from the
+  summed product of model weights and are not rescaled.}
 \label{fig:plotimp}
@@
 
-An alternative approach to evaluate variable relationships in a neural network is the Lek profile method \citep{Lek96,Gevrey03}.
-The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.
-The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method.
-However, it is one of few methods used to evaluate sensitivity in neural networks.
+An alternative approach to evaluate variable relationships in a neural
+network is the Lek profile method \citep{Lek96,Gevrey03}. The profile
+method differs fundamentally from the variable importance algorithms
+by evaluating the behavior of response variables across different
+values of the input variables. The method is generic and can be
+extended to any statistical model in \proglang{R} with a
+\code{predict} method. However, it is one of few methods used to
+evaluate sensitivity in neural networks.
 
-The \code{lekprofile} function evaluates the effects of input variables by returning a plot of model predictions across the range of values for each variable.
-The remaining explanatory variables are held constant when evaluating the effects of each input variable.
-The \code{lekprofile} function provides two options for setting constant values of unevaluated explanatory variables.
-The first option follows the original profile method by holding unevaluated variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}).
-This is implemented by creating a matrix where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables.
-All explanatory variables are held constant (e.g., median) while the variable of interest is sequenced from its minimum to maximum.
-This matrix is then used to predict values of the response variable from a fitted model object.
-This is repeated for each explanatory variable to obtain all response curves.
-Constant values are set in \code{lekprofile} by passing one or more values in the range 0--1 to the \code{group\_vals} argument.
-The default holds variables at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles (i.e., \code{group\_vals = c(0, 0.2, 0.4, 0.6, 0.8, 1)}).
+The \code{lekprofile} function evaluates the effects of input
+variables by returning a plot of model predictions across the range of
+values for each variable.  The remaining explanatory variables are
+held constant when evaluating the effects of each input variable.  The
+\code{lekprofile} function provides two options for setting constant
+values of unevaluated explanatory variables.  The first option follows
+the original profile method by holding unevaluated variables at
+different quantiles (e.g., minimum, 20th percentile, maximum;
+\cref{fig:plotlek_sens1,fig:plotlek_bars1}). This is implemented by
+creating a matrix where the number of rows is the number of
+observations in the original dataset and the number of columns is the
+number of explanatory variables. All explanatory variables are held
+constant (e.g., at the median) while the variable of interest is
+sequenced from its minimum to maximum. This matrix is then used to
+predict values of the response variable from a fitted model
+object. This is repeated for each explanatory variable to obtain all
+response curves.  Constant values are set in \code{lekprofile} by
+passing one or more values in the range 0--1 to the \code{group\_vals}
+argument.  The default holds variables at the minimum, 20th, 40th,
+60th, 80th, and maximum percentiles (i.e., \code{group\_vals = c(0,
+  0.2, 0.4, 0.6, 0.8, 1)}).
 
-A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data.
-Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable).
-The second option holds unevaluated variables at means defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}).
-Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer greater than one.
-The centers (means) of the clusters are then used as constants for the unevaluated variables.
-\citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.
-Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more interpretable response curves.
-Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
-For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
-Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
+A second implementation of \code{lekprofile} is to group the
+unevaluated explanatory variables by natural groupings defined by the
+data. Covariance among predictors may present unlikely scenarios if
+all unevaluated variables are held at the same level (e.g., high
+values for one variable may be unlikely with high values for a second
+variable). The second option holds unevaluated variables at means
+defined by natural clusters in the data
+(\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified
+using {\it k}-means clustering (\code{kmeans} from the base package
+\pkg{stats}; \citealt{Hartigan79}) of the input variables if the
+argument passed to \code{group_vals} is an integer greater than
+one. The centers (means) of the clusters are then used as constants
+for the unevaluated variables.  \citet{Beck14a} provide an example of
+the clustering method for \code{lekprofile} by evaluating response of
+a lake health index to different explanatory variables.  Lake clusters
+were identified given covariance among variables, such that holding
+explanatory variables at values defined by clusters created more
+interpretable response curves.  Both methods return similar plots,
+with additional options to visualize the groupings for unevaluated
+explanatory variables (\cref{fig:plotlek_bars}).  For the latter case,
+\code{group\_show = TRUE} will return a stacked bar plot for each
+group with heights within each bar proportional to the constant
+values.  Sensitivity profiles were created using the standard approach
+based on quantiles and using the alternative clustering method
+(\cref{fig:plotlek_sens}), including bar plots of the relative values
+for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
+%
 \begin{Schunk}
@@
 \end{Schunk}
-\begin{figure}[!ht]
+%
+\begin{figure}[t!]
 \centering
 \subfloat[][Response with quantiles]{
-\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
+\includegraphics[width = 0.9\textwidth, page = 1]{plotlek_sens.pdf}
 \label{fig:plotlek_sens1}
@@
 \subfloat[][Response with clusters]{
-\includegraphics[width = 0.9\textwidth, page = 2]{figs/plotlek_sens.pdf}
+\includegraphics[width = 0.9\textwidth, page = 2]{plotlek_sens.pdf}
 \label{fig:plotlek_sens2}
@@
 \caption{Sensitivity analysis of a neural network using the Lek profile method to evaluate the effects of explanatory variables.
-\Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles (minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles) and \cref{fig:plotlek_sens2} groups by cluster means (six groups).
+\Cref{fig:plotlek_sens1} groups unevaluated explanatory variables at quantiles (minimum, 20th, 40th, 60th, 80th, and maximum percentiles) and \cref{fig:plotlek_sens2} groups by cluster means (six groups).
 Values at which explanatory variables are held constant for each group are shown in \cref{fig:plotlek_bars1,fig:plotlek_bars2}.}
@@
 %%%%
-\begin{figure}
+\begin{figure}[t!]  
 \centering
 \subfloat[][Quantile groupings]{
-\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_bars.pdf}
+\includegraphics[width = 0.9\textwidth, page = 1]{plotlek_bars.pdf}
 \label{fig:plotlek_bars1}
@@
 \subfloat[][Cluster groupings]{
-\includegraphics[width = 0.9\textwidth, page = 2]{figs/plotlek_bars.pdf}
+\includegraphics[width = 0.9\textwidth, page = 2]{plotlek_bars.pdf}
 \label{fig:plotlek_bars2}
@@
 \caption{Bar plots for values of unevaluated explanatory variables in each group in \cref{fig:plotlek_sens1,fig:plotlek_sens2}.
-\Cref{fig:plotlek_bars1} shows default quantile groupings set at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles.
-For example, variables are held at negative values for group 1 (i.e., stacked bars with negative heights) for the minimum value, whereas group 6 holds variables at their maximum (largest positive heights).
+\Cref{fig:plotlek_bars1} shows default quantile groupings set at the minimum, 20th, 40th, 60th, 80th, and maximum percentiles.
+For example, variables are held at negative values for group~1 (i.e., stacked bars with negative heights) for the minimum value, whereas group~6 holds variables at their maximum (largest positive heights).
 \Cref{fig:plotlek_bars2} shows the cluster centers for each variable in each group.
@@
 
-Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.
-A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.
-A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.
-The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.
-In general, the following should be considered during initial development \citep{Ripley96, Lek00, Maier00}:
+Although \pkg{NeuralnetTools} provides several methods to extract
+information from a fitted neural network, it does not provide explicit
+guidance for developing the initial model.  A potentially more
+challenging aspect of using \ac{mlp} neural networks is understanding
+the effects of network architecture on model performance, appropriate
+use of training and validation datasets, and implications for the
+bias-variance tradeoff with model over- or under-fitting
+\citep{Maier00}.  A detailed discussion of these issues is beyond the
+scope of this paper, although an example application is presented
+below to emphasize the importance of these considerations.  The models
+presented above, including the \code{neuraldat} dataset, are contrived
+examples to illustrate use of the \pkg{NeuralNetTools} package and
+they do not demonstrate a comprehensive or practical application of
+model development.  In general, the following should be considered
+during initial development \citep{Ripley96, Lek00, Maier00}:
 \begin{itemize}
-\item Initial data pre-processing to normalize inputs, standardize response, and assess influence of outliers,
-\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, and pruning weights or inputs,
-\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
-\item Initial starting weights for the backpropagation algorithm, and
-\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
+\item Initial data pre-processing to normalize inputs, standardize response, and assess influence of outliers.
+\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, and pruning weights or inputs.
+\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc.
+\item Initial starting weights for the backpropagation algorithm.
+\item Criteria for stopping model training, e.g., error convergence tolerance, maximum number of iterations, minimum error on test dataset, etc.
 \end{itemize}
 
-A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate 1) use of the functions in \pkg{NeuralNetTools} to gain additional insight into relationships among variables, and 2) effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).
-This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.
-The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.
-Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).
+A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to
+demonstrate (1) the use of the functions in \pkg{NeuralNetTools} to
+gain additional insight into relationships among variables, and (2)
+the effects of training conditions on model conclusions.
+This dataset provides information on all flights departing New York
+City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights
+from the UA carrier in the month of December to identify variables
+that potentially influence arrival delays (\code{arr\_delay}, minutes)
+at the destination airport.  Factors potentially related to delays
+are selected from the dataset and include departure delay
+(\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours,
+minutes), arrival time (\code{arr\_time}, hours, minutes), travel time
+between destinations (\code{air\_time}, minutes), and distance flown
+(\code{distance}, miles).
 
-First, the apropriate month and airline carrier were selected, all explanatory variables were normalized, and the response variable was standardized to 0--1.
+First, the appropriate month and airline carrier are selected, all
+explanatory variables are scaled and centered, and the response variable is
+scaled to 0--1.
+%
 \begin{Schunk}
@@
 R> library("dplyr")
-R> tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
-+   select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
-+   mutate_each(funs(scale), -arr_delay) %>% 
+R> tomod <- filter(flights, month == 12 & carrier == "UA") %>% 
++    select(arr_delay, dep_delay, dep_time, arr_time, air_time,
++      distance) %>% mutate_each(funs(scale), -arr_delay) %>% 
 +    mutate_each(funs(as.numeric), -arr_delay) %>% 
@@
 \end{Schunk}
-Then, a standard \ac{mlp} with five hidden nodes was created with the \pkg{nnet} package to model the effects of selected variables on arrival delays.
-The entire dataset is used for the example but separate training and validation datasets should be used in practice.
+%
+Then, a standard MLP with five hidden nodes was created with the
+\pkg{nnet} package to model the effects of selected variables on
+arrival delays. The entire dataset is used for the example but
+separate training and validation datasets should be used in practice.
+%
 \begin{Schunk}
 \begin{Sinput}
-R> library(nnet)
-R> mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
+R> library("nnet")
+R> mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod,
++    trace = FALSE)  
 \end{Sinput}
 \end{Schunk}
+%
 The default output is limited to structural information about the model and methods for model predictions (see \code{str(mod)} and \code{?predict.nnet}).
 Using functions from \pkg{NeuralNetTools}, a more comprehensive understanding of the relationships between the variables is illustrated.
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
-\Cref{fig:flightplots} shows the information about arrival delays that can be obtained with the functions in \pkg{NeuralNetTools}.
-The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used as a general characterization of the relationships between variables.
-For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays.
-Similarly, a large positive weight is observed between I3 and H1, suggesting arrival time is positively associated with arrival delays.
-However, interpreting individual connection weights between layers is challenging.
-\Cref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more quantitative descriptions using information from both the \ac{nid} and model predictions.
-\Cref{fig:flightplotsb} shows variable importance using the \code{garson} and \code{olden} algorithms.
-The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.
-However, the \code{garson} function shows arrival time (\code{arr\_time}) as the third most important variable for arrival delays, whereas this is ranked highest by the \code{olden} function.
-Similar discrepancies between the two methods are observed for other variables, which are explained below.
-Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.
-Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group four that generally had early arrival times and long departure delays.
+%
+\Cref{fig:flightplots} shows the information about arrival delays that
+can be obtained with the functions in \pkg{NeuralNetTools}. The
+\ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be
+used as a general characterization of the relationships between
+variables.  For example, most of the connection weights from input
+node I5 are negative (gray), suggesting that distance traveled has an
+opposing relationship with arrival delays. Similarly, a large positive
+weight is observed between I3 and H1, suggesting arrival time is
+positively associated with arrival delays.  However, interpreting
+individual connection weights between layers is challenging.
+Figures~\ref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more
+quantitative descriptions using information from both the \ac{nid} and
+model predictions.  \Cref{fig:flightplotsb} shows variable importance
+using the \code{garson} and \code{olden} algorithms.  The
+\code{garson} function suggests time between destinations
+(\code{air\_time}) has the strongest relationship with arrival delays,
+similar to a strong positive association shown with the \code{olden}
+method. However, the \code{garson} function shows arrival time
+(\code{arr\_time}) as the third most important variable for arrival
+delays, whereas this is ranked highest by the \code{olden}
+function. Similar discrepancies between the two methods are observed
+for other variables, which are explained below. Finally, results from
+the \code{lekprofile} function (Figure~\ref{fig:flightplotsc}) confirm those
+in Figure~\ref{fig:flightplotsb}, with the addition of non-linear responses
+that vary by different groupings of the data.  Values for each
+variable in the different unevaluated groups (based on clustering)
+show that there were no obvious patterns between groups with the
+exception being group four that generally had early arrival times and
+long departure delays.
 
 %%%%
-\begin{figure}
+\begin{figure}[t!]
 \centering
 \subfloat[][\code{plotnet}]{
-\includegraphics[width = 0.5\textwidth, page = 1]{figs/flightplots1.pdf}
+\includegraphics[width = 0.5\textwidth, page = 1]{flightplots1.pdf}
 \label{fig:flightplotsa}
@@
 \subfloat[][\code{garson} (top) and \code{olden} (bottom)]{
-\includegraphics[width = 0.5\textwidth, page = 2]{figs/flightplots1.pdf}
+\includegraphics[width = 0.5\textwidth, page = 2]{flightplots1.pdf}
 \label{fig:flightplotsb}
@@
 \subfloat[][\code{lekprofile}]{
-\includegraphics[width = 1.05\textwidth, page = 1]{figs/flightplots2.pdf}
+\includegraphics[width = 1.05\textwidth, page = 1]{flightplots2.pdf}
 \label{fig:flightplotsc}
@@
 
-A second analysis was needed to show the effects of network architecture and initial starting weights on uncertainty in estimates of variable importance.
-Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.
-Each model had a random set of starting weights for the first training iteration.
-Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).
+A second analysis is needed to show the effects of network
+architecture and initial starting weights on uncertainty in estimates
+of variable importance. Models with one, five, or ten hidden nodes and
+100 separate models for each node level are created.  Each model has
+a random set of starting weights for the first training iteration.
+Importance estimates using \code{olden} are saved for each model and
+combined in a single plot to show overall variable importance as the
+median and 5th/95th percentiles from the 100 models for each node
+level.
 
-Several conclusions from \cref{fig:flightimp} provide further information to interpret the trends in \cref{fig:flightplots}.
-First, consistent relationships can be identified such that delays in arrival time were negatively related to distance and positively related to departure delays and air time.
-That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.
-No conclusions can be made for the other variables because the bounds of uncertainty included zero.
-Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).
-This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.
-Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.
-This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.
-Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
-\begin{figure}[!h]
+Several conclusions from \cref{fig:flightimp} provide further
+information to interpret the trends in \cref{fig:flightplots}.  First,
+consistent relationships can be identified such that delays in arrival
+time are negatively related to distance and positively related to
+departure delays and air time.  That is, flights arrived later than
+their scheduled time if flight time was long or if their departure was
+delayed, whereas flights arrived earlier than scheduled for longer
+distances.  No conclusions can be made for the other variables because
+the bounds of uncertainty include zero.  Second, the range of
+importance estimates varies between the models (i.e., one node varies
+between $\pm 1$ and the others between $\pm 3$).  This suggests that
+the relative importance estimates only have relevance within each
+model, whereas only the rankings (e.g., least, most important) can be
+compared between models.  Third and most important, the level of
+uncertainty for specific variables can be large between model fits for
+the same architecture.  This suggests that a single model can provide
+misleading information and therefore several models may be required to
+decrease uncertainty.  Additional considerations described above
+(e.g., criteria for stopping training, use of training and test
+datasets) can also affect the interpretation of model information and
+should be considered equally during model development.
+\begin{figure}[t!]
 \centering
 \subfloat[][Networks with one node]{
-\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
+\includegraphics[width = 0.95\textwidth, page = 1]{flightimp.pdf}
 \label{fig:flightimp1}
@@
 \subfloat[][Networks with five nodes]{
-\includegraphics[width = 0.95\textwidth, page = 2]{figs/flightimp.pdf}
+\includegraphics[width = 0.95\textwidth, page = 2]{flightimp.pdf}
 \label{fig:flightimp2}
@@
 \subfloat[][Networks with ten nodes]{
-\includegraphics[width = 0.95\textwidth, page = 3]{figs/flightimp.pdf}
+\includegraphics[width = 0.95\textwidth, page = 3]{flightimp.pdf}
 \label{fig:flightimp3}
@@
 
-The \pkg{NeuralNetTools} package provides a simple approach to improve the quality of information obtained from a feed-forward \ac{mlp} neural network.
-Functions can be used to visualize a neural network using a \acl{nid} (\code{plotnet}), evaluate variable importance (\code{garson}, \code{olden}), and conduct a sensitivity analysis (\code{lekprofile}).
-Although visualizing a neural network with \code{plotnet} is impractical for large models, the remaining functions can simplify model complexity to identify important relationships between variables.
-Methods are available for the most used \ac{CRAN} packages that can create neural networks (\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas additional methods could be added based on popularity of the remaining packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
+The \pkg{NeuralNetTools} package provides a simple approach to improve
+the quality of information obtained from a feed-forward \ac{mlp}
+neural network.  Functions can be used to visualize a neural network
+using a \acl{nid} (\code{plotnet}), evaluate variable importance
+(\code{garson}, \code{olden}), and conduct a sensitivity analysis
+(\code{lekprofile}).  Although visualizing a neural network with
+\code{plotnet} is impractical for large models, the remaining
+functions can simplify model complexity to identify important
+relationships between variables.  Methods are available for the most
+frequently used \ac{CRAN} packages that can create neural networks
+(\pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, \pkg{RSNNS}), whereas
+additional methods could be added based on popularity of the remaining
+packages (\pkg{AMORE}, \pkg{FCNN4R}, \pkg{monmlp}, \pkg{qrnn}).
 
-A primary objective of the package is to address the concern that supervised neural networks are `black boxes' that provide no information about underlying relationships between variables \citep{Paruelo97,Olden02}.
-Although neural networks are considered relatively complex statistical models, the theoretical foundation has many parallels with simpler statistical techniques that provide for evaluation of variable important \citep{Cheng94}.
-Moreover, the model fitting process minimizes error using a standard objective function such that conventional techniques to evaluate model sensitivity or performance (e.g., cross-validation) can be used with neural networks.
-As such, functions in \pkg{NeuralNetTools} can facilitate the selection of the optimal network architecture or can be used for post-hoc assessment.
+A primary objective of the package is to address the concern that
+supervised neural networks are ``black boxes'' that provide no
+information about underlying relationships between variables
+\citep{Paruelo97,Olden02}.  Although neural networks are considered
+relatively complex statistical models, the theoretical foundation has
+many parallels with simpler statistical techniques that provide for
+evaluation of variable importance \citep{Cheng94}.  Moreover, the model
+fitting process minimizes error using a standard objective function
+such that conventional techniques to evaluate model sensitivity or
+performance (e.g., cross-validation) can be used with neural networks.
+As such, functions in \pkg{NeuralNetTools} can facilitate the
+selection of the optimal network architecture or can be used for
+post-hoc assessment.
 
-Another important issue is determining when and how to apply neural networks given availability of alternative methods of analysis.
-The popularity of the \ac{mlp} neural network is partly to blame  for the generalizations and misperceptions about their benefits as modelling tools \citep{Burke97}.
-Perhaps an overstatement, the neural component is commonly advertised as a mathematical representation of the network of synaptic impulses in the human brain.
-Additionally, several examples have shown that the \ac{mlp} network may provide comparable predictive performance as similar statistical methods \citep{Feng02,Razi05,Beck14a}.
-A neural network should be considered a tool in the larger toolbox of data-intensive methods that should be used after examination of the tradeoffs between techniques, with particular emphasis on the specific needs of a dataset or research question.
-Considerations for choosing a method may include power given the sample size, expected linear or non-linear interactions between variables, distributional forms of the response, and other relevant considerations of exploratory data analysis \citep{Zuur10}.
-\pkg{NeuralNetTools} provides analysis tools that can inform evaluation and selection from among several alternative methods for exploratory data analysis.
+Another important issue is determining when and how to apply neural
+networks given availability of alternative methods of analysis.  The
+popularity of the \ac{mlp} neural network is partly to blame for the
+generalizations and misperceptions about their benefits as modeling
+tools \citep{Burke97}. Perhaps an overstatement, the neural component
+is commonly advertised as a mathematical representation of the network
+of synaptic impulses in the human brain.  Additionally, several
+examples have shown that the \ac{mlp} network may provide comparable
+predictive performance as similar statistical methods
+\citep{Feng02,Razi05,Beck14a}.  A neural network should be considered
+a tool in the larger toolbox of data-intensive methods that should be
+used after examination of the tradeoffs between techniques, with
+particular emphasis on the specific needs of a dataset or research
+question.  Considerations for choosing a method may include power
+given the sample size, expected linear or non-linear interactions
+between variables, distributional forms of the response, and other
+relevant considerations of exploratory data analysis \citep{Zuur10}.
+\pkg{NeuralNetTools} provides analysis tools that can inform
+evaluation and selection from among several alternative methods for
+exploratory data analysis.
 
-\section[Acknowledgments]{Acknowledgments}
+\section*{Acknowledgments}
 
-I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.
-Thanks to Sehan Lee and Marc Weber for reviewing an earlier draft.
-Contributions and suggestions from online users have also greatly improved the utility of the package.
-Funding for this project was supported in part by an Interdisciplinary Doctoral Fellowship provided by the Graduate School at the University of Minnesota to M.
-Beck.
+I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the
+University of Minnesota for general guidance during the development of
+this package.  Thanks to Sehan Lee and Marc Weber for reviewing an
+earlier draft.  Contributions and suggestions from online users have
+also greatly improved the utility of the package.  Funding for this
+project was supported in part by an Interdisciplinary Doctoral
+Fellowship provided by the Graduate School at the University of
+Minnesota to M.~Beck.
 
-% note that ref titles need to be in title-case 
-\input{refs.bbl}
+\bibliography{ref}
 

### 2018-05-03 nnt_manu.Rnw
working on final JSS edits

@@
 library("ggplot2")
-@
-% get online bib file
-<<echo = FALSE>>=
-refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
-refs <- rawToChar(refs$content)
-writeLines(refs, con = file('refs.bib'))
 @
@@
 toplo1 <- filter(imp_sums, nodes == 1) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 toplo2 <- filter(imp_sums, nodes == 5) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 toplo3 <- filter(imp_sums, nodes == 10) %>% 
-  mutate(variable = factor(variable, levels = levels(variable)[order(med)]))
+  mutate(variable = factor(variable, levels = variable[order(med)]))
 
@@
 wd <- 0.3
-ylims <- c(-3.93, 3.49)
+ylims <- with(imp_sums, c(min(lo), max(hi)))
 
@@
 % \bibliographystyle{jss}
-% \bibliography{refs}
-% this file is bbl edited by hand after running bibtex
-\input{CopyOfnnt_manu.bbl}
+\bibliography{refs}
 

### 2018-05-03 nnt_manu.tex
working on final JSS edits

@@
 % options and libs
-% get online bib file
 
@@
 Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.
-At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 95\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.
+At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 95\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 68\%.
 As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.

### 2018-05-03 final_JSS_edits/code.R

@@ -1,0 +1,175 @@
+# total downloads for neural network packages off CRAN
+# make sure to change cdate to time of writing
+library("cranlogs")
+today <- as.Date('2016-10-20')
+
+tots <- c(
+  sum(cran_downloads('nnet', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('neuralnet', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('RSNNS', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('AMORE', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('monmlp', from = '2000-01-01', to = today)$count),
+  sum(cran_downloads('qrnn', from = '2000-01-01', to = today)$count)
+)
+
+tots_nnt <- 100 * sum(tots[1:3])/sum(tots)
+tots_nnet <- 100 * tots[1]/sum(tots)
+
+## install.packages("NeuralNetTools")
+library("NeuralNetTools")
+
+set.seed(123)
+library("RSNNS")
+x <- neuraldat[, c('X1', 'X2', 'X3')]
+y <- neuraldat[, 'Y1']
+mod1 <- mlp(x, y, size = 5)
+library("neuralnet")
+mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
+library("nnet")
+mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
+
+par(mar = c(0, 0, 0, 0))
+plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
+plotnet(mod3)
+
+modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)
+
+par(mar = c(0, 0, 0, 0))
+plotnet(modskip, skip = TRUE)
+plotnet(modskip)
+
+pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01,
+  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
+mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
[…]

### 2018-05-04 final_JSS_edits/paper.tex
messing with reproducible stuff from jss

@@
 variables.  For example, most of the connection weights from input
-node I5 are negative (gray), suggesting that distance traveled has an
-opposing relationship with arrival delays. Similarly, a large positive
-weight is observed between I3 and H1, suggesting arrival time is
-positively associated with arrival delays.  However, interpreting
-individual connection weights between layers is challenging.
-Figures~\ref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more
-quantitative descriptions using information from both the \ac{nid} and
-model predictions.  \Cref{fig:flightplotsb} shows variable importance
-using the \code{garson} and \code{olden} algorithms.  The
-\code{garson} function suggests time between destinations
-(\code{air\_time}) has the strongest relationship with arrival delays,
-similar to a strong positive association shown with the \code{olden}
-method. However, the \code{garson} function shows arrival time
-(\code{arr\_time}) as the third most important variable for arrival
-delays, whereas this is ranked highest by the \code{olden}
+nodes I2 and I5 are strongly negative (gray), suggesting that departure 
+time and distance traveled has an opposing relationship with arrival 
+delays. Similarly, large positive weights are observed for I3 and I4, 
+suggesting arrival time and time in the air are positively associated 
+with arrival delays.  However, interpreting individual connection weights 
+between layers is challenging. Figures~\ref{fig:flightplotsb} and 
+\ref{fig:flightplotsc} provide more quantitative descriptions using 
+information from both the \ac{nid} and model predictions.  
+\Cref{fig:flightplotsb} shows variable importance using the \code{garson} 
+and \code{olden} algorithms.  The \code{garson} function suggests time 
+between destinations (\code{air\_time}) has the strongest relationship 
+with arrival delays, similar to a strong positive association shown 
+with the \code{olden} method. However, the \code{garson} function 
+shows arrival time (\code{arr\_time}) as the third most important 
+variable for arrival delays, whereas this is ranked highest by the \code{olden}
 function. Similar discrepancies between the two methods are observed
@@
 variable in the different unevaluated groups (based on clustering)
-show that there were no obvious patterns between groups with the
-exception being group four that generally had early arrival times and
-long departure delays.
+show that there were no obvious patterns between groups, with the
+exception being group one that generally had longer times in the air 
+and greater distance travelled.
 

### 2018-05-04 nnt_manu.Rnw
messing with reproducible stuff from jss

@@
 The entire dataset is used for the example but separate training and validation datasets should be used in practice.
-<<eval = F, eval = T, results = hide>>=
+<<eval = F, results = hide>>=
 library(nnet)

### 2018-05-04 final_JSS_edits/code.R

@@ -1,28 +1,30 @@
 # total downloads for neural network packages off CRAN
 # make sure to change cdate to time of writing
 library("cranlogs")
-today <- as.Date('2016-10-20')
+today <- as.Date("2016-10-20")
 
 tots <- c(
-  sum(cran_downloads('nnet', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('neuralnet', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('RSNNS', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('AMORE', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('monmlp', from = '2000-01-01', to = today)$count),
-  sum(cran_downloads('qrnn', from = '2000-01-01', to = today)$count)
+  sum(cran_downloads("nnet", from = "2000-01-01", to = today)$count),
+  sum(cran_downloads("neuralnet", from = "2000-01-01", to = today)$count),
+  sum(cran_downloads("RSNNS", from = "2000-01-01", to = today)$count),
+  sum(cran_downloads("FCNN4R", from = "2000-01-01", to = today)$count),
+  sum(cran_downloads("AMORE", from = "2000-01-01", to = today)$count),
+  sum(cran_downloads("monmlp", from = "2000-01-01", to = today)$count),
+  sum(cran_downloads("qrnn", from = "2000-01-01", to = today)$count)
 )
 
 tots_nnt <- 100 * sum(tots[1:3])/sum(tots)
+tots_nnt
 tots_nnet <- 100 * tots[1]/sum(tots)
+tots_nnet
 
-## install.packages("NeuralNetTools")
+# install.packages("NeuralNetTools")
 library("NeuralNetTools")
 
 set.seed(123)
 library("RSNNS")
-x <- neuraldat[, c('X1', 'X2', 'X3')]
-y <- neuraldat[, 'Y1']
+x <- neuraldat[, c("X1", "X2", "X3")]
+y <- neuraldat[, "Y1"]
 mod1 <- mlp(x, y, size = 5)
 library("neuralnet")
 mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
@@ -30,7 +32,7 @@ library("nnet")
 mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
 
[…]

### 2018-05-04 final_JSS_edits/paper.tex
almost there with final jss edits

@@
   Gulf Breeze, Florida, 32561, USA\\
-  Telephone: +1/850/934-2480\\
-  E-mail: \email{beck.marcus@epa.gov}
+  \textit{Current address:} Southern California Coastal Water Research Project\\
+  3535 Harbor Blvd., Suite 110\\
+  Costa Mesa, CA, 92626\\
+  Telephone: +1/714/755/3217\\
+  E-mail: \email{marcusb@sccwrp.org}
 }
@@
 \end{Schunk}
+%
+Note that the pruned network obtained with \pkg{RSNNS} and thus this
+plot might vary depending on the platform used.
 %
